---
title: "Getting started with pvstackr"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  pvstackr fits regression models to assessment data with plausible values,
  such as PISA, and reports the fixed-effect coefficients. This article shows
  how to install the package and write the `pv_fit()` call for the default
  method, `stack_direct`. It then reads a precomputed fit of the bundled
  synthetic data with `print()`, `get_estimates()` and a coefficient plot, and
  shows the columns of the estimate table that say whether pvstackr's reporting
  rule lets you read an interval as a confidence interval with nominal
  coverage.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Getting started with pvstackr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

## Plausible values

Large-scale assessments such as PISA do not publish a single test score for
each student. Because proficiency is not observed directly, each student
receives a set of *plausible values*: $M$ draws from that student's posterior
distribution of proficiency [@mislevy1991randomization;
@vondavier2009plausible]. The standard analysis fits the model once for each
plausible value, with the survey weights and a replicate-weight estimate of
each fit's sampling variance, and combines the $M$ results with Rubin's rules
for multiple imputation [@rubin1987multiple]. An analysis that uses only `PV1`
omits the between-imputation variance (the variation of the estimates across
plausible values), so its standard errors are too small.

## What pvstackr computes

The default method, `stack_direct`, fits one model to the stacked data ($M$
copies of the data, one per plausible value) and then calibrates the
fixed-effect draws of that fit to the *target*: the estimates, covariance
matrix and degrees of freedom that `pv_brr_target()` computes with the
standard analysis described above (a survey-weighted regression for each
plausible value, the BRR–Fay replicate weights for the sampling variance of
the coefficients, and Rubin's rules to combine the results). The calibration
transforms the draws so that their mean and covariance equal those of the
target. The setting `center = "target"` in `pv_control()`, the default and the
only value `stack_direct` accepts, centers the calibrated draws on the
target's estimates, so the reported estimates, standard errors and degrees of
freedom are those of the target: the Bayesian fit does not change them. It
supplies the calibrated draws, which can be used for quantities derived from
the coefficients, such as the difference between two of them, and diagnostics
that show how closely the fit agrees with the target. `stack_direct` needs one
Bayesian fit where a Bayesian analysis of each plausible value needs $M$, but
that fit has $M$ times as many rows.

`pv_fit()` also offers `per_pv`, which fits one model per plausible value and
combines the results with Rubin's rules, and `stack_psis`, which reweights one
stacked fit toward each plausible value with importance weights that you
supply. Both take the variance within each plausible value from the model's
posterior draws rather than from the replicate weights, so their intervals are
always descriptive. The `stack_direct` model is a linear regression weighted by
the survey weights; `stack_direct`, `stack_psis` and `pv_brr_target()` stop
with an error if the formula contains a random-effect term such as
`(1 | school)`. Only the fixed-effect coefficients (intercept and slopes) are
reported; other parameters of the model, such as the residual standard
deviation `sigma`, are neither calibrated nor reported.

By pvstackr's reporting rule, only a `stack_direct` fit whose target uses
Barnard–Rubin degrees of freedom has intervals that can be read as confidence
intervals with nominal coverage;
the example fit uses the classic Rubin degrees of freedom, so its intervals are
descriptive (see the [section on intervals](#intervals)).

The examples below read a fit of the small synthetic data set bundled with the
package, computed in advance ([the example data and fit](#example-fit)); the
article fits no model, so it runs in seconds.

## Installing pvstackr

Install pvstackr from GitHub, for example with the pak package:

```{r install-dev, eval = FALSE}
# install.packages("pak")
pak::pak("joonho112/pvstackr")
```

Then load it (the first chunk of this article did so):

```{r load-again, eval = FALSE}
library(pvstackr)
```

Reading the example fit needs no modeling engine. A fit on your own data
needs one: either the engine bundled with pvstackr or fitting functions of
your own. The bundled engine fits the model with the brms package, so it needs
brms and posterior installed.

## Calling `pv_fit()`

`pv_fit()` fits a model with any of the three methods. For `stack_direct`, the
default, a call on your own data has this form. The chunk is not run, because this article
computes no target and uses no engine:

```{r pv-fit-call, eval = FALSE}
fit <- pv_fit(
  data    = your_data,
  formula = OUTCOME ~ x + female,   # OUTCOME stands for each plausible value
  target  = your_target,            # made by pv_brr_target()
  method  = "stack_direct",
  control = pv_control(method = "stack_direct", backend = "brms")
)
```

- `data` is a data frame with one row per student: the plausible-value
  columns, the final survey weight, the replicate weights and the covariates.
- `formula` has the placeholder `OUTCOME` on the left-hand side. pvstackr
  replaces it by each plausible-value column in turn (`PV1READ`, `PV2READ`,
  …), so you write the model once instead of $M$ times.
- `target` is the design-based target, made by `pv_brr_target()` from the same
  data and formula ([The full analysis workflow](a2-the-workflow.html#target)
  shows the call); `pv_fit()` stops with an error when the data or the
  formula differ from those of the target.
- `control` holds the settings of the fit, made by `pv_control()` with the
  same `method`; `pv_fit()` stops with an error when the two differ.
  `backend = "brms"` selects the engine bundled with pvstackr;
  with the default, `backend = "none"`, a `stack_direct` fit stops with an
  error unless you pass fitting functions of your own.

Instead of the bundled engine you can pass three functions of your own to
`pv_fit()`: `fit_function`, `draws_function` and `diagnose_function`. [The full
analysis workflow](a2-the-workflow.html#own-engine) describes what each one
receives and returns.

## The example data and fit {#example-fit}

pvstackr includes a small synthetic data set and a `stack_direct` fit made from
it, so that the examples run without PISA files. Load both:

```{r load-example}
pisa_tiny <- read.csv(
  system.file("extdata", "pisa_tiny.csv", package = "pvstackr")
)

fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit
```

The data set, `pisa_tiny`, has `r nrow(pisa_tiny)` students in three schools
of one made-up country (`CNT` is `"SYN"`). Besides the student and school
identifiers `CNTSTUID` and `CNTSCHID`, it has a continuous covariate `x`, a 0/1
indicator `female`, two plausible values for reading, `PV1READ` and `PV2READ`
($M = 2$), the final survey weight `W_FSTUWT` and four replicate weights,
`W_FSTURWT1` to `W_FSTURWT4` ($R = 4$):

```{r example-data}
str(pisa_tiny)
```

The values were made up for the examples and the package tests; the data
contain no PISA records, so the numbers illustrate the output and support no
substantive conclusion. PISA 2022 provides ten plausible values per domain and
80 replicate weights [@oecd2024pisa]; the example has two and four, which keeps
the objects small. [Real PISA data: access, memory and
reproducibility](a5-real-pisa-guidance.html) describes how to work with real
PISA files.

The example fit was not made by a sampler. Its target was computed from
`pisa_tiny` by `pv_brr_target()` with the classic degrees of freedom, and the
fit was made with fitting functions written for the example, which ignore the
data and return draws built from the target's estimates and standard errors
together with a typed-in sampler record. The fit was saved without its
calibrated draws (`return_draws = FALSE` in `pv_control()`).

Because its draws were built around the target, the stacked fit agrees with
the target by construction. The diagnostics that compare the two show
differences of about $10^{-14}$ target standard errors, where a fit made by a
sampler would differ by its Monte Carlo error ([Reading and reporting the
results](a3-reading-results.html#target-and-diagnostics) shows these values).
The example therefore shows what the output of a fit looks like, not how well
the method works.

## Reading the fit

### The overview from `print()`

`print()` shows an overview of the fit:

```{r print-fit}
print(fit)
```

The first lines give the method, the status and the number of fixed effects.
The `target` line gives the source of the target: `external_brr_fay_rubin`
marks a target made by `pv_brr_target()` from the BRR–Fay replicate weights and
Rubin's rules. The `draws` line says that the calibrated draws were not kept,
and the `diagnostics` line names the parts that `get_diagnostics(fit)` returns.
The last line notes that the intervals are descriptive, which [Reading the
intervals](#intervals) explains.

`status: ok` means that none of pvstackr's checks raised a warning or blocked
the fit. It is not a statement that a sampler converged. For a `stack_direct`
fit the checks cover the sampler diagnostics (R-hat, effective sample sizes,
divergent transitions), the difference between the stacked fit and the target
before the calibration, the numerical stability of the calibration, and any
prior that you set (an explicit prior always gives a warning). A fit with
status `"warning"` still has estimates, and `print()`
then lists the reason codes; a fit with status `"blocked"` has an empty
estimate table. "Status and checks" in `?pvstackr_object_contracts` gives the
thresholds of the checks.

### The estimate table

`get_estimates()` returns the estimate table, a data frame with one row per
fixed effect. The chunk shows 8 of its 17 columns:

```{r get-estimates}
est <- get_estimates(fit)

est[, c("term", "estimate", "se", "df",
        "conf_low", "conf_high",
        "interval_role", "coverage_claim_allowed")]
```

The terms are the intercept `b_Intercept` and the slopes `b_x` and
`b_female`: pvstackr names each coefficient after its column of the model
matrix, with the prefix `b_`. In these synthetic data the intercept is about
458 points, the slope of `x` about 47 points per unit of `x`, and the
coefficient of `female` about 2 points. `se` is the standard error, and `df`
the degrees of freedom used for the interval from `conf_low` to `conf_high`.
The last two columns say how each interval can be read. [Reading and reporting
the results](a3-reading-results.html#estimate-table) describes all 17 columns.

### A plot of the slopes

The plot shows the two slopes with their intervals. The intercept is left out,
since at about 458 points it would stretch the axis, and a dashed line marks
zero:

```{r coef-figure, fig.width = 7, fig.height = 3, fig.cap = "Slopes of the example fit, from synthetic data, with their 95% descriptive intervals. The intercept, about 458 points, is not shown. The dashed line marks zero.", fig.alt = "Dot-and-interval plot of the two slopes of the example fit, on a horizontal axis in synthetic reading-score points. b_x, at the top, is about 47 with a short interval from about 44 to 49, well to the right of zero. b_female, at the bottom, is about 2 with a long interval from about minus 42 to 46, which contains zero. A dashed vertical line marks zero."}
slopes <- est[est$term != "b_Intercept", ]
slopes <- slopes[order(slopes$term), ]

y    <- seq_len(nrow(slopes))
xlim <- range(c(slopes$conf_low, slopes$conf_high, 0))

op <- par(mar = c(4.5, 7, 1, 1))
plot(
  slopes$estimate, y,
  xlim = xlim, ylim = c(0.5, nrow(slopes) + 0.5),
  yaxt = "n", ylab = "",
  xlab = "Coefficient (synthetic reading-score points)",
  pch = 19, cex = 1.4, col = "#1f6f9c"
)
abline(v = 0, lty = 2, col = "grey50")
segments(slopes$conf_low, y, slopes$conf_high, y, lwd = 2, col = "#1f6f9c")
points(slopes$estimate, y, pch = 19, cex = 1.4, col = "#1f6f9c")
axis(2, at = y, labels = slopes$term, las = 1)
par(op)
```

The interval of `b_x`, from about 44 to 49, lies well to the right of zero;
the interval of `b_female`, from about −42 to 46, contains zero. Both are
descriptive intervals; [Reading the intervals](#intervals) explains what that
means and why the intervals of this example are wide.

### The draws

The example fit was saved without its calibrated draws, so `get_draws()`
returns `NULL`:

```{r get-draws}
get_draws(fit)
```

A fit made with the default, `return_draws = TRUE`, keeps them. The estimate
table does not use the draws, because its values are the target's.

## Reading the intervals {#intervals}

`coverage_claim_allowed`, the last column of the estimate table above, is
`TRUE` when pvstackr's reporting rule lets you read the interval as a
confidence interval with nominal coverage (a coverage-claimable interval) and
`FALSE` when
pvstackr makes no statement about its coverage (a descriptive interval).
`interval_role`, the column before it, names the case. In the example every
row is `descriptive_classic_rubin` with
`coverage_claim_allowed = FALSE`, because the target of the fit uses the
classic Rubin degrees of freedom (`df_method = "classic"`, the default of
`pv_brr_target()`); the note in `print(fit)` says the same.

By pvstackr's reporting rule, `coverage_claim_allowed` is `TRUE` only for a
`stack_direct` fit whose target was made with `df_method = "barnard_rubin"`
and a positive `df_complete`; every other interval, including every `per_pv`
and `stack_psis` interval, is descriptive. Report an interval as a
confidence interval only when `coverage_claim_allowed` is `TRUE`, and as a
descriptive interval otherwise. [Reading and reporting the
results](a3-reading-results.html#interval-columns) explains the rule and the
six values of `interval_role`.

The intervals of the example are wide compared with their standard errors.
Each interval is the estimate plus or minus a $t$ quantile with `df` degrees of
freedom times `se`. The classic degrees of freedom of @rubin1987multiple are
$(M - 1)/\lambda^2$, where $\lambda$, the fraction of missing information (FMI)
of the coefficient, is the share of its variance that comes from the
differences between the plausible values. With $M = 2$ they cannot be smaller
than 1, and they are close to 1 when $\lambda$ is close to 1, as the `df`
column of the estimate table shows (1.02, 1.40 and 1.01). With so few degrees
of freedom, a 95% interval reaches about 12 standard errors on each side of
the estimate (6.6 for `b_x`), where an interval based on the normal
distribution would reach 1.96.

The FMI is stored in the target as `get_target(fit)$fmi`, not in the estimate
table; [Reading and reporting the results](a3-reading-results.html#fmi) shows
its values for the example and explains why they are large.

[The full analysis workflow](a2-the-workflow.html) goes through an analysis of
your own data, from checking the columns to reading the result, and [Reading
and reporting the results](a3-reading-results.html) describes everything a fit
contains and what to report. [Comparing the three fitting
methods](a4-comparing-methods.html) compares `stack_direct` with `per_pv` and
`stack_psis`, and the Method articles, starting with [Plausible values, survey
weights and notation](m1-foundations-and-notation.html), give the formulas and
assumptions behind the target, the stacked fit and the calibration.

Bug reports and feature requests:
<https://github.com/joonho112/pvstackr/issues>.

## References
