---
title: "The full analysis workflow"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  This article goes through an analysis with pvstackr in order: checking the
  plausible-value and weight columns with `pv_design()`, computing the
  design-based target with `pv_brr_target()`, fitting the model with
  `pv_fit()`, and reading the estimate table. The first two steps run on the
  bundled synthetic data; the fit is shown as code and read from the example
  fit that ships with the package. The article also explains why
  `stack_direct` requires `center = "target"` and what your own fitting
  functions receive and must return.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{The full analysis workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

To analyze your own data with pvstackr, go through these steps in order:

| Step | Function | Result |
|:--|:--|:--|
| Check the plausible-value and weight columns (optional) | `pv_design()` | the column names, the numbers of plausible values ($M$) and replicate weights ($R$), and the Fay coefficient |
| Compute the design-based target | `pv_brr_target()` | the estimates of the fixed effects, their covariance matrix, standard errors and degrees of freedom |
| Fit the model | `pv_fit()` with settings from `pv_control()` | a fit whose estimate table holds the target's numbers, with calibrated draws and diagnostics |
| Compare with fits made by other methods (optional) | `pv_compare_methods()` | the estimates of two or more fits side by side |
| Read the result | `summary()`, `get_estimates()` | the estimate table |

The comparison is the subject of [Comparing the three fitting
methods](a4-comparing-methods.html). The target
and the fit use the same model, a linear regression of the plausible values on
the covariates, weighted by the survey weights, and only its fixed effects are
reported. pvstackr fits no school-level variance, and the only other parameter
of the stacked fit, the residual standard deviation `sigma`, is neither
calibrated nor reported.

The examples use `pisa_tiny`, a small synthetic data set included in the
package. The first two steps run below. Fitting needs a Bayesian sampler, so
the article shows the fitting calls without running them and reads the example
fit that comes with pvstackr. The example fit and its synthetic data are
described in [Getting started with
pvstackr](a1-getting-started.html#example-fit).

## Checking the columns with `pv_design()` {#check-columns}

`pv_design()` finds and checks the columns that hold the plausible values, the
final survey weight and the replicate weights, and records the Fay coefficient
$k$ that was used to make the replicate weights. It fits no model. The step is
optional: `pv_brr_target()` takes the same column arguments and finds
PISA-style plausible-value and replicate-weight columns itself; there too, give
`pv_suffix` for names such as `PV1READ`, and name the final weight, which has
no default. A design object is a quick check that a new file
has the columns you expect before anything is computed. Read the bundled data
and check its columns:

```{r design}
pisa_tiny <- read.csv(
  system.file("extdata", "pisa_tiny.csv", package = "pvstackr")
)

design <- pv_design(
  pisa_tiny,
  formula     = OUTCOME ~ x + female,   # OUTCOME stands for each plausible value
  pv_suffix   = "READ",                 # finds PV1READ and PV2READ
  expected_M  = 2L,                     # stop unless 2 plausible values are found
  expected_R  = 4L,                     # stop unless 4 replicate weights are found
  id_cols     = "CNTSTUID"             # the column that identifies each row
)

design
```

The print shows what was found: 12 rows, the formula `OUTCOME ~ x + female`,
2 plausible values ($M = 2$), the final weight `W_FSTUWT`, 4 replicate weights
($R = 4$), the Fay coefficient `fay_k = 0.5`, and `design hash`, a checksum of
the formula, the declared columns and their values.

The formula has the placeholder `OUTCOME` on the left-hand side. You write the
model once, and pvstackr replaces `OUTCOME` by each plausible-value column in
turn (`PV1READ`, then `PV2READ`). The survey weights do not go into the formula:
a `weights()` term is an error. Name the weight columns in `weight_col` and
`rep_weight_cols`; the defaults of `pv_design()` match the PISA names
`W_FSTUWT` and `W_FSTURWT1`, `W_FSTURWT2`, and so on.

PISA 2022 files name the plausible values by subject, such as `PV1READ`,
`PV2READ`, … for reading, so `pv_suffix = "READ"` selects the reading values;
the default suffix `""` matches only bare names such as `PV1`. `expected_M` and
`expected_R` are optional: they stop the function when a different number of
columns is found, for example because the suffix is wrong. `id_cols` names the
column that identifies each row (here the student ID); its values must be
unique. [Real PISA data: access, memory and
reproducibility](a5-real-pisa-guidance.html#design) shows the call for a PISA
2022 file.

`pv_design()` checks only the form of the formula; the covariates are checked
by `pv_brr_target()`, which stops with an error if one of them has missing
values, so remove incomplete rows first. No other pvstackr function takes the
design object itself. Its fields hold the columns that were found, and you pass
them on to `pv_brr_target()`:

```{r design-fields}
design$pv_cols
design$weight_col
design$rep_weight_cols
```

`pv_design()` finds the columns with two functions that you can also call on
their own, for example to look at a new file:

```{r detect-columns}
detect_pisa_pv_columns(pisa_tiny, suffix = "READ")
detect_pisa_brr_replicate_weights(pisa_tiny)
```

## Computing the target with `pv_brr_target()` {#target}

The target is the design-based estimate that the fit is calibrated to and whose
numbers it reports. `pv_brr_target()` computes it without a Bayesian model. For
each plausible value it fits the regression by weighted least squares, once
with the final weight and once with each of the $R$ replicate weights, and
estimates the covariance matrix of the coefficients from the spread of the
replicate estimates. It then combines the $M$ results with Rubin's rules. That
makes $M(R + 1)$ weighted least-squares fits, 10 here. Pass the data, the
formula and the columns from the design:

```{r target}
target <- pv_brr_target(
  pisa_tiny,
  formula         = OUTCOME ~ x + female,
  pv_cols         = design$pv_cols,
  weight_col      = design$weight_col,
  rep_weight_cols = design$rep_weight_cols,
  fay_k           = design$fay_k,
  id_cols         = design$id_cols
)

target
```

The print shows 3 fixed effects, 2 plausible values, 4 replicate weights,
`fay_k = 0.5`, the degrees-of-freedom rule (`df method: classic`), the interval
label `descriptive_classic_rubin` and the source label `external_brr_fay_rubin`.
A `stack_direct` fit copies the interval label into its estimate table and
repeats the source label in the column `target_source`.

For plausible value $m$, let $\hat\beta_m$ be the estimates with the final
weight and $\hat\beta_m^{(r)}$ those with replicate weight $r$. The BRR–Fay
replicate covariance of $\hat\beta_m$ is

$$
\hat U_m = \frac{1}{R(1-k)^2} \sum_{r=1}^{R}
  \big(\hat\beta_m^{(r)} - \hat\beta_m\big)\big(\hat\beta_m^{(r)} - \hat\beta_m\big)^\top ,
$$

where $k$ is the Fay coefficient `fay_k`. In each replicate, Fay's method
multiplies the weights of one half-sample by $2 - k$ and those of the other by
$k$, where plain BRR uses 2 and 0; the factor $1/(1-k)^2$ corrects the variance
for this smaller change of the weights [@judkins1990fay]. Rubin's rules then
combine the $M$ results [@rubin1987multiple]: the estimate $\bar\beta$ is the
average of the $\hat\beta_m$, and its covariance matrix is

$$
T_{\text{MI}} = \bar U + \Big(1 + \frac{1}{M}\Big) B ,
$$

where $\bar U$, the average of the $\hat U_m$, is the sampling covariance within
a plausible value, and $B$, the covariance matrix of the $M$ estimates
$\hat\beta_m$, is the variation between plausible values. The standard errors
are the square roots of the diagonal of $T_{\text{MI}}$.

The degrees of freedom of each coefficient $\ell$ follow the rule chosen with
`df_method`:

- `"classic"` (the default) is the rule of @rubin1987multiple,
  $\nu_\ell = (M - 1)/\lambda_\ell^2$, where
  $\lambda_\ell = (1 + 1/M)\,B_{\ell\ell}/T_{\text{MI},\ell\ell}$ is the share
  of the coefficient's total variance that comes from the variation between
  plausible values (stored in the target as `fmi` and as `lambda`).
- `"barnard_rubin"` is the rule of @barnard1999small, which corrects the
  classic value when the complete-data degrees of freedom are small. It needs
  `df_complete`, the complete-data degrees of freedom of the replicate-weight
  variance, which you state.

The rule also sets the interval labels that the fit copies. A target made with
`"classic"`, like this one, gives descriptive intervals; only a target made
with `"barnard_rubin"` gives `interval_role = "coverage_barnard_rubin"` and
`coverage_claim_allowed = TRUE`. [Reading and reporting the
results](a3-reading-results.html#interval-columns) explains these columns.

The target keeps each piece as a field: `beta` ($\bar\beta$), `U_bar`
($\bar U$), `B`, `T_MI`, `se`, `df`, `fmi` ($\lambda_\ell$),
`fay_variance_multiplier` (the factor $1/\{R(1-k)^2\}$, which is 1 here) and
`target_hash`, a checksum of its contents:

```{r target-fields}
target$beta
target$T_MI
sqrt(diag(target$T_MI))   # these become the reported standard errors
```

These are the standard errors that a `stack_direct` fit reports. [The
design-based target: BRR–Fay replicate weights and Rubin's
rules](m2-brr-fay-target.html) derives the formulas above.

The target has fixed effects only. `pv_brr_target()` stops with an error when
the formula has a random-effect term such as `(1 | CNTSCHID)`, and so do
`stack_direct` and `stack_psis` fits:

```{r target-group-term, error = TRUE}
pv_brr_target(
  pisa_tiny,
  formula         = OUTCOME ~ x + (1 | CNTSCHID),
  pv_cols         = design$pv_cols,
  weight_col      = design$weight_col,
  rep_weight_cols = design$rep_weight_cols,
  fay_k           = design$fay_k
)
```

The message means that in pvstackr 0.2.x a target, and therefore a
`stack_direct` fit, can have fixed effects only.

## Fitting the model with `pv_fit()` {#fitting}

`pv_fit()` with `method = "stack_direct"`, the default, fits one Bayesian
model to the stacked data: $M$ copies of the data, one for each plausible
value, with that plausible value as the outcome, so $N \cdot M$ rows for $N$
students. Row $i$ of each copy has the weight $\tilde w_i/M$, where
$\tilde w_i = W_i/\overline W$ is the student's final weight $W_i$ divided by
the mean final weight $\overline W$, so the stacked log-likelihood is the
average of the $M$ survey-weighted log-likelihoods, one for each plausible
value. With these weights, the weighted least-squares estimate from the
stacked data equals the target's $\bar\beta$, the average of the estimates for
the separate plausible values. The stacked fixed-effect point identity,
Theorem 4.1 of the companion preprint [@lee2026onemcmc], states the conditions
for this equality; [One stacked fit and the Rubin point
estimate](m3-stacked-bridge.html) explains it.

The identity concerns the point estimate only. The covariance of the stacked
draws does not contain the variation between plausible values, $B$
[@lee2026onemcmc], and its part within a plausible value comes from the model,
not from the replicate weights. pvstackr therefore calibrates the stacked fit to the target.
The Cholesky calibration correction (CCC) transforms the fixed-effect draws so
that their mean and covariance equal the target's $\bar\beta$ and
$T_{\text{MI}}$.

### Loading the example fit

The example fit that comes with pvstackr was made from the same data and
target:

```{r example-fit}
fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit

fit
```

The print shows the method, the status, the number of fixed effects, the source
label of the target, that the calibrated draws were not kept, the names of the
diagnostics, and a note that the intervals are descriptive; the note appears
because the target uses the classic rule for the degrees of freedom (see the
degrees-of-freedom rules in [Computing the target](#target)). `status: ok` means
that none of pvstackr's checks gave a warning or a block, not that a sampler
converged; "Status and checks" in `?pvstackr_object_contracts` lists the checks
and their thresholds.

### The call for your own data

On your own data, call `pv_fit()` with the same data and formula that built the
target; `pv_fit()` stops with an error when they differ. The fit also needs an
engine. The bundled one, selected with
`pv_control(backend = "brms")`, fits the stacked model with the brms function
`brm()`:

```{r fit-bundled-engine, eval = FALSE}
fit <- pv_fit(
  data    = pisa_tiny,
  formula = OUTCOME ~ x + female,
  target  = target,              # from pv_brr_target() above
  method  = "stack_direct",
  control = pv_control(method = "stack_direct", backend = "brms")
)
```

The bundled engine needs the brms and posterior packages, which pvstackr
suggests but does not require. If cmdstanr is installed, CmdStan must be
configured, otherwise the fit stops with an error instead of switching to
rstan; if cmdstanr is not installed, brms uses rstan. `pv_control()` also sets
the sampler: by default 4 chains of 2000 iterations each, of which 1000 are
warm-up. `cache_dir`, an argument of `pv_fit()` that is passed on to
`pv_fit_direct()`, defaults to `"cache"`: the bundled engine saves the brms fit
in a folder `cache` in the working directory, which it creates. brms reuses
that file as long as the model and the data are unchanged, even when `chains`,
`iter`, `warmup` or `seed` change: a fit with another number of chains or of
draws per chain is then blocked, and a new seed returns the old draws without
a warning. Set `cache_dir = NULL`, or give a new `cache_stem`, when you change
these settings (see `?pv_fit_direct`).

To use another engine, pass three functions of your own instead; [Using your
own fitting functions](#own-engine) describes what each one receives and
returns:

```{r fit-own-functions, eval = FALSE}
fit <- pv_fit(
  data              = pisa_tiny,
  formula           = OUTCOME ~ x + female,
  target            = target,
  method            = "stack_direct",
  control           = pv_control(method = "stack_direct"),
  fit_function      = your_fit_function,      # fits the stacked model
  draws_function    = your_draws_function,    # returns the draws of that fit
  diagnose_function = your_diagnose_function, # returns its sampler diagnostics
  cache_dir         = NULL                    # no cache file for your function
)
```

## Reading the result

`summary()` repeats the overview of `print()` and adds the main columns of the
estimate table:

```{r summary}
summary(fit)
```

`get_estimates()` returns the estimate table as a data frame with one row per
fixed effect and 17 columns, of which the chunk below shows eight:

```{r get-estimates}
est <- get_estimates(fit)

est[, c("term", "estimate", "se", "df",
        "conf_low", "conf_high",
        "interval_role", "coverage_claim_allowed")]
```

These are the rows that [Getting started with pvstackr](a1-getting-started.html)
reads. The estimates and standard errors are the target's $\bar\beta$ and
$\sqrt{\operatorname{diag}(T_{\text{MI}})}$ computed above, for the reason
given in [the next section](#center-target), and `df` holds the target's
classic degrees of freedom, between 1.0 and 1.4 here, which is why the
intervals are so wide. [Reading and reporting the
results](a3-reading-results.html#interval-columns) explains the last two
columns.

## Why `stack_direct` requires `center = "target"` {#center-target}

The setting `center` of `pv_control()` says where the calibrated draws are
centered. `stack_direct` requires the default, `center = "target"`: CCC gives the
fixed-effect draws the target's mean $\bar\beta$ and covariance
$T_{\text{MI}}$, and the estimate table copies the target's numbers. `estimate`
is $\bar\beta$, `se` is $\sqrt{\operatorname{diag}(T_{\text{MI}})}$, and `df`,
`df_method`, `df_complete`, `interval_role` and `coverage_claim_allowed` come
from the target; none of them is computed from the stacked fit. The stacked fit
supplies the calibrated draws, which you can use for quantities derived from
the coefficients, and the diagnostics that compare the fit with the target.

This is pvstackr's rule. It makes every number in a row of the estimate table
come from one design-based calculation: the estimate, its standard error, the
degrees of freedom and the interval label are all the target's. The mean of the
stacked draws differs from $\bar\beta$ by Monte Carlo error, and further when
the prior is not flat, so pvstackr does not report it. It uses it in a check
instead: before the calibration, `delta_c_max` measures the largest difference
between a target estimate and the mean of the stacked draws, in target
standard errors, and a value of 0.01 or more gives the fit the status
`"warning"`, 0.05 or more the status `"blocked"`. [Calibrating the fixed-effect
draws (CCC)](m4-ccc-calibration.html) describes the calibration and its
diagnostics.

`pv_control()` also accepts `center = "posterior"`, which would keep the mean
of the stacked draws and calibrate only their covariance. No fitting function
uses it: `pv_fit()` with `method = "stack_direct"` stops with an error when it
is set, and `per_pv` and `stack_psis` ignore `center`.

The interval labels come from the target as well, and the calibration does not
change them. The example target uses the classic rule, so every row of the
estimate table above is descriptive. [Reading and reporting the
results](a3-reading-results.html#interval-columns) gives the rule that sets
these labels for every method.

## Using your own fitting functions {#own-engine}

Instead of the bundled engine, `stack_direct` can use three functions that you
pass to `pv_fit()`: `fit_function` fits the stacked model, `draws_function`
returns its draws, and `diagnose_function` returns its sampler diagnostics.
When you pass `fit_function`, pvstackr uses it whatever the `backend` value
(leave the default; the value is passed to your function and recorded in the
fit), `draws_function` is required as well, and without `diagnose_function` the fit
is blocked. Without `fit_function`, `pv_fit()` stops with an error unless
`backend = "brms"` selects the bundled engine. The skeleton below is not run;
its comments say what each function receives and must return:

```{r own-functions-sketch, eval = FALSE}
# fit_function: fit the stacked model and return the fitted object. pvstackr
# calls it with the named arguments formula, data, family, prior, chains, iter,
# warmup, cores, seed, backend, file and file_refit, plus the elements of
# additional_args. There is no weights argument: the weights are in the
# formula term weights(.pvstackr_weight) and in the data column
# .pvstackr_weight.
my_fit_function <- function(formula, data, ...) {
  # ... call your Bayesian engine here ...
}

# draws_function: return the draws of that fit as a numeric matrix with one
# row per draw. Name the fixed-effect columns as in target$fe_names
# (b_Intercept, b_x, ...) or after the stacked formula (b_pvstackrMM001, ...),
# or pass param_map.
my_draws_function <- function(backend_fit, ...) {
  # ... extract a draws matrix from backend_fit ...
}

# diagnose_function: return the sampler diagnostics as a named list with six
# values. If it is missing, fails or leaves out a value, the fit is blocked.
my_diagnose_function <- function(backend_fit, ...) {
  # ... return list(rhat_max = , ess_bulk_min = , ess_tail_min = ,
  #                 divergences = , chains = , post_warmup_draws_per_chain = )
}

fit <- pv_fit(
  data              = your_real_pisa_data,
  formula           = OUTCOME ~ x + female,
  target            = your_target,
  method            = "stack_direct",
  control           = pv_control(method = "stack_direct"),
  fit_function      = my_fit_function,
  draws_function    = my_draws_function,
  diagnose_function = my_diagnose_function,
  cache_dir         = NULL
)
```

`fit_function` receives the stacked data: one copy of your data per plausible
value, with the outcome in `.pvstackr_y`, the row weight $\tilde w_i/M$ in
`.pvstackr_weight`, and the columns of the design matrix, the intercept
included, as `pvstackrMM001`, `pvstackrMM002`, and so on. For the model of this
article its formula, in brms syntax, is
`.pvstackr_y | weights(.pvstackr_weight) ~ 0 + pvstackrMM001 + pvstackrMM002 + pvstackrMM003`.
A function that ignores the weights fits an unweighted model. It may return
any object that your other two functions accept.

`draws_function` takes that object and returns the draws as a numeric matrix or
data frame with one row per draw (at least two). The columns whose names start
with `b_` are the fixed effects; a `sigma` column is taken as the residual
standard deviation, which is not reported, and other columns such as `lp__` are
dropped. When your fixed effects have other names, say which columns they are
with `param_map`.

`diagnose_function` takes the same object and returns a named list with six
values: `rhat_max` (the largest R-hat), `ess_bulk_min` and `ess_tail_min` (the
smallest bulk and tail effective sample sizes, totals over all chains),
`divergences` (the number of divergent transitions), `chains` and
`post_warmup_draws_per_chain`. If the function is missing or fails, or any of
the six values is missing, the last two included, the fit is blocked
(`sampler_diagnostics_incomplete`) and has no estimates. The last two must also
equal `chains` and `iter - warmup` in `pv_control()`, or the fit is blocked as
well. pvstackr then checks R-hat, the effective sample sizes and the
divergences by the thresholds under "Status and checks" in
`?pvstackr_object_contracts`.

The other arguments of `fit_function` carry settings: `chains`, `iter`,
`warmup`, `cores`, `seed` and `backend` come from `pv_control()`, `family` is
always `stats::gaussian()`, `prior` is `NULL` unless you give one (see
`?pv_fit_direct`), and `additional_args`, a named list in the `pv_fit()` call,
adds arguments of your own. With the default `cache_dir = "cache"`, your
function receives `file = "cache/pvstackr-stack-direct"` and
`file_refit = "on_change"`, but pvstackr does not create the folder; with
`cache_dir = NULL` it receives `file = NULL` and `file_refit = "never"`.

You do not have to write all three functions. The functions of the bundled
engine are exported as `pv_backend_brms_fit_function()`,
`pv_backend_brms_draws_function()` and `pv_backend_brms_sampler_diagnostics()`,
so you can pass them and replace only the one that differs for your engine.
`?pv_fit_direct` gives the full lists of arguments and return values (see
`fit_function`, `draws_function`, `param_map`, `diagnose_function`,
`cache_dir` and `additional_args`), and `?pv_backend_brms_fit_function`
describes the bundled functions.

Once you have a fit, [Reading and reporting the
results](a3-reading-results.html) describes everything it contains and what to
report. For real PISA files, [Real PISA data: access, memory and
reproducibility](a5-real-pisa-guidance.html) covers data access, memory and
running time: the stacked data have $N \cdot M$ rows, and the target takes
$M(R + 1)$ weighted least-squares fits.

## References
