---
title: "Comparing the three fitting methods"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  pvstackr has three fitting methods, `stack_direct`, `per_pv` and
  `stack_psis`. This article describes how they differ, lines up three example
  fits with `pv_compare_methods()`, and reads the estimates and the agreement
  diagnostics of the comparison. It ends with two cautions, on the standard
  errors and on the Pareto k-hat values of `stack_psis`, and with pvstackr's
  recommendation for choosing a method.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Comparing the three fitting methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

pvstackr fits a model to plausible-value data with one of three methods:
`stack_direct` (the default), `per_pv` and `stack_psis`.

The example compares three fits of the small synthetic data set bundled with
pvstackr. The `stack_direct` fit is the example fit that ships with the
package; [Getting started with pvstackr](a1-getting-started.html#example-fit)
describes how it was made. The `per_pv` fit is made from random numbers near
the example estimates instead of posterior draws, because a real `per_pv` fit
needs one Bayesian fit per plausible value. The `stack_psis` fit gets
placeholder importance weights and Pareto $\hat k$ values, both explained in
[the three methods](#methods-table), that no program computed; since no
program is named, pvstackr blocks the fit (PSIS status
`provenance_incomplete`). The numbers therefore show the layout of the output
and say nothing about how the methods compare on real data.

## The three methods {#methods-table}

All three methods report only the fixed effects, and all three combine the
results for the $M$ plausible values with Rubin's rules [@rubin1987multiple],
so their standard errors include the variation between plausible values. They
differ in the number of Bayesian fits they need and in where the variance
within a plausible value comes from:

| Method | Bayesian fits | Variance within a plausible value | Intervals | Bundled engine |
|:--|:--|:--|:--|:--|
| `stack_direct` (default) | 1 | BRR–Fay replicate weights, through the target | coverage-claimable only with an external Barnard–Rubin target, otherwise descriptive | brms, with `pv_control(backend = "brms")` |
| `per_pv` | $M$ | posterior draws of each fit | always descriptive | none |
| `stack_psis` | 1 | posterior draws of the stacked fit, reweighted | always descriptive | none |

A coverage-claimable interval is one that pvstackr's reporting rule lets you
read as a confidence interval with nominal coverage
(`coverage_claim_allowed = TRUE`); [Reading and reporting the
results](a3-reading-results.html#interval-columns) gives the rule and the
`interval_role` label of each case.

`stack_direct` fits one model to the stacked data ($M$ copies of the data, one
per plausible value) and calibrates its fixed-effect draws to the target from
`pv_brr_target()`; its estimate table reports the target's estimates,
standard errors and degrees of freedom. The target itself takes $M(R + 1)$
weighted least-squares fits, which need no sampler. [The full analysis
workflow](a2-the-workflow.html#fitting) describes both steps.

`per_pv` combines one Bayesian fit per plausible value. You supply the fits,
either as their posterior draws (`per_pv_draws`) or as your own fitting
functions, which pvstackr calls once per plausible value. pvstackr does not
pass the survey weights to these functions, so a weighted fit has to take them
from the data. Scale them as pvstackr scales the weights of the stacked fit, by
dividing the final weight by its mean: with flat priors on the coefficients,
the posterior means do not depend on the scale of the weights, but the
posterior standard deviations do. PISA final
weights count the students that each sampled student represents
[@oecd2024pisa], so fits that use them unscaled give posterior standard
deviations that are far too small, and `per_pv` standard errors that are too
small. `?pv_fit_reference` lists what these functions receive and return.
pvstackr also collects no sampler diagnostics from them: a `per_pv`
fit always has status `"ok"`, so check the convergence of each fit yourself.

`stack_psis` reweights the draws of one fit to the stacked data toward each
plausible value. For each plausible value you supply importance weights and a
Pareto $\hat k$ value, computed outside pvstackr, for example by Pareto
smoothed importance sampling (PSIS) with the loo package
[@vehtari2017practical; @vehtari2024pareto]; $\hat k$ shows how reliable the
reweighted estimates are. pvstackr computes no part of PSIS: it divides each
column of weights by its sum, reweights the draws and combines the results
with Rubin's rules. It blocks the fit unless every $\hat k$ is below
`psis_k_threshold` in `pv_control()` (0.7 by default) and the program that
made the weights is named with `psis_producer` and `psis_producer_version`.
`?pv_fit_stack_psis` lists the inputs.

The number of Bayesian fits says how many samplers run, not how long the
analysis takes.

## Running a comparison

`pv_compare_methods()` takes two or more finished fits and lines up their
estimate tables by fixed-effect name; it fits no model itself. The three fits
below have the same fixed effects, `b_Intercept`, `b_x` and `b_female`. The
`stack_direct` fit is read from the package. The other two are made from draws
passed in directly: `pv_fit_reference()`, which runs the `per_pv` method, takes
a list of draw matrices, one per plausible value, and `pv_fit_stack_psis()`,
which runs `stack_psis`, takes a matrix of stacked draws with the weights and
the $\hat k$ values. No sampler runs and no PSIS is computed. The `stack_psis`
call names no program, because no program made its weights.

```{r build-fits}
fe <- c("b_Intercept", "b_x", "b_female")

fit_direct <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit

# Made-up draws for the per_pv fit: for each plausible value, 300 random values
# per fixed effect, with standard deviation 2, around 458, 47 and 2, near the
# example estimates. They stand in for posterior draws; no model was fitted.
set.seed(1)
mk <- function(mu) {
  d <- matrix(rnorm(300 * 3, sd = 2), ncol = 3, dimnames = list(NULL, fe))
  sweep(d, 2L, mu, "+")
}

fit_per_pv <- pv_fit_reference(
  per_pv_draws = list(PV1READ = mk(c(458, 47, 2)),
                      PV2READ = mk(c(458, 47, 2))),
  control      = pv_control(method = "per_pv")
)

# Made-up stacked draws with equal placeholder weights and Pareto k-hat values
# of 0.2. No PSIS was run, so no program is named, and pvstackr blocks the fit
# (its PSIS status is provenance_incomplete).
sd_mat <- mk(c(458, 47, 2))
fit_psis <- pv_fit_stack_psis(
  stacked_draws = sd_mat,
  pv_cols       = c("PV1READ", "PV2READ"),
  psis_weights  = matrix(1 / nrow(sd_mat), nrow = nrow(sd_mat), ncol = 2),
  pareto_k      = c(0.2, 0.2),
  control       = pv_control(method = "stack_psis")
)
```

The comparison is one call:

```{r compare}
cmp <- pv_compare_methods(
  stack_direct = fit_direct,
  per_pv       = fit_per_pv,
  stack_psis   = fit_psis
)

cmp
```

The print first names the reference fit, `per_pv`, from which the differences
in the comparison are taken. By default it is the first `per_pv` fit that is
not blocked, or, without one, the first fit that is not blocked; choose another with the argument `reference_method`. It then lists
the label and the method of each fit, the number of fixed effects (3) and the
blocked fit, `stack_psis`. The line `provenance note:` appears in every
comparison; [the agreement diagnostics](#agreement) explain it. The line
`interval note:` says that the compared intervals are descriptive.

The comparison has a row for each fit and each fixed effect that appears in any
of the fits. A blocked fit keeps its rows, with `NA` values and its reason
code, so that it stays visible in the comparison.

## The estimates side by side

`get_estimates()` on a comparison returns one row per fit and fixed effect,
here $3 \times 3 = 9$ rows. For each row it gives the fit's estimate, standard
error, degrees of freedom, interval and interval labels, the method, status and
reason codes of the fit, and the estimate and standard error of the reference
fit with the differences from them (see [the agreement
diagnostics](#agreement)); the source labels and checksums of the fits are in
`get_diagnostics()`. The
columns below put each estimate next to its interval and the two columns that
say how the interval can be read:

```{r aligned-estimates}
est <- get_estimates(cmp)

est[, c("method_label", "term", "estimate", "se",
        "conf_low", "conf_high",
        "interval_role", "coverage_claim_allowed")]
```

The `stack_direct` rows hold the numbers of the example fit, which are those of
its target. The `per_pv` estimates are close to them only because the made-up
draws were placed around 458, 47 and 2, and their standard errors, about 2,
come from the standard deviation of those draws. The `stack_psis` rows are `NA`
because that fit is blocked.

Two columns say how each interval can be read:

- `interval_role` is `descriptive_classic_rubin` for `stack_direct`, because
  the example target uses the classic Rubin degrees of freedom, and
  `reference_classic_rubin` for `per_pv`. It is `NA` for the blocked
  `stack_psis` rows; a `stack_psis` fit with estimates has
  `psis_classic_rubin` or `psis_barnard_rubin`.
- `coverage_claim_allowed` is `FALSE` for the two fits with estimates and `NA`
  for the blocked one. With a target made with Barnard–Rubin degrees of
  freedom, the `stack_direct` rows would have `TRUE`; the `per_pv` and
  `stack_psis` rows have `FALSE` whatever their degrees of freedom.

The intervals also differ in width. For the slope of `x`:

```{r interval-widths}
bx <- est[est$term == "b_x",
          c("method_label", "se", "conf_low", "conf_high")]
bx$width <- bx$conf_high - bx$conf_low
bx
```

The width of an interval depends on the standard error and on the degrees of
freedom. The `stack_direct` interval is 4.9 points wide although its standard
error is 0.37, because the example target has only 1.40 classic degrees of
freedom for `b_x`. The `per_pv` interval is 8.1 points wide, about four times
its standard error of 2.06, which comes from the made-up draws. Neither width
says anything about real data. The blocked `stack_psis` row has no interval.

The figure shows the estimates and intervals of the two slopes; the intercept,
about 458, is left out because its scale would compress the axis. The blocked
`stack_psis` fit has nothing to draw.

```{r overlay-figure, fig.width = 7, fig.height = 3.4, fig.cap = "Estimates and 95% intervals of the two slopes, b_x and b_female, from the stack_direct and per_pv fits. The blocked stack_psis fit has no estimates to draw. The dashed line marks zero. The per_pv values come from made-up draws.", fig.alt = "Dot-and-interval plot with one row for b_female, at the bottom, and one for b_x, at the top. In each row the stack_direct estimate, in blue, is drawn slightly below the per_pv estimate, in yellow, each with its 95 percent interval. The stack_direct interval for b_female runs from about −42 to 46 and covers most of the horizontal axis; the per_pv interval for b_female, from about −2 to 6, and the two intervals for b_x, between about 43 and 51, are short. A dashed vertical line marks zero. There is no point or interval for stack_psis."}
slopes  <- est[est$term != "b_Intercept",
               c("method_label", "term", "estimate", "conf_low", "conf_high")]
slopes  <- slopes[is.finite(slopes$estimate), , drop = FALSE]
methods <- unique(slopes$method_label)
terms   <- sort(unique(slopes$term))

offsets <- setNames(seq(-0.18, 0.18, length.out = length(methods)), methods)
cols    <- setNames(c("#1f6f9c", "#f2c14e", "grey45")[seq_along(methods)], methods)

xlim <- range(c(slopes$conf_low, slopes$conf_high, 0), na.rm = TRUE)

op <- par(mar = c(4.5, 7, 1, 1))
plot(
  NA, xlim = xlim, ylim = c(0.5, length(terms) + 0.5),
  yaxt = "n", ylab = "",
  xlab = "Coefficient (synthetic reading-score points)"
)
abline(v = 0, lty = 2, col = "grey50")
for (mlab in methods) {
  sub <- slopes[slopes$method_label == mlab, ]
  yy  <- match(sub$term, terms) + offsets[[mlab]]
  segments(sub$conf_low, yy, sub$conf_high, yy, lwd = 2, col = cols[[mlab]])
  points(sub$estimate, yy, pch = 19, cex = 1.3, col = cols[[mlab]])
}
axis(2, at = seq_along(terms), labels = terms, las = 1)
legend("topleft", legend = methods, col = cols, pch = 19, lwd = 2,
       bty = "n", cex = 0.9)
par(op)
```

Both intervals for `b_x` lie well above zero, and both intervals for
`b_female` include it. The `stack_direct` interval for `b_female`, from about
−42 to 46, is much wider than the `per_pv` one, from about −2 to 6, for the
reasons given above.

On real data, the estimates of the three methods are close when they all fit
the same survey-weighted model. A `stack_direct` fit reports the target's
$\bar\beta$, the average over the plausible values of the survey-weighted
least-squares estimates. A `per_pv` fit reports the average of the posterior
means of your fits. If each of these fits is the same survey-weighted
regression, with the weights taken from the data and flat priors on the fixed
effects (in brms, write the intercept as `0 + Intercept`, because its default
prior on the intercept is not flat), its posterior mean is the weighted least-squares estimate for its
plausible value, apart from Monte Carlo error; the `per_pv` estimates then
differ from $\bar\beta$ only by that error.

The same holds for a stacked fit, which `stack_direct` calibrates and
`stack_psis` reweights. When its rows are weighted by $\tilde w_i/M$, where
$\tilde w_i$ is the final survey weight divided by its mean, the weighted
least-squares estimate from the stacked data equals $\bar\beta$ exactly in
pvstackr's model, a weighted linear regression without random effects. The
stacked fixed-effect point identity, Theorem 4.1 of the companion preprint
[@lee2026onemcmc], states the conditions for this equality, and [One stacked
fit and the Rubin point estimate](m3-stacked-bridge.html) explains it. The
reweighted means of `stack_psis` approximate the posterior means for the
separate plausible values, and the Pareto $\hat k$ values show how reliable
these approximations are [@vehtari2024pareto]. The standard errors are another
matter, which the first of [the cautions](#cautions) explains.

## Agreement diagnostics {#agreement}

`get_diagnostics()` on a comparison returns a list:

```{r diag-keys}
dg <- get_diagnostics(cmp)
names(dg)
```

`reference_method` and `methods` repeat what `print()` shows, `statuses` gives
the status of each fit, and `blocked_methods` and `warning_methods` list the
fits with status `"blocked"` or `"warning"`. `timing` gives the number of model
fits of each method and, when you pass times that you measured with the
argument `timings`, the elapsed seconds; pvstackr does not time the fits
itself. The other three elements describe how the fits agree.

`agreement` has one row per fit. `max_abs_z_diff` is the largest, over the
fixed effects, of the absolute difference between the fit's estimate and the
reference estimate, divided by
$\sqrt{\mathrm{se}^2 + \mathrm{se}_{\text{ref}}^2}$. `max_abs_log_se_ratio` is
the largest absolute log of the ratio of the fit's standard error to the
reference standard error:

```{r agreement}
dg$agreement
```

The reference fit, `per_pv`, has zeros. For `stack_direct`, `max_abs_z_diff`
is 0.07: the estimates are close because the made-up `per_pv` draws were
placed near them. `max_abs_log_se_ratio` is 1.71 and comes from `b_x`, whose
`stack_direct` standard error, 0.37, is 0.18 times the `per_pv` one, 2.06. The
blocked `stack_psis` fit has no estimates (`n_available` is 0) and therefore
no values.

The estimate table of the comparison gives the same measures for each row
(`estimate_diff`, `se_ratio` and `abs_z_diff`), and `agreement_band` groups
`abs_z_diff` by pvstackr's cut-offs: `"close"` below 0.1, `"moderate"` from
0.1, `"different"` from 0.5, and `"not_available"` for a blocked fit. These values describe how far apart the fits
are; they are not a statistical test. In a real comparison, `max_abs_z_diff`
shows whether the fits agree on the estimates and `max_abs_log_se_ratio` how
far their standard errors differ; [the cautions](#cautions) explain why the
two need not go together.

`target_overlap` records whether fits share their numbers:

```{r target-overlap}
str(dg$target_overlap)
```

Every flag is `FALSE` here: the `stack_direct` fit has the target made by
`pv_brr_target()`, the `per_pv` fit has its own combination of its draws, and
the blocked fit has neither. The flags become `TRUE` when two or more fits
have the same target (`shared_target_hash`, and `shared_external_target` when
they are `stack_direct` fits calibrated to the same target) or the same
combined result (`shared_pooling_hash`), or when a fit has the target or the
combined result of the reference fit (`shares_reference_target`,
`shares_reference_pooling`). `shared_target_sources` lists the source labels
that two or more fits have, and `independence_caveat_required` is `TRUE` when
any flag is `TRUE` or that list is not empty. `independence_caveat` holds the
text of the `provenance note:` line.

Agreement between fits that share their numbers follows by construction: two
`stack_direct` fits calibrated to the same target report the same estimates
and standard errors, whatever their draws. Even without shared numbers, fits
of the same data and plausible values are not independent, so their agreement
does not confirm a result.

`method_diagnostics` has one row per fit and many columns, which
`?pv_compare_methods` lists; the chunk shows some of them:

```{r method-diagnostics}
dg$method_diagnostics[, c("method_label", "interval_role",
                          "coverage_claim_allowed", "target_source",
                          "psis_status", "pareto_k_max", "n_fits")]
```

`target_source` labels where the numbers of a fit come from:
`external_brr_fay_rubin` for the target from `pv_brr_target()` and
`per_pv_rubin_draws` for Rubin's rules applied to the `per_pv` draws. It is
`NA` for the blocked fit. A `stack_psis` fit with estimates has the label
`stack_psis_rubin_pooling` but no target object: `get_target()` returns `NULL`
for it. `n_fits` is the number of model fits: 1 for `stack_direct`, $M = 2$
for `per_pv` and 1 for `stack_psis`. For the blocked fit, `psis_status` is
`"provenance_incomplete"` because no program was named, and `pareto_k_max`,
the largest of the $\hat k$ values supplied, is the 0.2 typed into the
example; pvstackr records it but cannot tell how it was obtained.

pvstackr checks the $\hat k$ values against one cut-off, `psis_k_threshold`
in `pv_control()`: 0.7 by default, which is also the largest value allowed. A
`stack_psis` fit is blocked unless the $\hat k$ of every plausible value is
below it. [Choosing a method and reading its
intervals](m5-methods-and-coverage.html#psis) gives the cut-offs of the
literature and how to set `psis_k_threshold` to one of them.

## Cautions in reading a comparison {#cautions}

### Standard errors from the model and from the replicate weights

Fits that agree on the estimates can still differ in their standard errors,
because the methods take the variance within a plausible value from different
sources. The target of `stack_direct` takes it from the BRR–Fay replicate
weights, which PISA provides for estimating the sampling variance under its
design. In such a design, clustering (students sampled within schools) and
unequal weights usually increase the sampling variance, while stratification
reduces it [@oecd2024pisa]. `per_pv` and `stack_psis` take the variance from
the posterior draws of a model. The draws reflect the clustering only as far
as the model includes it (a single-level regression treats the students as
independent), and weighting the likelihood by the survey weights does not by
itself make their variance design-based [@williams2021uncertainty]. The two
variances can therefore
differ even when the estimates agree. The example cannot show this, because
its `per_pv` standard errors come from the made-up draws.

This difference is the reason for pvstackr's reporting rule in [the three
methods](#methods-table). Close agreement between the estimates of two methods does not tell
you which standard errors to report.

### Pareto $\hat k$ values check only the importance weights

For a `stack_psis` fit, a $\hat k$ below the cut-off of @vehtari2024pareto,
$\min(1 - 1/\log_{10} S, 0.7)$ for $S$ draws, indicates that the importance
ratios for that plausible value pass this diagnostic; the reliability of a
weighted mean or covariance depends on the Pareto $\hat k$ of that quantity,
which can be larger [@vehtari2024pareto, pp. 7, 14]. With fewer than about
2,000 draws this cut-off is below pvstackr's default of 0.7. It does not
change where the variance comes
from: the standard errors of a `stack_psis` fit still come from the posterior
draws of the stacked model, as in the previous caution.

pvstackr cannot check the $\hat k$ values either. It computes neither them nor
the weights, so it asks you to name the program that produced them, with
`psis_producer` and `psis_producer_version`, and records the name without
checking it: a producer name alone is not evidence that PSIS was run. In the
example the value 0.2 was typed in and no program is named, so the fit stays
blocked.

$\hat k$ also depends on the data, the model and the posteriors that the
weights aim at, so check it in every analysis. In the application of the
companion preprint to PISA 2022, one stacked fit per country was reweighted,
in the same way in both countries, toward the posteriors of a model with a
school random intercept, with the survey weights placed in two stages; every
$\hat k$ was below 0.7 in the United States and above it in Korea
[@lee2026onemcmc, sec. 6.4 and app. D.1]. pvstackr 0.2.x does not fit that
model, so these values are not direct evidence about `stack_psis` fits of the
single-level model that pvstackr fits ([Choosing a method and reading its
intervals](m5-methods-and-coverage.html) gives the details).

## Choosing a method {#choosing}

pvstackr recommends the following (a package rule, not a result of the
companion preprint):

- Report a `stack_direct` fit. Its estimate table holds the numbers of the
  design-based target, its intervals are the only ones that can be
  coverage-claimable (with a Barnard–Rubin target), and it needs one Bayesian
  fit.
- Use `per_pv` to compare the `stack_direct` estimates with one Bayesian fit
  per plausible value. Its intervals are descriptive, and pvstackr does not
  check the convergence of its fits.
- Use `stack_psis` only as a check on a stacked fit, with importance weights
  and $\hat k$ values from a PSIS program. Its intervals are descriptive as
  well.

The companion preprint notes that for a single, final, high-stakes model one
fit per plausible value remains a sound default [@lee2026onemcmc, sec. 7].

[Reading and reporting the results](a3-reading-results.html#interval-columns)
explains the interval labels of every method and what to report for a fit.
Among the Method articles, [Choosing a method and reading its
intervals](m5-methods-and-coverage.html) compares the methods and their
intervals in more detail, and [One stacked fit and the Rubin point
estimate](m3-stacked-bridge.html) explains the stacked fixed-effect point
identity, whose proof is in the preprint.

## References
