---
title: "A4: Comparing methods — choosing stack_direct, per_pv, or stack_psis"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  pvstackr ships three fitting methods. This vignette uses `pv_compare_methods()`
  to align `stack_direct`, `per_pv`, and `stack_psis` on the same data and shows
  how to read the agreement diagnostics. It states the two cautions plainly —
  design variance is not optional, and a small Pareto-k does not imply a correct
  `stack_psis` variance — and gives the rule of thumb: use `stack_direct`, keep
  `stack_psis` as a cross-check, and treat `per_pv` as the reference.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{A4: Comparing methods — choosing stack_direct, per_pv, or stack_psis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

Vignettes A1–A3 lived entirely inside one method: load a `stack_direct` fit, read
its estimate table, and decide what is safe to report. This vignette steps back
to the choice *between* methods. pvstackr ships three — `stack_direct`, `per_pv`,
and `stack_psis` — and `pv_compare_methods()` aligns them on the same
fixed-effect terms so you can read them side by side. The point of the article is
not to crown a winner on a toy dataset; it is to show the **comparison machinery**
and the **interval semantics** that separate the methods, and to state the two
cautions that govern when agreement means something.

::: {.alert .alert-warning}
A note on the worked example below. To compare three methods offline we feed
`per_pv` and `stack_psis` **synthetic, injected** posterior draws (random noise
centred near the fixture's coefficients), because the package ships only one
cached fit and runs no live MCMC. Their point estimates are therefore *manufactured
for illustration* and will not coincide with `stack_direct` the way a real
comparison would. Read this article for the **structure** of a comparison and the
**interval semantics**, not as a substantive agreement result.
:::

## 1. The three methods in one table

The three methods answer the same question — *what is the fixed-effect block
$\beta_{\text{FE}}$?* — but they get their **variance** from different places, and
that difference is the whole story of when a row may carry a coverage claim.

| Method | Covariance basis | Interval status | Fit count |
|---|---|---|---|
| `stack_direct` *(default)* | design-based external $T_{\text{MI}}$ (Rubin / BRR–Fay) | **conditional** — coverage-claimable only with an external Barnard–Rubin BRR–Fay target; descriptive under classic df | **1** |
| `per_pv` | model-based posterior | descriptive / reference | **$M$** |
| `stack_psis` | model-based posterior **+ PSIS** reweighting | descriptive / reference | **1** (+ PSIS) |

In one sentence each:

- **`stack_direct`** fits one stacked model and calibrates it (via CCC) to an
  **external, design-based** Rubin / BRR–Fay target. Because that target's
  covariance is design-based, `stack_direct` rows are the only ones the package
  ever marks `coverage_claim_allowed = TRUE` [@judkins1990fay; @rubin1987multiple].
- **`per_pv`** is the orthodox path: fit the model once per plausible value and
  pool the $M$ model-based results with Rubin's rules. It is the natural
  **reference**, but its covariance is *model-based*, so its intervals stay
  descriptive.
- **`stack_psis`** reweights a single stacked draw cloud with per-PV
  Pareto-smoothed importance sampling [@vehtari2017practical; @vehtari2024pareto],
  pooling the PSIS-weighted summaries. One fit, model-based covariance — a
  **cross-check**, never the deliverable.

::: {.alert .alert-warning}
The **"fit count" column is topology, not speed.** It records how many model fits
a method's architecture requires — one stacked fit versus $M$ separate ones — not
a benchmarked runtime. v0.1 makes no "faster" claim; see the honesty triad in
Section 7.
:::

Two scope reminders carried over from A1–A3, because they bound this whole
comparison: the reportable surface is **fixed effects only**, and **coverage
lives on exactly one path** (`stack_direct` against an external Barnard–Rubin
BRR–Fay target). Everything below respects both.

## 2. Run a comparison — `pv_compare_methods()`

`pv_compare_methods()` takes two or more already-built `pvstackr_fit` objects and
aligns their fixed-effect estimate tables. It does **not** fit anything — you hand
it finished fits — so the only setup is building three fits that share the same
fixed-effect names (`b_Intercept`, `b_x`, `b_female`).

The first fit is the cached `stack_direct` fixture you already met in A1. The
other two are built from **injected synthetic draws**: `pv_fit_reference()` accepts
a list of per-PV posterior draw matrices, and `pv_fit_stack_psis()` accepts a
stacked draw matrix with supplied PSIS weights and Pareto-$\hat k$ values. No live
MCMC runs anywhere below.

```{r build-fits}
fe <- c("b_Intercept", "b_x", "b_female")

fit_direct <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit

# Synthetic per-PV draw clouds, centred near the fixture's coefficients
# (~458 / 47 / 2). These are random noise for illustration, NOT real posteriors.
set.seed(1)
mk <- function(mu) {
  d <- matrix(rnorm(300 * 3, sd = 2), ncol = 3, dimnames = list(NULL, fe))
  sweep(d, 2L, mu, "+")
}

fit_per_pv <- pv_fit_reference(
  per_pv_draws = list(PV1READ = mk(c(458, 47, 2)),
                      PV2READ = mk(c(458, 47, 2))),
  control      = pv_control(method = "per_pv")
)

# Synthetic stacked draw cloud + equal (placeholder) PSIS weights and small
# Pareto-k. Again injected for illustration; nothing is importance-sampled live.
sd_mat <- mk(c(458, 47, 2))
fit_psis <- pv_fit_stack_psis(
  stacked_draws = sd_mat,
  pv_cols       = c("PV1READ", "PV2READ"),
  psis_weights  = matrix(1 / nrow(sd_mat), nrow = nrow(sd_mat), ncol = 2),
  pareto_k      = c(0.2, 0.2),
  control       = pv_control(method = "stack_psis")
)
```

With three fits in hand, the comparison is one call:

```{r compare}
cmp <- pv_compare_methods(
  stack_direct = fit_direct,
  per_pv       = fit_per_pv,
  stack_psis   = fit_psis
)

cmp
```

The `print()` box names the **reference method** (here `per_pv`, the default when a
non-blocked `per_pv` fit is present), lists the three labelled methods, and
confirms the comparison aligned on **3 shared fixed-effect terms**. It then prints
two honest one-liners that recur throughout this article:

- a **provenance note** — agreement bands are *descriptive* and shared
  target/pooling/source metadata is not independent corroboration (Section 4); and
- an **interval note** — at least one compared interval is *descriptive rather
  than coverage-claimable* (Sections 3, 5–7).

Comparison aligns methods on their **shared** fixed-effect terms: every method
here reports `b_Intercept`, `b_x`, and `b_female`, so all three line up cleanly. A
method missing a term would still get a row, with its numeric fields left `NA`
rather than silently dropped.

## 3. The three methods side by side

`get_estimates()` on a comparison returns the **aligned** table: one row per
(method, term) pair — here $3 \times 3 = 9$ rows. The slice that matters for
choosing a method puts the point estimate next to its interval and the
interval's licensing metadata:

```{r aligned-estimates}
est <- get_estimates(cmp)

est[, c("method_label", "term", "estimate", "se",
        "conf_low", "conf_high",
        "interval_role", "coverage_claim_allowed")]
```

Read this table on two axes.

**The points do not match here — and that is the synthetic setup, not a finding.**
The three methods' estimates differ by more than they should because the `per_pv`
and `stack_psis` rows were built from *independent random noise*, not from the
same data as `stack_direct`. In a **real** comparison, `stack_direct` and `per_pv`
points agree closely: this is the **stacked-MLE point identity** — one stacked,
$1/M$-weighted fit recovers the Rubin-pooled per-PV point estimate. That identity
is *exact* only under its regularity conditions (flat prior, linear-Gaussian,
common design, a common plug-in covariance) and holds **approximately, not
exactly**, in production with REML variance components and real PISA data. The
precise statement and its caveat are Theorem 2.2 in the **Method track, M3**.

**The interval semantics are the real teaching point.** Even where points agree,
two columns separate the methods and never collapse:

- **`interval_role`** records the *kind* of interval — `descriptive_classic_rubin`
  for `stack_direct` (this classic-df fixture), `reference_classic_rubin` for
  `per_pv`, and `psis_classic_rubin` for `stack_psis`.
- **`coverage_claim_allowed`** is `FALSE` for **every** row here: the fixture's
  `stack_direct` target uses *classic* Rubin df (so it is descriptive by design,
  as in A2/A3), and `per_pv` / `stack_psis` are *always* descriptive regardless of
  their df.

Interval **width** also differs sharply, and width is where the methods genuinely
diverge. Look at the slope on `x`:

```{r interval-widths}
bx <- est[est$term == "b_x",
          c("method_label", "se", "conf_low", "conf_high")]
bx$width <- bx$conf_high - bx$conf_low
bx
```

The three widths are not interchangeable — and in this synthetic run the
model-based methods happen to produce *wider* intervals than the calibrated
`stack_direct` fit. Do not read that ordering as a general result: it is an
artifact of the injected draw clouds (the `per_pv` and `stack_psis` clouds carry
their own noisy spread). The durable lesson is that **width and the coverage flag
are method-specific** — comparing intervals across methods means comparing
*different kinds of object*, which is exactly what `interval_role` is there to flag.

A figure makes the per-term overlay easy to read. We plot the two slope
coefficients (dropping the intercept, whose ~458 scale would flatten the axis) and
overlay all three methods, each offset slightly so the intervals do not collide:

```{r overlay-figure, fig.width = 7, fig.height = 3.4, fig.cap = "The two slope coefficients (b_x, b_female) under all three methods, with their 95% intervals overlaid. The dashed line marks zero. Because the per_pv and stack_psis rows are built from synthetic injected draws, their points and widths are illustrative only — the figure shows the comparison structure and the per-method interval semantics, not a substantive agreement result.", fig.alt = "A horizontal dot-and-interval plot with two rows, one for the coefficient on x near 47 and one for the coefficient on female near 2. Within each row the three methods stack_direct, per_pv, and stack_psis are drawn at slightly offset heights in different colours, each with its own 95 percent interval. A dashed vertical reference line is drawn at zero."}
slopes  <- est[est$term != "b_Intercept",
               c("method_label", "term", "estimate", "conf_low", "conf_high")]
methods <- unique(slopes$method_label)
terms   <- sort(unique(slopes$term))

offsets <- setNames(seq(-0.18, 0.18, length.out = length(methods)), methods)
cols    <- setNames(c("#1f6f9c", "#f2c14e", "grey45")[seq_along(methods)], methods)

xlim <- range(c(slopes$conf_low, slopes$conf_high, 0), na.rm = TRUE)

op <- par(mar = c(4.5, 7, 1, 1))
plot(
  NA, xlim = xlim, ylim = c(0.5, length(terms) + 0.5),
  yaxt = "n", ylab = "",
  xlab = "Coefficient (synthetic reading-score points)"
)
abline(v = 0, lty = 2, col = "grey50")
for (mlab in methods) {
  sub <- slopes[slopes$method_label == mlab, ]
  yy  <- match(sub$term, terms) + offsets[[mlab]]
  segments(sub$conf_low, yy, sub$conf_high, yy, lwd = 2, col = cols[[mlab]])
  points(sub$estimate, yy, pch = 19, cex = 1.3, col = cols[[mlab]])
}
axis(2, at = seq_along(terms), labels = terms, las = 1)
legend("topright", legend = methods, col = cols, pch = 19, lwd = 2,
       bty = "n", cex = 0.9)
par(op)
```

The intervals all comfortably span zero for `b_female` and all sit clear of zero
for `b_x` — but their *widths* differ by method, and only the `stack_direct` row
could ever (with a Barnard–Rubin target) carry a coverage claim. That last point
is the bridge to the cautions.

## 4. Agreement diagnostics

`get_diagnostics()` on a comparison returns a named list. Its keys:

```{r diag-keys}
dg <- get_diagnostics(cmp)
names(dg)
```

Two of these carry the interpretive weight.

**`agreement`** summarises, per method, how far each method's rows sit from the
reference (`per_pv` here). The two columns to read are `max_abs_z_diff` (the
largest standardised point gap from the reference, across terms) and
`max_abs_log_se_ratio` (the largest log standard-error ratio — i.e. how differently
the method spreads its intervals):

```{r agreement}
dg$agreement
```

The reference compares to itself, so its `max_abs_z_diff` and
`max_abs_log_se_ratio` are exactly zero. For `stack_direct` and `stack_psis` the
*point* gaps are large here only because the injected draws are synthetic; in a
real comparison the informative pattern is **agree on points, disagree on
variance** — methods that share a target tend to match on the point (nearly an
algebraic identity) while differing on the variance, which comes from genuinely
different machinery. The SE-ratio column is the one that correctly reflects three
different covariance bases.

**`target_overlap`** is the honesty gate on reading agreement as corroboration. It
flags whether the compared methods share provenance — an external target, a target
hash, a pooling hash, or a target-source family:

```{r target-overlap}
str(dg$target_overlap)
```

In this synthetic run `independence_caveat_required` is `FALSE`, because the three
injected fits happen to carry **distinct** hashes and only `stack_direct` holds the
`external_brr_fay_rubin` source. But the field exists precisely for the common case
where it flips to `TRUE`: when two methods are calibrated to the *same* external
target (or pool the same draws), close agreement is **mechanically guaranteed** and
tells you nothing about correctness. The `independence_caveat` string spells this
out, and it is the same message the `print()` box surfaces.

The per-method diagnostic table carries the rest, including the PSIS row's
Pareto-$\hat k$:

```{r method-diagnostics}
dg$method_diagnostics[, c("method_label", "interval_role",
                          "coverage_claim_allowed", "target_source",
                          "psis_status", "pareto_k_max", "n_fits")]
```

A few reads from this table. `target_source` is **provenance vocabulary, not a
formal target object**: the `stack_psis` row reads `stack_psis_rubin_pooling`, but
`get_target()` returns `NULL` for a `stack_psis` fit — the label records how the
covariance was built, not that a target object exists. The `n_fits` column is the
topology count from Section 1 (`stack_direct` = 1, `per_pv` = $M$ = 2,
`stack_psis` = 1). And for `stack_psis`, `psis_status` and `pareto_k_max` carry the
PSIS diagnostics: here `status == "ok"` with `pareto_k_max = 0.2`, comfortably
inside the **Vehtari et al. thresholds** [@vehtari2024pareto] —

- $\hat k < 0.5$: **good**;
- $0.5 \le \hat k < 0.7$: **borderline**;
- $\hat k \ge 0.7$: **unreliable**.

::: {.alert .alert-warning}
**Agreement is descriptive, never independent corroboration when methods share a
target, pooling, or source.** Three methods reading the *same* information are
expected to agree; their concordance is then uninformative about whether the
shared answer is right. Read `target_overlap` before you read any agreement band
as reassurance. Section 6 shows how this bites `stack_psis` specifically.
:::

## 5. Caution 1 — design variance is not optional

The first caution is why `per_pv` and `stack_psis` are *descriptive* while only
`stack_direct` is coverage-claimable. It is a statement about **variance**, not
about points.

A weighted analysis can get the **point** estimate exactly right and still report
intervals that are far too narrow, because **model-based standard errors ignore the
survey design**. Complex PISA sampling — stratification, clustering, unequal
selection — inflates the true sampling variance above what an i.i.d.-style
model-based posterior assumes. The fix is to take the variance from the design, not
the model: the **BRR–Fay** replicate-weight sandwich, pooled across plausible
values into the Rubin total $T_{\text{MI}} = \bar U + (1 + 1/M)\,B$, restores
nominal coverage [@judkins1990fay; @rubin1987multiple]. That is exactly the
external target `stack_direct` calibrates to.

::: {.alert .alert-info}
The illustrative evidence for "model-based SEs under-cover; design-based BRR–Fay
$T_{\text{MI}}$ restores nominal coverage" comes from the **companion methods
paper (in preparation)**. It is described here qualitatively and deliberately
carries **no hard-coded numbers** — the package neither bundles real PISA nor
presents simulation results as its own output.
:::

This is the whole reason the package draws the line where it does. `per_pv` and
`stack_psis` produce *model-based* covariances, so however sensible their points,
their intervals cannot bear a design-coverage claim — the package marks them
descriptive. Only a `stack_direct` fit calibrated to the **external Barnard–Rubin**
BRR–Fay target earns `coverage_claim_allowed = TRUE`. (And, as A2/A3 stressed, even
`stack_direct` stays descriptive under *classic* df — as on this fixture.)

## 6. Caution 2 — small Pareto-$\hat k$ ≠ correct variance

The second caution is the trap specific to `stack_psis`, and it is subtle because
the diagnostic that looks reassuring can be reassuring about the *wrong thing*.

A small Pareto-$\hat k$ tells you the **importance-sampling step is stable** — the
weights are well-behaved. It does **not** tell you the resulting **variance** is
right. The reason is structural: `stack_psis` reweights **one** stacked draw cloud
to stand in for all $M$ plausible values, so the per-PV "imputations" are drawn
from the *same* cloud and are therefore **correlated**. Rubin's between-imputation
variance $B$ assumes independent imputations; correlated ones shrink $B$, and the
pooled `stack_psis` interval can run **narrow even when $\hat k$ is small**.

Worse, $\hat k$ is **specification-dependent**. It can sit comfortably below $0.5$
in a simplified, low-dimensional reading demo and climb **above $1$** in a full
production specification — the regime where PSIS is formally unreliable. A green
$\hat k$ on a toy model is no guarantee of a green $\hat k$ on the model you
actually report.

::: {.alert .alert-warning}
Treat `stack_psis` as a **cross-check, never the deliverable.** A small
Pareto-$\hat k$ is necessary but **not sufficient** for a trustworthy interval;
the correlated-imputation narrowing and the specification dependence of $\hat k$
are described qualitatively in the **companion methods paper (in preparation)**.
The package hard-codes **no** $\hat k$ values as facts — the `pareto_k = 0.2` used
above is an injected illustration, nothing more
[@vehtari2017practical; @vehtari2024pareto].
:::

## 7. Which to use (rule of thumb)

The choice collapses to one sentence per method:

- **Report `stack_direct`.** It is the only path that is coverage-claimable (when
  calibrated to an external Barnard–Rubin BRR–Fay target), from **one** stacked
  fit. This is the deliverable.
- **Use `stack_psis` as a one-fit cross-check.** Watch the Pareto-$\hat k$, and —
  per Caution 2 — *do not trust a small $\hat k$ alone*. Never report it as the
  headline.
- **Use `per_pv` as the orthodox reference.** The classic fit-it-$M$-times-and-pool
  answer is the natural sanity check on `stack_direct`'s point estimate; its
  intervals stay descriptive.

And the **honesty triad** that governs every claim in this article:

1. **Fixed effects only.** v0.1 calibrates and reports $\beta_{\text{FE}}$;
   variance components are out of scope.
2. **Coverage on one path only.** `coverage_claim_allowed = TRUE` requires
   `stack_direct` **and** an external Barnard–Rubin BRR–Fay target; `per_pv` and
   `stack_psis` are always descriptive.
3. **No speed claims.** "One fit" is *topology* — an architecture statement, not a
   benchmarked runtime. v0.1 makes no efficiency claim.

## 8. Where to next

You can now align three methods and read their agreement honestly. From here:

- **A3 · Reading results & what to report** — the interval-metadata columns
  (`interval_role`, `coverage_claim_allowed`, `df_method`, `df_complete`) read here
  across methods, treated in depth for a single fit, plus the fraction of missing
  information and a reporting checklist.
- **A5 · Real PISA data guidance** — running these comparisons on genuine PISA
  files: licensing and non-affiliation, the design declaration, and the
  memory/runtime cost of $M(R+1)$ replicate fits.

For the underlying theory, the **Method track** carries the proofs this article
gestures at: **M5 · Methods, PSIS, and coverage** states the precise method
contrast, the PSIS weights and Pareto-$\hat k$ thresholds, and *why only
`stack_direct` is coverage-claimable*; **M3 · The stacked fractional bridge** proves
the stacked-MLE point identity (and where it stops). Start the track at **M1 ·
Foundations and notation**.

Bug reports and feature requests:
<https://github.com/joonho112/pvstackr/issues>.

### Session info

```{r session-info}
sessionInfo()
```

## References
