---
title: "A5: Real PISA data — licensing, memory, and reproducibility"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  How to connect pvstackr to real PISA data without bundling licensed survey
  files. This guidance covers licensing and non-affiliation, declaring the
  plausible-value design on a real cycle, the target and fit shape, the memory and
  runtime footprint (about N times M stacked rows, and about M times R-plus-one
  target fits), and a reproducibility checklist. It is guidance, not a worked
  substantive analysis; the package ships no real PISA records.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{A5: Real PISA data — licensing, memory, and reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
# This is a GUIDANCE vignette. The package ships NO real PISA data (the bundled
# pisa_tiny fixture is synthetic), so every illustrative real-PISA chunk below is
# set eval = FALSE GLOBALLY here and shows only the shape of the call. The closing
# session-info chunk re-enables eval = TRUE explicitly so it still runs.
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  out.width = "85%",
  eval      = FALSE
)
```

```{r library}
library(pvstackr)
```

## 1. Purpose

Vignettes A1–A4 ran end to end on the bundled **synthetic** `pisa_tiny` fixture:
you loaded a cached fit (A1), built the pipeline that creates one (A2), read what
is safe to report (A3), and compared the three methods (A4). This article is
different in kind. It is **guidance** for pointing the same workflow at a
**real** PISA cycle on your own machine — it is not a worked substantive analysis,
and it computes no real-data numbers.

The reason is deliberate and worth stating plainly:

- **pvstackr does not ship real PISA records.** The only example data in the
  package is the synthetic `pisa_tiny` fixture (one make-believe country
  `CNT == "SYN"`, $M = 2$ plausible values, $R = 4$ replicate weights), built
  solely so examples and package checks run offline without licensed files.
  Nothing in this vignette bundles, mirrors, or reproduces genuine PISA microdata.
- **pvstackr is not affiliated with or endorsed by OECD/PISA.** In package
  documentation "PISA" is used **descriptively**, for the survey-data and
  column-name conventions that applied users bring to their own authorized
  analyses (`PV1MATH`, `W_FSTUWT`, `W_FSTURWT1`, `CNTSTUID`, and so on)
  [@oecd2024pisa].

Because the package has no real data to run, **every real-PISA code chunk below is
`eval = FALSE`**: it shows the *shape* of the call you would make on your own
authorized extract, never a live result. Only the setup, the `library()` line, and
the closing session-info chunk actually execute.

The road map for this article: data access and licensing → declare the design on
a real cycle → the target and fit shape → memory and runtime → a reproducibility
checklist → the companion methods paper, described qualitatively.

## 2. Data access and licensing

### 2.1 Download only from an authorized source

Obtain real PISA files **only** from an authorized source — the official OECD PISA
release portal or an institutionally approved mirror. Before you use a file,
review the **license, citation, redistribution, and privacy terms** for the
**specific cycle and files** you download: these vary by cycle and release, and
the terms attached to the files govern, not any convention assumed here
[@oecd2024pisa]. The plausible-value design that the data encode — $M$ draws per
student from a latent-proficiency posterior — is itself a property of how PISA is
constructed and scored [@mislevy1991randomization; @vondavier2009plausible].

### 2.2 The package license does not license your data

The **MIT license for `pvstackr` does not license users' real PISA data**, derived
extracts, or any third-party files you process with it. The package license covers
the *software*; your *data* are governed by the cycle- and file-specific terms of
your source. Follow those terms in full — attribution requirements, redistribution
limits, third-party-rights rules, and privacy obligations all remain yours to
honour.

### 2.3 Keep microdata, extracts, and caches out of the repository

Do **not** commit real PISA microdata, large derived extracts, or backend caches
to a source-controlled package or project repository. Keep them outside the source
tree — for example under a local directory referenced by an environment variable,
so paths never leak into version control:

```{r data-dir, purl = FALSE}
pisa_dir <- Sys.getenv("PISA_DATA_DIR")
stopifnot(nzchar(pisa_dir))

student_file <- file.path(pisa_dir, "student_file_for_your_cycle")
```

### 2.4 Read with the reader for your format, and keep only what you need

File names and formats vary by cycle and release. Use the reader that matches your
**authorized** file format — an SPSS, SAS, transport, or converted local extract —
and then select only the columns the model needs: the country filter, the
plausible values, the final weight, the replicate weights, the identifiers, and
the covariates. Narrowing the columns at read time is the single cheapest thing
you can do for the memory footprint discussed in Section 5.

```{r read-extract, purl = FALSE}
# Example only: replace this with the reader for YOUR authorized file format,
# e.g. haven::read_sav() for SPSS, a SAS/transport reader, or a converted CSV.
# pvstackr does not ship a PISA reader and does not bundle any real PISA file.
pisa_raw <- haven::read_sav(student_file)

pisa_country <- subset(pisa_raw, CNT == "USA")
pisa_country <- pisa_country[
  ,
  c(
    "CNT",
    "CNTSCHID",
    "CNTSTUID",
    "ESCS",
    "ST004D01T",
    paste0("PV", 1:10, "MATH"),   # M = 10 plausible values (mathematics)
    "W_FSTUWT",                    # final student weight
    paste0("W_FSTURWT", 1:80)     # R = 80 BRR replicate weights
  )
]
```

## 3. Design declaration

Declare the plausible-value design **before** fitting, exactly as in A2 — only now
the arguments describe a real cycle. The single most important habit here is to
**verify every structural argument against the official technical documentation for
that cycle**: the plausible-value suffix, the plausible-value count `expected_M`,
the replicate-weight count `expected_R`, the Fay coefficient `fay_k`, and the
identifier columns `id_cols`. PISA mathematics commonly uses M = 10, R = 80, and
Fay $k = 0.5$ [@oecd2024pisa; @judkins1990fay], but you confirm these per cycle
rather than assume them — `pv_design()` asserts the counts you pass, so a mismatch
surfaces immediately instead of silently mis-pooling.

```{r design, purl = FALSE}
design <- pv_design(
  data       = pisa_country,
  formula    = OUTCOME ~ ESCS + ST004D01T,   # OUTCOME is a placeholder
  pv_suffix  = "MATH",                        # matches PV1MATH ... PV10MATH
  expected_M = 10L,                           # assert exactly 10 plausible values
  expected_R = 80L,                           # assert exactly 80 replicate weights
  fay_k      = 0.5,                           # Fay coefficient for this cycle
  id_cols    = "CNTSTUID"                     # column(s) identifying a unique row
)

design
```

Three points about the call shape, carried over from A2 and specialised to a real
cycle:

- **`OUTCOME` is a placeholder, not a column.** pvstackr substitutes each
  plausible-value column (`PV1MATH`, …, `PV10MATH`) in turn when it builds the
  target and the stacked data, so you write the model once rather than $M$ times
  [@mislevy1991randomization; @vondavier2009plausible].
- **`pv_suffix = "MATH"` selects the domain.** Modern subject-suffixed columns such
  as `PV1MATH` require the matching suffix. Leaving `pv_suffix = ""` (the bare
  names such as `PV1`, `PV2`, … default) is appropriate **only** when your analytic
  extract has *intentionally renamed* the plausible values to bare names; on an
  official file it would fail to detect the domain block.
- **No `weights()` in the formula.** Survey weights are *design* information, not
  model terms. Pass the final weight and the replicate weights through the design
  and target arguments (auto-detected here from the `W_FSTUWT` / `W_FSTURWT*`
  columns), never inside the formula.

## 4. Target and fit shape

This section mirrors A2 (§3–6) and A3, specialised to a real cycle; consult those
vignettes for the line-by-line reading of the objects.

### 4.1 stack_direct requires an external BRR–Fay target

A reportable `stack_direct` fit requires an **external**
Rubin / BRR–Fay fixed-effect target built by `pv_brr_target()`. The target is
**fixed-effect-only**: it forms a BRR–Fay sandwich covariance per plausible value
and pools the $M$ results with Rubin's rules into the total
$T_{\text{MI}} = \bar U + (1 + 1/M)\,B$ [@rubin1987multiple; @barnard1999small].
Random-effect (group) terms such as `(1 | CNTSCHID)` or `(1 || CNTSCHID)` are
**not accepted by the BRR-Fay target engine**, by design — see A2 §3 for the
error this raises.

```{r target, purl = FALSE}
target <- pv_brr_target(
  data            = design$data,
  formula         = design$formula,
  pv_cols         = design$pv_cols,
  weight_col      = design$weight_col,
  rep_weight_cols = design$rep_weight_cols,
  fay_k           = design$fay_k,
  id_cols         = design$id_cols
)

target
```

### 4.2 Fit shape — `center = "target"` and the backend adapter

With a design and a target in hand, `pv_fit(method = "stack_direct")` fits **one**
stacked model on the N * M rows and calibrates it (via CCC) to the external target.
On real data you either ask for the bundled backend with
`pv_control(backend = "brms")`, or supply your own engine through the three
adapter slots (`fit_function`, `draws_function`, `diagnose_function` — all three
are required for a reportable fit). This vignette runs **no** live backend, so
the chunk shows only the shape; the adapter contracts are sketched in A2 §7.

```{r fit, purl = FALSE}
fit <- pv_fit(
  data           = design$data,
  formula        = design$formula,
  target         = target,
  method         = "stack_direct",
  control        = pv_control(
                     method           = "stack_direct",
                     backend          = "injected",
                     center           = "target",     # required for a reportable fit
                     iter             = 2000L,
                     warmup           = 1000L,
                     chains           = 4L,
                     cores            = 4L,
                     seed             = 20260607,
                     return_draws     = FALSE,
                     keep_data        = FALSE,
                     keep_backend_fit = FALSE,
                     keep_log_lik     = FALSE
                   ),
  fit_function   = your_fit_function,    # your modelling backend (see A2 §7)
  draws_function = your_draws_function   # extract posterior draws for CCC
)

summary(fit)
get_estimates(fit)
```

### 4.3 How to read the result

Read `get_estimates(fit)` exactly as in A3. For the default `control$center =
"target"` path, the reportable fixed-effect estimate and standard error are the
external target values: the estimate is the target coefficient after CCC centering,
and the standard error is $\sqrt{\operatorname{diag}(T_{\text{MI}})}$ with the
target's degrees of freedom and interval metadata. The stacked fit supplies the
calibrated draw cloud (when retained) and the center-separation agreement
diagnostics — it is **not** the source of the headline fixed-effect numbers, and
must not be read as replacing the external target. Reportable `stack_direct` fits
require `control$center = "target"`; posterior-centered CCC output is
diagnostic/exploratory only, because it keeps the raw stacked posterior center
while borrowing the external covariance (A2 §6).

Use the interval-metadata columns when interpreting results (A3 §4). In this
package stage, `coverage_claim_allowed = TRUE` is reserved for external
Barnard–Rubin BRR–Fay `stack_direct` target rows
[@barnard1999small; @rubin1987multiple; @judkins1990fay]; `per_pv` and
`stack_psis` intervals are descriptive/reference intervals even when their Rubin
pooling uses Barnard–Rubin df and `df_complete`. Read the `coverage_claim_allowed`
column rather than inferring from the method name.

## 5. Memory and runtime

Start with **one country, one domain, and a small fixed-effect formula**, and keep
only the columns the analysis needs (Section 2.4). Two footprint facts then let you
size a job before you launch it. Both are statements about the **footprint and
topology** of the computation — *not* speed or "faster" claims (the honesty triad,
Section 7).

**Stacked-data preparation expands the row count.** Each plausible value
contributes a copy of the data at weight $1/M$, so a design with `N` rows and `M`
plausible values produces about `N * M` stacked rows before the live Bayesian fit.
A modest national sample at M = 10 therefore presents the backend with roughly ten
times its raw row count.

**The target step runs many weighted fixed-effect fits.** `pv_brr_target()`
estimates a weighted fixed-effect fit for the point and for each of the `R`
replicate weights, across all `M` plausible values: about `M * (R + 1)` weighted
fits per model. With common PISA settings (M = 10, R = 80) that is about
`10 * 81 = 810` weighted least-squares fits before any live Bayesian sampling. This
stage is ordinary weighted regression, but on a large national extract it is still
worth validating on a reduced country/domain slice first.

The live backend usually dominates the footprint. Practise backend hygiene:

- **set a seed and record every control setting** (Section 6), so a run is exactly
  reproducible;
- **write backend caches outside the repository**, alongside the microdata
  (Section 2.3);
- keep **`keep_data = FALSE`**, **`keep_backend_fit = FALSE`**, and
  **`keep_log_lik = FALSE`** unless you specifically need those artifacts — each
  one can dominate the size of a saved fit;
- use **`return_draws = FALSE`** for lightweight reporting when calibrated draws
  are not needed downstream (the fixed-effect report is fully determined by the
  target, A3 §6);
- **increase `chains`, `iter`, and `cores` only after the design and target hashes
  are stable** — scaling the backend before the inputs are frozen wastes the most
  expensive part of the run.

## 6. Reproducibility checklist

Record these items with each real-data analysis, so a reviewer can tell a
reproducible applied workflow from the tiny synthetic examples used in package
tests. The hashes are stable content fingerprints carried on the design, target,
and estimate objects (A2 §2–3, A3 §3) — recording them lets you prove which inputs
produced which numbers.

- **Cycle provenance.** PISA cycle and release date; the authorized data source.
- **Filters.** Country / domain filters applied to the extract.
- **Variables.** The model variables and the plausible-value suffix (`pv_suffix`).
- **Weights.** The final weight, the replicate-weight prefix and count, and the
  Fay coefficient `fay_k`.
- **Package.** The `pvstackr` package version
  (`utils::packageVersion("pvstackr")`).
- **Design / target hashes.** `design_hash`, `row_support_hash`, `pv_value_hash`,
  `weight_design_hash`, and the target's `target_hash`.
- **Backend.** The backend, the seed, and `chains`, `iter`, `warmup`, and `cores`.
- **Retention flags.** Whether reportable draws, backend fits, data, or
  log-likelihoods were retained (`return_draws`, `keep_backend_fit`, `keep_data`,
  `keep_log_lik`).

```{r repro-record, purl = FALSE}
# Sketch only (eval = FALSE): assemble a compact provenance record from the
# objects above. Adapt the field names to your project's logging conventions.
provenance <- list(
  pisa_cycle         = "2022",                  # cycle + release date
  data_source        = "OECD PISA portal",      # authorized source
  filters            = "CNT == 'USA'; domain = MATH",
  variables          = all.vars(design$formula),
  pv_suffix          = "MATH",
  weight_col         = design$weight_col,
  rep_weight_prefix  = "W_FSTURWT",
  R                  = length(design$rep_weight_cols),
  fay_k              = design$fay_k,
  pkg_version        = as.character(utils::packageVersion("pvstackr")),
  design_hash        = design$design_hash,
  row_support_hash   = design$row_support_hash,
  pv_value_hash      = design$pv_value_hash,
  weight_design_hash = design$weight_design_hash,
  target_hash        = target$target_hash,
  backend            = "injected",
  seed               = 20260607,
  chains             = 4L,
  iter               = 2000L,
  warmup             = 1000L,
  cores              = 4L,
  return_draws       = FALSE,
  keep_backend_fit   = FALSE,
  keep_data          = FALSE,
  keep_log_lik       = FALSE
)

str(provenance)
```

These records separate a reproducible applied workflow from the tiny synthetic
examples used in package tests — and they are what the `target_hash` and friends
read off in A3 §3 are ultimately *for*.

## 7. The companion methods paper (qualitative)

The substantive PISA findings — the empirical evidence that motivates the package —
live in a **companion methods paper, currently in preparation**, not in
`pvstackr`. The package neither bundles real PISA nor presents contested
coefficients as its own output, so this section describes that work **qualitatively
and without a formal citation**, and **hard-codes no real-PISA coefficient** as a
package fact. Treat everything here as motivation to read the paper when it
appears, not as a result you may quote from the package.

What the paper supports, stated qualitatively:

- **Point bridge at real PISA scale.** The companion paper checks that the stacked
  fixed-effect point used by `stack_direct` agrees closely with the orthodox
  per-PV Rubin point under the stated production conditions. The reportable
  standard error in `stack_direct` is imposed from the external BRR-Fay /
  Barnard-Rubin target. The package's `per_pv` path remains model-based and
  descriptive, so its standard error should not be described as matching the
  design-based target. This is a topology statement about one stacked fit rather
  than $M$ backend fits, never an efficiency or "faster" claim.
- **Caution 1 — design variance is not optional.** A correct *weighted point* with
  *model-based* standard errors can under-cover severely under complex PISA
  sampling; the design-based BRR–Fay total $T_{\text{MI}}$ restores nominal
  coverage. This is exactly why `stack_direct` is the only coverage-claimable path
  (A4 §5; the proof is **M5**).
- **Caution 2 — small Pareto-$\hat k$ ≠ correct `stack_psis` variance.** Reusing
  one draw cloud makes the per-PV imputations correlated, so `stack_psis` intervals
  can run narrow even when $\hat k$ is small; and $\hat k$ is
  **specification-dependent** — small in the simplified low-dimensional reading demo
  but well above the reliability threshold in the full paper specification, where
  Pipeline B is unreliable. Treat `stack_psis` as a cross-check, never the
  deliverable (A4 §6; **M5**).
- **The "PISA reversal" headline.** The paper's cautionary empirical result is that
  proper PISA **two-stage survey weighting overturns a naive between-school SES-gap
  conclusion**: a between-school socioeconomic gap that looks large and significant
  under naive weighting shrinks and loses significance once the exact two-stage
  weights are applied. This is a *survey-weighting* lesson about the **target** —
  complementary to, and independent of, the stacked-calibration machinery the
  package implements (the target is the subject of **M2**).

Three things this section must **not** do, and does not:

- it does **not** cite any single coefficient as "the result" — the naive *before*
  value is precisely what the reversal overturns, so hard-coding it would
  contradict the paper's own point;
- it makes **no** speed or efficiency claim — "one fit" is a statement about the
  *topology* of the method, not a benchmarked runtime; and
- it makes **no** "verified" simulation claim — the mathematical identities are
  stated as results *under their stated conditions*, not as established simulation
  runs.

For the underlying theory the paper rests on, the **Method track** carries the
derivations this section only gestures at: **M2** (the external BRR–Fay target),
**M4** (CCC calibration), and **M5** (the precise method contrast, PSIS, and why
only `stack_direct` is coverage-claimable). Start the track at **M1 · Foundations
and notation**.

## 8. Where to next

- **A1 · Getting started** and **A2 · The end-to-end workflow** — the same
  `pv_design()` → `pv_brr_target()` → `pv_fit()` pipeline, run end to end on the
  synthetic fixture (and the conceptual live-backend sketch in A2 §7).
- **A3 · Reading results** — the estimate table and the interval-metadata columns
  whose hashes the checklist above records.
- **A4 · Comparing methods** — `pv_compare_methods()` and the two cautions, on
  which Section 7 draws.

For the mathematics — the BRR–Fay target (**M2**), the stacked fractional bridge
(**M3**), CCC (**M4**), and why only `stack_direct` is coverage-claimable
(**M5**) — start the Method track at **M1 · Foundations and notation**.

Bug reports and feature requests:
<https://github.com/joonho112/pvstackr/issues>.

### Session info

```{r session-info, eval = TRUE}
sessionInfo()
```

## References
