---
title: "Plausible values, survey weights and notation"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  This article defines the quantities that pvstackr works with: plausible
  values, the final and replicate survey weights with Fay's coefficient, the
  survey-weighted linear model whose fixed effects pvstackr reports, and the
  weighted likelihood of each plausible value. It lists the symbols and the
  main formulas of the Method articles and reads the design of the bundled
  synthetic data.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Plausible values, survey weights and notation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

## Plausible values and the replicate-weight design {#pv-design}

### Plausible values

Large-scale assessments such as PISA do not observe proficiency directly. Each
student receives $M$ plausible values: draws from the student's posterior
distribution of proficiency given the item responses and background variables
[@mislevy1991randomization; @vondavier2009plausible]. They are multiple
imputations of a latent variable, so an analysis fits its model once for each
plausible value and combines the $M$ results with Rubin's rules
[@rubin1987multiple]. This convention presumes that the analysis model is
congenial to the model that generated the plausible values, as the companion
preprint notes [@lee2026onemcmc, sec. 2]; pvstackr does not check it. The
variation of the $M$ estimates, the
between-imputation variance, reflects the uncertainty that remains about
proficiency; an analysis that uses only `PV1` omits it, so its standard errors
are too small. In pvstackr the plausible values are $M$ columns of the data,
such as `PV1READ` and `PV2READ`, and the model formula names the outcome
`OUTCOME`, which pvstackr replaces by each of these columns in turn.

### Final and replicate weights

PISA selects schools and then students within schools, and the selection
probabilities vary. The final weight $W_i$ of student $i$ combines the inverse
selection probabilities of the school and of the student with adjustment
factors for non-response and trimming [@oecd2024pisa]. The data also carry $R$
replicate weights $W_i^{(r)}$, $r = 1, \dots, R$. An estimate is computed with
the final weight and again with each replicate weight, and the spread of the
$R$ replicate estimates around the full-sample estimate estimates its sampling
variance under the design.

### Fay's method

PISA builds its replicate weights by balanced repeated replication (BRR) with
Fay's method [@oecd2024pisa]. In BRR the sample of each variance stratum is
split into two halves, and a replicate multiplies the weights of one half by 2
and those of the other by 0. Fay's method multiplies them by $2 - k$ and $k$
instead, where $0 \le k < 1$ is the Fay coefficient, so each replicate changes
the weights by $1 - k$ times as much as in BRR. For linear statistics the
squared deviations of the replicate estimates from the full-sample estimate are
then smaller by the factor $(1 - k)^2$, and Fay's variance estimate, the mean
of these squared deviations over the $R$ replicates, is multiplied by
$1/(1 - k)^2$ [@judkins1990fay]. pvstackr applies this estimate to the vector
of regression coefficients, with the factor $1/\{R(1 - k)^2\}$ in front of the
sum of the outer products of the replicate deviations. The target stores this
factor as `fay_variance_multiplier`, and [The design-based target: BRR–Fay
replicate weights and Rubin's rules](m2-brr-fay-target.html#replicate-covariance)
gives the formula.

In PISA 2022 the sampled schools are paired into variance strata (with a
triple, which gets other factors, where a stratum has an odd number of
schools), and each of the 80 replicates multiplies the weights of one school in
each pair by 1.5 and those of the other by 0.5, so $k = 0.5$. The non-response
adjustments and the trimming are then repeated for each replicate, so the
replicate weights differ slightly from these factors, and they are columns of
the data file. PISA 2022 also provides ten plausible values per domain, and the factor
$1/\{R(1 - k)^2\} = 1/(80 \times 0.25) = 0.05$ is the one in its variance
formula [@oecd2024pisa]. The bundled synthetic example data are much smaller:

| Data | $M$ | $R$ | $k$ | $1/\{R(1-k)^2\}$ |
|:--|:--:|:--:|:--:|:--:|
| PISA 2022 | 10 | 80 | 0.5 | 0.05 |
| Bundled example data, `pisa_tiny` | 2 | 4 | 0.5 | 1 |

In the example the factor is exactly 1 because $R(1 - k)^2 = 4 \times 0.25 = 1$,
a property of these two numbers only.

`pv_brr_target()` takes the final weight and at least two replicate weights as
columns of the data (`weight_col`, `rep_weight_cols`) and the Fay coefficient
as `fay_k` (default 0.5). It does not check `fay_k` against the replicate
weights, and it requires every weight to be positive, so plain BRR weights,
which are zero for half of the sample, are not accepted.

## The model pvstackr fits {#model}

pvstackr works with one model: a linear regression of the plausible value on
the covariates, weighted by the survey weights. For plausible value $m$ and
student $i = 1, \dots, N$,

$$
y_i^{(m)} = x_i^\top \beta + \varepsilon_i^{(m)}, \qquad
\varepsilon_i^{(m)} \sim N(0, \sigma^2),
$$

where $x_i$ is the row of the $N \times p$ design matrix $X$ for student $i$
(the intercept and the covariates of the formula), $\beta$ holds the $p$
fixed-effect coefficients, and $\sigma$ is the residual standard deviation. The
covariates, and with them $X$, are the same for every plausible value; only the
outcome changes. The formula of the examples, `OUTCOME ~ x + female`, has
$p = 3$ coefficients, which pvstackr names `b_Intercept`, `b_x` and `b_female`.

pvstackr uses this model in two places, both with the survey weights.
`pv_brr_target()` computes, for each plausible value, the weighted
least-squares estimate of $\beta$ with the final weights (`lm.wfit()` from the
stats package) and its covariance from the replicate weights; neither step uses
the normal distribution of the errors. The stacked fit of `stack_direct` fits
the model, normal errors included, once to all $M$ plausible values together,
with the survey weights in its likelihood ([below](#estimand)). It uses a
Bayesian engine, either the brms engine bundled with pvstackr or fitting
functions that you supply, and it accepts only the Gaussian family with the
identity link; any other family stops it with an error. Only the fixed effects
are reported: the stacked fit also estimates $\sigma$, which pvstackr neither
calibrates nor reports.

The model has no random effects. A formula with a random-effect term such as
`(1 | CNTSCHID)` stops `pv_brr_target()`, `stack_direct` and `stack_psis` with
an error, so pvstackr 0.2.x fits no school-level variance. `per_pv` fits no
model itself: it pools, with Rubin's rules, draws for each plausible value that
you supply or that your fitting function returns. It does not check the
formula for random-effect terms and does not add the survey weights to it, so
for `per_pv` the model is the one that your function fits.

### School means as covariates

pvstackr computes no school or cluster means. A school mean that you compute
and store as a column, for example the weighted mean $\bar x_j$ of a covariate
$x$ over the sampled students of school $j$, is an ordinary covariate. When
pvstackr stacks the data, it copies every covariate unchanged into each of the
$M$ copies, so such a column has the same value for every plausible value. The
companion preprint [@lee2026onemcmc] needs this invariance. Its stacked
fixed-effect point identity (Theorem 4.1 of the preprint) requires, as
condition (R2), a design matrix common to all plausible values, which holds
because the covariates, the analyzed rows and the cluster membership do not
change with the plausible value (Appendix A.2 of the preprint). [One stacked
fit and the Rubin point estimate](m3-stacked-bridge.html#conditions) lists the
conditions.

With a school mean among the covariates, the regression can separate the
association within schools from the association between schools. For student
$i$ in school $j$, the within–between parameterisation enters the deviation
$x_{ij} - \bar x_j$ with slope $\beta^W$ and the school mean $\bar x_j$ with
slope $\beta^B$. It is a reparameterization of the model that enters $x_{ij}$
and $\bar x_j$ together, the form of @mundlak1978pooling, who added the group
means of the covariates to an error-components model:

$$
\beta^W (x_{ij} - \bar x_j) + \beta^B \bar x_j
  = \beta^W x_{ij} + (\beta^B - \beta^W)\, \bar x_j .
$$

In pvstackr both forms are covariates of the single-level model above; neither
adds a school random effect.

### The model of the companion preprint

The preprint develops the stacked fit and its calibration for a two-level
model with a school random intercept (its Eq. (2)). In the notation of this
article,

$$
y_{ij}^{(m)} = \beta_0 + \beta^W (x_{ij} - \bar x_j) + \beta^B \bar x_j
  + \zeta_j + \varepsilon_{ij}^{(m)}, \qquad
\zeta_j \sim N(0, \tau^2), \quad \varepsilon_{ij}^{(m)} \sim N(0, \sigma^2),
$$

where $\bar x_j$ is the school mean of the covariate, which the preprint's
application estimates by the weighted mean of the sampled students; the
preprint writes
$\theta_{ij}^{(m)}$ for the plausible value. Its weighted likelihood multiplies
each student's log-likelihood contribution by the student's normalized weight
inside the integral over the random intercept $\zeta_j$ (its Eq. (3)), and the
variance components $\tau^2$ and $\sigma^2$ are nuisance parameters. The
preprint's default target is computed from model-based fits of each plausible
value, by restricted maximum likelihood, with a replicate-weight target as an
alternative (its Appendices C.2 and C.3). pvstackr 0.2.x does not fit
this model, and it computes only the replicate-weight target.

## The estimand and the survey-weighted likelihood {#estimand}

The estimand is $\beta$, the vector of the $p$ fixed-effect coefficients (the
intercept and the slopes) of the model above; $\sigma$ is a nuisance
parameter. For each coefficient $\beta_\ell$, $\ell = 1, \dots, p$, pvstackr
reports an estimate, a standard error, degrees of freedom and an interval.

The survey weights enter through the likelihood. Let $\overline W$ be the mean
of the final weights and $\tilde w_i = W_i / \overline W$ the normalized weight
of student $i$, which has mean 1. The survey-weighted likelihood for plausible
value $m$ is

$$
\tilde L_m(\beta, \sigma) = \prod_{i=1}^{N}
  \phi\big(y_i^{(m)};\, x_i^\top \beta,\, \sigma^2\big)^{\tilde w_i},
$$

where $\phi(\cdot\,; \mu, \sigma^2)$ is the normal density with mean $\mu$ and
variance $\sigma^2$: each student's log-likelihood contribution is multiplied
by $\tilde w_i$. This is the preprint's weighted likelihood (its Eq. (3), whose
weights are also normalized to mean 1) for a model without the random
intercept. For any fixed $\sigma$, $\tilde L_m$ is largest, as a function of
$\beta$, at the weighted least-squares estimate $\hat\beta_m$, the estimate
that `pv_brr_target()` computes for plausible value $m$ with the final weights;
the scale of the weights does not change it. `pv_brr_target()` also estimates
the covariance $\hat U_m$ of $\hat\beta_m$ from the replicate weights and
combines the $M$ results with Rubin's rules into $\bar\beta$, the average of
the $\hat\beta_m$, and its covariance $T_{\text{MI}}$ ([The design-based
target: BRR–Fay replicate weights and Rubin's
rules](m2-brr-fay-target.html#rubin)).

The stacked fit of `stack_direct` uses the stacked data: one copy of the data
for each plausible value, $N \cdot M$ rows in all, where row $i$ of copy $m$ has
the outcome $y_i^{(m)}$ and the weight $\tilde w_i / M$. The weights of each
copy sum to $N/M$, and those of all rows to $N$. The bundled brms engine
multiplies the log-likelihood of each row by its weight, so the log likelihood
of the stacked fit is the average of the $M$ survey-weighted log-likelihoods,
the stacked objective, and its posterior is the stacked fractional posterior of
the preprint (its Eq. (4)),

$$
q(\beta, \sigma) \propto p(\beta, \sigma) \prod_{m=1}^{M}
  \tilde L_m(\beta, \sigma)^{1/M},
$$

where $p(\beta, \sigma)$ is the prior. A fitting function of your own receives
the weights in the formula term `weights(.pvstackr_weight)` and the data column
`.pvstackr_weight`, and has to use them in the same way ([The full analysis
workflow](a2-the-workflow.html#own-engine)).
The stacked fixed-effect point identity
gives conditions under which the maximizer of the stacked objective in $\beta$
equals $\bar\beta$; [One stacked fit and the Rubin point
estimate](m3-stacked-bridge.html#point-identity) states them and applies them
to pvstackr's model. The identity concerns the point estimate only. The
variance comes from the target: the calibration gives the fixed-effect draws
of the stacked fit the target's mean $\bar\beta$ and covariance $T_{\text{MI}}$
([Calibrating the fixed-effect draws (CCC)](m4-ccc-calibration.html#map)).

## Symbols and formulas {#notation}

The Method articles use the symbols below, and each article defines them again
where it first uses them. The last column gives the argument, field or column
of pvstackr that holds the quantity.

### Symbols

The indices:

| Symbol | Meaning | Range |
|:--|:--|:--|
| $i$ | student (a row of the data) | $1, \dots, N$ |
| $m$ | plausible value | $1, \dots, M$ |
| $r$ | replicate weight | $1, \dots, R$ |
| $\ell$ | fixed-effect coefficient | $1, \dots, p$ |
| $s$ | posterior draw | $1, \dots, S$ |
| $j$ | school, where a school mean is discussed | — |

The letter $k$ is not an index: it denotes the Fay coefficient, and the
Pareto shape estimate of importance sampling is always written Pareto
$\hat k$.

The data, the weights and the model:

| Symbol | Meaning | In pvstackr |
|:--|:--|:--|
| $y_i^{(m)}$ | outcome of student $i$ under plausible value $m$ | plausible-value columns such as `PV1READ` (`OUTCOME` in the formula) |
| $x_i$, $X$ | covariate row of student $i$; the $N \times p$ design matrix, the same for every $m$ | model matrix of the formula |
| $\beta$, $\beta_\ell$ | fixed-effect coefficients; coefficient $\ell$ | rows of the estimate table (`term`: `b_Intercept`, `b_x`, …) |
| $\sigma$ | residual standard deviation, not reported | — |
| $W_i$ | final survey weight of student $i$ | `weight_col` |
| $\overline W$ | mean of the final weights | — |
| $\tilde w_i = W_i / \overline W$ | normalized final weight (mean 1) | — |
| $\tilde w_i / M$ | weight of row $i$ in each of the $M$ stacked copies | column `.pvstackr_weight` of the stacked data |
| $W_i^{(r)}$ | replicate weight $r$ of student $i$ | `rep_weight_cols` |
| $k$ | Fay coefficient | `fay_k` (PISA: 0.5) |
| $\tilde L_m(\beta, \sigma)$ | survey-weighted likelihood for plausible value $m$ | — |

The target, the stacked fit and the calibration:

| Symbol | Meaning | In pvstackr |
|:--|:--|:--|
| $\hat\beta_m$ | weighted least-squares estimate for plausible value $m$ with the final weights | `per_pv[[m]]$beta` in the target |
| $\hat\beta_m^{(r)}$ | the same with replicate weight $r$ | columns of `per_pv[[m]]$replicate_beta` |
| $\hat U_m$ | BRR–Fay replicate covariance of $\hat\beta_m$ | `per_pv[[m]]$U` |
| $\bar\beta$ | combined estimate, the average of the $\hat\beta_m$ | `beta`, `beta_bar` |
| $\bar U$ | average covariance within plausible values | `U_bar` |
| $B$ | covariance between plausible values | `B` |
| $T_{\text{MI}}$ | total covariance of $\bar\beta$ | `T_MI` |
| $\lambda_\ell$ | fraction of missing information of coefficient $\ell$ (Barnard and Rubin's approximation) | `fmi`, `lambda` (same values) |
| $\mathrm{riv}_\ell$ | relative increase in variance | `riv` |
| $\nu_\ell$ | classic degrees of freedom | `df` with `df_method = "classic"` |
| $\nu_{\text{com}}$ | complete-data degrees of freedom, stated by the user | `df_complete` |
| $\nu_{\text{obs},\ell}$, $\nu_{\text{BR},\ell}$ | Barnard–Rubin observed-data and adjusted degrees of freedom | `df` with `df_method = "barnard_rubin"` |
| $q(\beta, \sigma)$ | stacked fractional posterior | — |
| $\beta_s$ | fixed-effect draw $s$ of the stacked fit, before calibration | — |
| $\bar\beta^{\text{raw}}$, $\Sigma_{\text{raw}}$ | mean and covariance of these draws | — |
| $L_{\text{raw}}$, $L_{\text{tgt}}$ | lower Cholesky factors of $\Sigma_{\text{raw}}$ and $T_{\text{MI}}$ | — |
| $A$ | calibration matrix | — |
| $\beta^{\text{cal}}_s$ | calibrated draw | rows of `get_draws(fit)` |
| $\kappa_A$ | condition number of $A$, the ratio of its largest to its smallest singular value | `kappa_A` in `get_diagnostics(fit)$ccc` |
| $\Delta_c$ | center separation | `delta_c_max` in `get_diagnostics(fit)$ccc` |
| $\omega_s^{(m)}$ | importance weight of draw $s$ for plausible value $m$, supplied by the user; pvstackr divides each column by its sum | `psis_weights` (one row per draw, one column per plausible value) |
| $\hat k_m$ | Pareto $\hat k$ for plausible value $m$, supplied by the user | `pareto_k` |

[Choosing a method and reading its
intervals](m5-methods-and-coverage.html#three-methods) also uses $\hat\beta_m$
and $\hat U_m$ for the per-PV estimate and covariance of `per_pv` and
`stack_psis`.

### Main formulas

The target of `pv_brr_target()` is derived in [The design-based target: BRR–Fay
replicate weights and Rubin's rules](m2-brr-fay-target.html): the BRR–Fay
replicate covariance of each $\hat\beta_m$,

$$
\hat U_m = \frac{1}{R(1-k)^2} \sum_{r=1}^{R}
  \big(\hat\beta_m^{(r)} - \hat\beta_m\big)\big(\hat\beta_m^{(r)} - \hat\beta_m\big)^\top ,
$$

Rubin's rules,

$$
\bar\beta = \frac{1}{M} \sum_{m=1}^{M} \hat\beta_m, \qquad
\bar U = \frac{1}{M} \sum_{m=1}^{M} \hat U_m,
$$

$$
B = \frac{1}{M-1} \sum_{m=1}^{M} \big(\hat\beta_m - \bar\beta\big)\big(\hat\beta_m - \bar\beta\big)^\top, \qquad
T_{\text{MI}} = \bar U + \Big(1 + \frac{1}{M}\Big) B ,
$$

and, for each coefficient $\ell$, the fraction of missing information, the
relative increase in variance, and the classic and Barnard–Rubin degrees of
freedom:

$$
\lambda_\ell = \frac{(1 + 1/M)\, B_{\ell\ell}}{T_{\text{MI},\ell\ell}}, \qquad
\mathrm{riv}_\ell = \frac{(1 + 1/M)\, B_{\ell\ell}}{\bar U_{\ell\ell}}, \qquad
\nu_\ell = \frac{M - 1}{\lambda_\ell^2},
$$

$$
\nu_{\text{obs},\ell} = \frac{\nu_{\text{com}} + 1}{\nu_{\text{com}} + 3}\,
  \nu_{\text{com}} (1 - \lambda_\ell), \qquad
\nu_{\text{BR},\ell} = \Big(\frac{1}{\nu_\ell} + \frac{1}{\nu_{\text{obs},\ell}}\Big)^{-1} .
$$

The stacked fractional posterior $q(\beta, \sigma)$ is given
[above](#estimand); [One stacked fit and the Rubin point
estimate](m3-stacked-bridge.html#stacked-objective) derives the stacked
objective and the point identity.

The calibration map and the center separation are derived in [Calibrating the
fixed-effect draws (CCC)](m4-ccc-calibration.html). With `center = "target"`,
which `stack_direct` requires, each fixed-effect draw of the stacked fit is
mapped to

$$
\beta^{\text{cal}}_s = \bar\beta + A\big(\beta_s - \bar\beta^{\text{raw}}\big),
  \qquad A = L_{\text{tgt}} L_{\text{raw}}^{-1},
$$

where $L_{\text{raw}} L_{\text{raw}}^\top = \Sigma_{\text{raw}}$ and
$L_{\text{tgt}} L_{\text{tgt}}^\top = T_{\text{MI}}$, so that the calibrated
draws have mean $\bar\beta$ and covariance $T_{\text{MI}}$. Before the
calibration, pvstackr measures how far the stacked fit is from the target by

$$
\Delta_c = \max_\ell
  \frac{\lvert \bar\beta_\ell - \bar\beta^{\text{raw}}_\ell \rvert}
       {\sqrt{T_{\text{MI},\ell\ell}}} ,
$$

stored as `delta_c_max`; `delta_c_rel` is the root mean square of the same
ratios.

The pooling of `per_pv` and `stack_psis`, with the importance weights
$\omega_s^{(m)}$ and the Pareto $\hat k_m$ of `stack_psis`, is given in
[Choosing a method and reading its intervals](m5-methods-and-coverage.html#three-methods).

## The symbols on the example data

pvstackr includes a small synthetic data set, `pisa_tiny`, and a `stack_direct`
fit made from it; [Getting started with
pvstackr](a1-getting-started.html#example-fit) describes both. The chunks below
only read the data and the target of the fit; they estimate nothing.

```{r example-data-shape}
pisa_tiny <- read.csv(
  system.file("extdata", "pisa_tiny.csv", package = "pvstackr")
)

dim(pisa_tiny)        # 12 students, 12 columns
names(pisa_tiny)
```

The rows are `r nrow(pisa_tiny)` students in three schools of one made-up
country (`CNT` is `"SYN"`), so $N = 12$; `CNTSCHID` and `CNTSTUID` identify the
schools and the students. The covariates are `x` and the 0/1 indicator
`female`. `PV1READ` and `PV2READ` are the $M = 2$ plausible values $y_i^{(1)}$
and $y_i^{(2)}$, `W_FSTUWT` is the final weight $W_i$, and `W_FSTURWT1` to
`W_FSTURWT4` are the $R = 4$ replicate weights $W_i^{(r)}$. The replicate
weights were typed in for the example rather than formed from half-samples by
Fay's method; `fay_k = 0.5` is the value declared for them.

`detect_pisa_pv_columns()` and `detect_pisa_brr_replicate_weights()` find the
plausible-value and replicate-weight columns by their names:

```{r detect-design}
detect_pisa_pv_columns(pisa_tiny, suffix = "READ")   # the M = 2 plausible values
detect_pisa_brr_replicate_weights(pisa_tiny)         # the R = 4 replicate weights
```

The plausible-value names carry the subject suffix `READ`, as in PISA 2022
files, so the call gives `suffix = "READ"`. With the default suffix `""`,
`detect_pisa_pv_columns()` looks for bare names such as `PV1` and stops with an
error on these data.

The target of the example fit records the same design and the names of the
coefficients:

```{r read-target}
fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit
tg <- get_target(fit)

c(M = tg$M, R = tg$R, fay_k = tg$fay_k)   # M, R and the Fay coefficient k
tg$fe_names                               # the p = 3 fixed-effect coefficients
```

The three coefficients `b_Intercept`, `b_x` and `b_female` of the formula
`OUTCOME ~ x + female` make up $\beta$ in this example, and the estimate table
of the fit has one row for each.

The next Method article, [The design-based target: BRR–Fay replicate weights
and Rubin's rules](m2-brr-fay-target.html), derives the target from these
quantities.

## References
