---
title: "The design-based target: BRR–Fay replicate weights and Rubin's rules"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  The function `pv_brr_target()` computes the design-based target that a
  `stack_direct` fit reports: for each plausible value, weighted
  least-squares estimates of the fixed effects and their BRR–Fay replicate
  covariance, combined over the plausible values with Rubin's rules. This
  article gives these formulas, the classic and Barnard–Rubin degrees of
  freedom and the fraction of missing information, with the source of each.
  It then recomputes the combined quantities of the example target from the
  estimates and covariances stored in it for each plausible value.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{The design-based target: BRR–Fay replicate weights and Rubin's rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

## The target and the fit that reports it

`pv_brr_target()` estimates the fixed effects of a survey-weighted linear
regression from plausible-value data, together with their covariance matrix.
It fits no Bayesian model. For each plausible value it fits the regression by
weighted least squares, once with the final weight and once with each
replicate weight, and estimates the sampling covariance of the coefficients
from the spread of the replicate estimates. Rubin's rules then combine the
results for the $M$ plausible values. The replicate weights are assumed to
come from balanced repeated replication (BRR) with Fay's method, as in PISA.
Because the sampling covariance comes from these replicate weights of the
survey design, not from a model, the target is called design-based.

The result, an object of class `pvstackr_brr_target`, holds the combined
estimates $\bar\beta$, their covariance matrix $T_{\text{MI}}$, the standard
errors, the degrees of freedom and the fraction of missing information;
`?pv_brr_target` lists all its fields. The regression has fixed effects only:
a random-effect term such as `(1 | school)` in the formula stops
`pv_brr_target()` with an error ([Plausible values, survey weights and
notation](m1-foundations-and-notation.html#model) describes the model and the
symbols).

A `stack_direct` fit reports this target: its estimate table copies the
target's numbers ([The full analysis
workflow](a2-the-workflow.html#center-target)), and `get_target(fit)` returns
the target.

pvstackr takes the covariance from the target because a single stacked fit
cannot supply it. The draws of the stacked fit do not contain the variation
between plausible values, $B$, which exists only across separate analyses of
the plausible values, as the companion preprint notes
[@lee2026onemcmc, sec. 4.4]. Their variance within a plausible value comes
from the model, not from the replicate weights. The Cholesky calibration
correction (CCC) gives the stacked draws the mean $\bar\beta$ and the
covariance $T_{\text{MI}}$ of the target ([Calibrating the fixed-effect draws
(CCC)](m4-ccc-calibration.html)).

pvstackr's reporting rule labels the intervals of a `stack_direct` fit
coverage-claimable, that is, readable as confidence intervals with nominal
coverage, only when its target uses the Barnard–Rubin degrees of freedom
described below ([Reading and reporting the
results](a3-reading-results.html#interval-columns) states the rule in full).

The code below reads the target of the example fit that ships with pvstackr
and recomputes parts of it with base R; it fits no model. The target was
computed by `pv_brr_target()` from the bundled synthetic data, which [Getting
started with pvstackr](a1-getting-started.html#example-fit) describes together
with the example fit:

```{r common-load}
fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit
tg <- get_target(fit)   # the target of the fit, a pvstackr_brr_target object

class(tg)
c(M = tg$M, R = tg$R, fay_k = tg$fay_k)
```

The synthetic data have $M = 2$ plausible values, $R = 4$ replicate weights
and the Fay coefficient $k = 0.5$.

## The BRR–Fay replicate covariance {#replicate-covariance}

For plausible value $m$, let $\hat\beta_m$ be the weighted least-squares
estimate of the $p$ fixed-effect coefficients with the final weights $W_i$,
and $\hat\beta_m^{(r)}$ the estimate with replicate weight $W_i^{(r)}$, for
$r = 1, \dots, R$. pvstackr computes them with `lm.wfit()` from the stats
package, $M(R + 1)$ fits in all. The BRR–Fay replicate covariance of
$\hat\beta_m$ is

$$
\hat U_m = \frac{1}{R(1-k)^2} \sum_{r=1}^{R}
  \big(\hat\beta_m^{(r)} - \hat\beta_m\big)\big(\hat\beta_m^{(r)} - \hat\beta_m\big)^\top ,
$$

where $k$ is the Fay coefficient `fay_k`. The sum adds the outer products of
the deviations of the replicate estimates from the estimate with the final
weights. $\hat U_m$ is a replication estimate of variance: it needs only the
replicate estimates, not a linearization of the estimator. In the terms of
multiple imputation, $\hat\beta_m$ and $\hat U_m$ are the complete-data
estimate and its variance for the data set completed with plausible value
$m$.

Fay's method changes the weights less than plain BRR: in each replicate it
multiplies the weights of one half-sample by $2 - k$ and those of the other
by $k$, where plain BRR uses 2 and 0 ([Plausible values, survey weights and
notation](m1-foundations-and-notation.html#pv-design) describes the replicate
weights). For a linear statistic the replicate deviations are then $1 - k$
times those of plain BRR, so their mean square is too small by the factor
$(1-k)^2$, and multiplying by $1/(1-k)^2$ corrects it [@judkins1990fay,
pp. 224–225]. The factor $1/R$ averages over the replicates. pvstackr requires
positive replicate weights, so it does not accept plain BRR weights, which are
zero for half of the sample.

The target stores the factor $1/\{R(1-k)^2\}$ as `fay_variance_multiplier`.
For PISA 2022, with $R = 80$ and $k = 0.5$, it is $1/(80 \cdot 0.25) = 0.05$,
the constant in the variance formula of the PISA 2022 Technical Report
[@oecd2024pisa, ch. 10]. In the example, $R = 4$ and $k = 0.5$ give
$1/(4 \cdot 0.25) = 1$, a value that follows from the small example design.

For each plausible value, `per_pv[[m]]` holds `beta` ($\hat\beta_m$), `U`
($\hat U_m$), `replicate_beta` (the $\hat\beta_m^{(r)}$, one column per
replicate weight) and `replicate_diff` (the deviations
$\hat\beta_m^{(r)} - \hat\beta_m$). For the first plausible value:

```{r brrfay-peek}
tg$fay_variance_multiplier                 # 1/(R (1 - k)^2), 1 in the example
dim(tg$per_pv[[1]]$replicate_diff)         # 3 coefficients x 4 replicate weights
str(tg$per_pv[[1]]$replicate_diff)
```

Each column holds the deviations of the three coefficients for one replicate
weight. $\hat U_1$, stored as `per_pv[[1]]$U`, is `fay_variance_multiplier`
times the sum of the outer products of these columns.

## Rubin's rules {#rubin}

Rubin's rules combine the $M$ estimates and covariances
[@rubin1987multiple, pp. 76–77]:

$$
\bar\beta = \frac{1}{M}\sum_{m=1}^{M} \hat\beta_m, \qquad
\bar U = \frac{1}{M}\sum_{m=1}^{M} \hat U_m, \qquad
B = \frac{1}{M-1}\sum_{m=1}^{M}
  (\hat\beta_m - \bar\beta)(\hat\beta_m - \bar\beta)^\top,
$$

$$
T_{\text{MI}} = \bar U + \Big(1 + \frac{1}{M}\Big) B .
$$

- $\bar\beta$ (field `beta_bar`, also stored as `beta`) is the average of the
  $M$ estimates, the target's estimate of the coefficients.
- $\bar U$ (`U_bar`) is the average of the replicate covariances: the
  complete-data covariance, that is, the sampling covariance an estimate would
  have if the plausible value were the observed outcome.
- $B$ (`B`) is the covariance matrix of the $M$ estimates, with divisor
  $M - 1$: the variation between plausible values, which comes from the
  uncertainty about each student's proficiency.
- $T_{\text{MI}}$ (`T_MI`, also stored as `total_var`) is the total
  covariance matrix of $\bar\beta$. The standard errors (`se`) are the square
  roots of its diagonal.

The factor $1 + 1/M$ adds $B/M$ to $\bar U + B$. It accounts for the
simulation variance of $\bar\beta$, which averages a finite number of
plausible values: $B/M$ estimates the variance of $\bar\beta$ around the
average over infinitely many plausible values [@rubin1987multiple,
pp. 89–90; @barnard1999small, p. 950].

For a single statistic this is the combination that the PISA 2022 Technical
Report gives for estimates from plausible values, with the sampling variance
of each plausible value from the replicate weights [@oecd2024pisa, ch. 11].
pvstackr applies it to the whole coefficient vector and keeps the full
matrices: CCC uses all of $T_{\text{MI}}$, not only its diagonal, as the
covariance of the calibrated draws ([Calibrating the fixed-effect draws
(CCC)](m4-ccc-calibration.html#map)).

## Degrees of freedom {#degrees-of-freedom}

A `stack_direct` fit forms each interval as the estimate plus or minus a $t$
quantile times the standard error. The quantile uses the degrees of freedom of
the target and the level `conf_level` of `pv_control()` (0.95 by default); the
`conf_level` stored in the target does not set it. Both rules for the degrees
of freedom work coefficient by coefficient with the ratio

$$
\lambda_\ell = \frac{(1 + 1/M)\,B_{\ell\ell}}{T_{\text{MI},\ell\ell}} ,
$$

the share of the total variance of coefficient $\ell$ that comes from the
variation between plausible values ([the fraction of missing
information](#fmi) returns to it). The argument `df_method` of
`pv_brr_target()` chooses the rule.

`df_method = "classic"`, the default, is the rule of
@rubin1987multiple [p. 77]:

$$
\nu_\ell=(M-1)/\lambda_\ell^{2} .
$$

Rubin writes it as $(M - 1)(1 + 1/\mathrm{riv}_\ell)^2$, with the relative increase in
variance $\mathrm{riv}_\ell = (1 + 1/M)B_{\ell\ell}/\bar U_{\ell\ell}$; the two forms are
equal because $\lambda_\ell = \mathrm{riv}_\ell/(1 + \mathrm{riv}_\ell)$. The rule is derived under
the assumption that the complete-data degrees of freedom, those the analysis
would have if no information were missing, are infinite
[@barnard1999small, p. 948]. Because $\lambda_\ell \le 1$, $\nu_\ell$ is never
smaller than $M - 1$.

`df_method = "barnard_rubin"` is the rule of
@barnard1999small [pp. 948–949]. It needs the complete-data degrees of
freedom $\nu_{\text{com}}$, which you give as `df_complete`, and combines the
classic value with an estimate of the observed-data degrees of freedom:

$$
\nu_{\text{obs},\ell}=\frac{\nu_{\text{com}}+1}{\nu_{\text{com}}+3}\,
  \nu_{\text{com}}\,(1-\lambda_\ell),
\qquad
\nu_{\text{BR},\ell}=\left(\frac{1}{\nu_\ell}+\frac{1}{\nu_{\text{obs},\ell}}\right)^{-1} .
$$

Barnard and Rubin derived this adjustment because the classic value can be
many times larger than $\nu_{\text{com}}$ when $\nu_{\text{com}}$ is small and
little information is missing. It corrects for small complete-data degrees of
freedom, not for a small number of plausible values. $\nu_{\text{BR},\ell}$ is
never larger than $\nu_\ell$ or $\nu_{\text{com}}$, increases with
$\nu_{\text{com}}$, and equals $\nu_\ell$ when $\nu_{\text{com}}$ is infinite.
Unlike $\nu_\ell$, it can be smaller than $M - 1$: when $\lambda_\ell$ is close
to 1, $\nu_{\text{obs},\ell}$ and therefore $\nu_{\text{BR},\ell}$ are small.

`df_complete` is the complete-data degrees of freedom of the replicate-weight
variance: the degrees of freedom the analysis would have if the outcome were
observed directly instead of through plausible values. You state it, as one
positive number or as one number per coefficient; pvstackr does not choose it
and does not check it against the replicate weights. `df_complete = Inf` is
accepted and gives the classic degrees of freedom. The adjustment matters when
the number is small.

`df` holds the degrees of freedom of the chosen rule, and `df_classic` holds
the classic values under both rules. With `"classic"`, the field
`df_complete` of the target is `NULL`, and the estimate table of a fit shows
`NA` in its column `df_complete`. `df_method` also sets the interval labels
that a `stack_direct` fit copies: `"classic"` gives
`interval_role = "descriptive_classic_rubin"` and
`coverage_claim_allowed = FALSE`, and `"barnard_rubin"` gives
`"coverage_barnard_rubin"` and `TRUE` for any positive `df_complete`, `Inf`
included. These labels follow pvstackr's reporting rule (see [where the
formulas and the interval labels come from](#sources)).

The example target uses the classic rule and has no `df_complete`, so this
article prints the formula but does not evaluate it.

## Fraction of missing information {#fmi}

pvstackr stores $\lambda_\ell$ in the target as `fmi` and again as `lambda`;
the two fields are identical, and the estimate table has no column for them.
@barnard1999small [p. 948] use this ratio as an approximation to the fraction
of missing information (FMI) of coefficient $\ell$.
@rubin1987multiple [p. 77] estimates the fraction of missing information with
an additional term that depends on the degrees of freedom:

$$
\gamma_\ell = \frac{\mathrm{riv}_\ell + 2/(\nu_\ell + 3)}{\mathrm{riv}_\ell + 1} ,
$$

where $\mathrm{riv}_\ell$ is the relative increase in variance, stored as `riv`, and
$\nu_\ell$ are the classic degrees of freedom. Since
$\lambda_\ell = \mathrm{riv}_\ell/(1 + \mathrm{riv}_\ell)$, $\gamma_\ell$ exceeds $\lambda_\ell$ by
$2/\{(\nu_\ell + 3)(\mathrm{riv}_\ell + 1)\}$, and the two are close when $\nu_\ell$ is
large. pvstackr computes only $\lambda_\ell$, so an FMI reported from pvstackr
is $\lambda_\ell$.

$\lambda_\ell$ is close to 1 when the variation between plausible values is
much larger than the replicate-weight variance, that is, when $\mathrm{riv}_\ell$ is
large, whatever the number of plausible values. The classic degrees of
freedom are then close to their smallest value, $M - 1$. This is the case in
the example, where $M - 1 = 1$ and the relative increase in variance is large
([the checks on the example target](#checks) show the values). The FMI of an
analysis of real data depends on its data and model; the synthetic example
says nothing about it.

## Where the formulas and the interval labels come from {#sources}

The replicate covariance is the variance estimator of Fay's method for
balanced repeated replication [@judkins1990fay], and the combination is
Rubin's rules [@rubin1987multiple] with the degrees of freedom of
@rubin1987multiple or @barnard1999small. The PISA 2022 Technical Report
computes standard errors from plausible values and replicate weights in the
same way, one statistic at a time [@oecd2024pisa, chs. 10–11]. pvstackr
applies these formulas to the coefficients of its survey-weighted linear
regression.

The companion preprint [@lee2026onemcmc] builds its default target from
model-based fits of each plausible value and describes a target from replicate
weights as an alternative (Appendix C of the preprint). Its simulation study
compares estimators whose variances and targets come from model-based fits,
on simulated data without replicate weights [@lee2026onemcmc, sec. 5]. It
therefore provides no evidence about a BRR–Fay target, and it does not compare
model-based with design-based standard errors. [Real PISA data: access,
memory and reproducibility](a5-real-pisa-guidance.html#companion-preprint)
summarizes what the preprint shows.

The interval labels `interval_role` and `coverage_claim_allowed` are
pvstackr's reporting rule, not a result of any of these sources. Under this
rule only a `stack_direct` fit whose target was made with
`df_method = "barnard_rubin"` and a positive `df_complete` has intervals
labeled coverage-claimable. [Choosing a method and reading its
intervals](m5-methods-and-coverage.html#interval-rule) gives the reason for
the rule, and [Reading and reporting the
results](a3-reading-results.html#interval-columns) explains how to read the
labels.

## Checks on the example target {#checks}

The chunks below recompute the combined quantities of the example target from
the estimates and covariances of its two plausible values, stored in
`tg$per_pv`. These are arithmetic checks: they show that the stored
`beta_bar`, `U_bar`, `B`, `T_MI` and `fmi` are Rubin's rules applied to the
stored pieces for each plausible value. They take those pieces as given and
do not repeat the weighted least-squares fits.

```{r rubin-pooling}
M     <- tg$M                                    # 2
betas <- sapply(tg$per_pv, function(p) p$beta)   # 3 x M: one column per plausible value
Us    <- lapply(tg$per_pv, function(p) p$U)      # the replicate covariances U_m

beta_bar <- rowMeans(betas)                      # average of the estimates
U_bar    <- Reduce(`+`, Us) / M                  # average replicate covariance
dev      <- betas - beta_bar
B        <- (dev %*% t(dev)) / (M - 1)           # covariance between plausible values
T_MI     <- U_bar + (1 + 1/M) * B                # total covariance

all.equal(unname(beta_bar), unname(tg$beta_bar))
all.equal(unname(U_bar),    unname(tg$U_bar))
all.equal(unname(B),        unname(tg$B))
all.equal(unname(T_MI),     unname(tg$T_MI))
```

The four comparisons return `TRUE`: the recomputed and the stored values agree
within the default tolerance of `all.equal()`, a relative difference of about
$1.5 \times 10^{-8}$. The ratio $\lambda_\ell$ follows from $B$ and
$T_{\text{MI}}$ in the same way:

```{r fmi-recompute}
fmi <- (1 + 1/M) * diag(B) / diag(T_MI)          # lambda for each coefficient
all.equal(unname(fmi), unname(tg$fmi))
```

The target also stores the relative increase in variance and the degrees of
freedom, with the rule that gave them:

```{r df-table}
data.frame(
  term      = tg$fe_names,
  fmi       = round(unname(tg$fmi), 3),
  riv       = round(unname(tg$riv), 2),
  df        = round(unname(tg$df),  2),
  df_method = unname(tg$df_method)
)
```

`df_method` is `"classic"` for every coefficient, so `df` holds
$(M-1)/\lambda_\ell^2$: with the rounded values of `fmi`,
$1/0.990^2 \approx 1.02$, $1/0.844^2 \approx 1.40$ and
$1/0.993^2 \approx 1.01$. The FMI is large because `riv` is large (about 95,
5 and 146), so the degrees of freedom are near their smallest possible value,
$M - 1 = 1$, most closely for the intercept and `b_female`.

The example target uses classic degrees of freedom, so it has no
complete-data degrees of freedom:

```{r df-complete}
tg$df_complete   # NULL for a classic target
```

The estimate table of the fit shows `NA` in its column `df_complete`, and by
pvstackr's reporting rule the intervals of the fit are labeled
`descriptive_classic_rubin`, with `coverage_claim_allowed = FALSE`.

For one weighted fit to the stacked data, in which every row has the weight
$\tilde w_i/M$ (with $\tilde w_i = W_i/\overline W$, the final weight divided
by its mean), the stacked fixed-effect point identity,
Theorem 4.1 of the preprint [@lee2026onemcmc], gives the conditions under
which the fixed-effect estimate equals $\bar\beta$; [One stacked fit and the
Rubin point estimate](m3-stacked-bridge.html#point-identity) explains it.
[Calibrating the fixed-effect draws (CCC)](m4-ccc-calibration.html) shows how
the stacked draws are given the covariance $T_{\text{MI}}$.

The numbers above were computed in this session:

```{r session-info}
sessionInfo()
```

## References
