---
title: "M3: The stacked fractional bridge — one fit, the Rubin point, and where it stops"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  The third page of the Method track derives the bridge that lets a single model
  fit stand in for the multiple-imputation point estimate. It explains what
  "stacked fractional" means — one model fit on all M plausible values at once,
  with every row down-weighted by one over M — and writes the stacked objective
  as the average of the per-plausible-value likelihoods. From that objective it
  states the point identity of the companion paper's Theorem 2.2: under a short
  list of regularity conditions the stacked fixed-effect estimate equals the
  Rubin average that M2 derived, and it sketches in prose why this holds. It is
  careful about scope: this is a fixed-effect point identity, exact only under
  those conditions and approximate in production, and it says nothing about
  variance or coverage. A read-only panel confirms the identity on the cached
  fixture under the reportable centering convention.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{M3: The stacked fractional bridge — one fit, the Rubin point, and where it stops}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

M2 built the external variance answer — the BRR–Fay sandwich pooled by the Rubin
rules into $T_{\text{MI}}$, with the Rubin mean $\bar\beta$ falling out as the
reportable *point*. It got there the orthodox way: fit the model once per
plausible value and combine the $M$ results [@mislevy1991randomization]. This page
asks a different question. Can a **single** model fit — on all $M$ plausible
values stacked together — recover that same point $\bar\beta$, without ever forming
the $M$ separate fits and averaging them? The answer is the **stacked fractional
bridge**, and the identity it delivers (Theorem 2.2) is the subject of M3.

Two warnings up front, because they bound everything below and the rest of the
page keeps returning to them. First, what the bridge delivers is a **point**
identity for the fixed effects, *not* a variance or coverage statement — the
variance answer is M2's external target, and the calibration that delivers it
to the stacked fit is M4. Second, "one stacked fit instead of $M$" is a
statement about **architecture**, the shape of the computation, and never a
claim about speed.

::: {.alert .alert-info}
**Scope of this page.** M3 derives the bridge (EQ-BRIDGE) and the point identity
(EQ-THM22), and verifies the identity on the bundled synthetic fixture. It
estimates nothing — there is no MCMC anywhere below, only read-only accessor
calls and a `data.frame`. The *live* stacked fit is produced by **brms/Stan**
[@burkner2017brms]; this vignette runs none of it, reading instead from the
cached fixture. The variance side is **M2** ($T_{\text{MI}}$) and **M4** (the CCC
calibration); which fits may carry a coverage claim is **M5**.
:::

The shared light-path load — the same one used across the Method track — gives us
the cached fit and its external target to point at:

```{r common-load}
fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit
tg <- get_target(fit)   # external Rubin / BRR–Fay target (M2's object)

c(M = tg$M, R = tg$R, fay_k = tg$fay_k)   # 2, 4, 0.5
```

The design is the tiny synthetic one M1 introduced: $M = 2$ plausible values,
$R = 4$ replicate weights, Fay coefficient $k = 0.5$.

## 1. The stacking idea

Start from the orthodox picture (M2, EQ-MODEL). You have $M$ plausible values per
student, and the multiple-imputation recipe fits the model **$M$ separate times**
— once to each plausible value — producing $M$ per-PV estimates
$\hat\beta_1,\dots,\hat\beta_M$, which the Rubin rules then average into
$\bar\beta$. The fit is repeated; the combining happens afterwards.

Stacking rearranges the *topology* of that computation. Instead of $M$ tables of
$N$ students fed to $M$ fits, lay all $M$ plausible values **end to end into one
long table** of $N\cdot M$ rows: student $ij$ appears $M$ times, once carrying
$y_{ij}^{(1)}$, once $y_{ij}^{(2)}$, and so on, with the covariates
($x_{ij}$, the cluster mean $\bar x_{\cdot j}$, the cluster index $j$) copied
across all $M$ of its rows. Then fit the model **once** to this stacked table.

If you stacked naïvely and stopped there, each student would count $M$ times and
the likelihood would behave as if the sample were $M$ times larger — wrong. The
fix is the **fractional** part: give every row weight $1/M$. With that weight a
student contributes a total mass of $M\cdot(1/M)=1$ across its $M$ copies, exactly
one student's worth, and the stacked log-likelihood becomes the *average* of the
$M$ per-PV log-likelihoods rather than their sum. That weighting is the entire
content of "stacked fractional": one design of $N\cdot M$ rows, each down-weighted
by $1/M$.

::: {.alert .alert-warning}
**"One stacked fit instead of $M$" is topology, not speed.** It describes the
shape of the computation — a single fit on $N\cdot M$ stacked rows in place of $M$
fits on $N$ rows each — not a benchmarked runtime. This vignette makes **no**
"faster" claim and reports no timings; the comparison is architectural, exactly
as A4 §7 frames the "fit count" column. The point of the bridge is *what the
single fit recovers*, which is the rest of this page.
:::

Two pieces of the M1 vocabulary do the work here. $L_m^{0}(\psi)$ is the
**unweighted per-PV likelihood** M1 §4 named — the likelihood of the parameters
$\psi$ under EQ-MODEL fit to plausible value $m$ alone. And $N$ is the number of
students, so the stacked design has the $N\cdot M$ rows just described. The next
section writes the stacked objective in those symbols.

## 2. The stacked fractional objective

Stacking the $M$ plausible values and weighting every row by $1/M$ turns the
per-PV likelihoods of Section 1 into a single objective. Writing
$q_{\text{SWL}}(\psi)$ for the **stacked weighted likelihood** at parameters
$\psi$, the bridge — **tag EQ-BRIDGE** — is

$$
\log q_{\text{SWL}}(\psi)=\log p(\psi)+\tfrac1M\textstyle\sum_m \log L_m^{0}(\psi)+C.
$$

Read it left to right:

- $\log p(\psi)$ is the **prior** — the same prior the live brms fit places on the
  parameters [@burkner2017brms]. Under the Theorem 2.2 conditions below it is flat
  on the fixed effects, so it shifts the objective by a constant and does not move
  the fixed-effect optimum.
- $\tfrac1M\sum_m \log L_m^{0}(\psi)$ is the **averaged per-PV log-likelihood** —
  the heart of the bridge. Each $\log L_m^{0}(\psi)$ is one plausible value's
  contribution; the $\tfrac1M$ in front is precisely the row weight from Section 1
  carried through to the log scale. This is what makes the stacked likelihood the
  *average* per-PV likelihood rather than their product or their sum: the $1/M$
  weight is the whole reason the $\sum_m$ is an average.
- $C$ collects terms constant in $\psi$ (the row-weighting and stacking constants);
  it does not affect the location of the maximum.

So $q_{\text{SWL}}$ is exactly the object Section 1 described — one fit on the
stacked rows, each weighted $1/M$ — written as a likelihood. The fixed-effect
estimate it produces when maximised, $\hat\beta_{\text{FE}}^{q}$, is the quantity
the next section identifies with the Rubin mean. (M1 introduced both
$q_{\text{SWL}}$ and $L_m^{0}$ in the symbol table; M3 is where they are assembled.)

## 3. Theorem 2.2 — the point identity

The reason to build $q_{\text{SWL}}$ is that maximising it returns the Rubin point.
Let $\hat\beta_{\text{FE}}^{q}$ be the fixed-effect estimate that maximises the
stacked objective EQ-BRIDGE. The companion methods paper's **Theorem 2.2** —
**tag EQ-THM22** — states that, under the regularity conditions listed below,

$$
\hat\beta_{\text{FE}}^{q}=\bar\beta,
$$

where $\bar\beta=\tfrac1M\sum_m\hat\beta_m$ is the Rubin mean from M2 (EQ-TMI)
[@rubin1987multiple]. In words: **one stacked, $1/M$-weighted fit recovers the
multiple-imputation point estimate** that the orthodox $M$-fits-and-pool recipe
produces — the same point, reached from the other direction.

The identity is exact only under a set of **regularity conditions, labelled
collectively R0–R5** in the canonical facts sheet. They are:

- a **flat prior** on the fixed effects (so $\log p(\psi)$ does not move the
  fixed-effect optimum);
- a **linear-Gaussian** model (EQ-MODEL with Gaussian residuals);
- a **common design $X$** across plausible values (the same covariate matrix in
  every $L_m^{0}$ — the plausible values differ only in the outcome column);
- **PV-invariant cluster means** $\bar x_{\cdot j}$ (computed once from the full
  sample and reused across all $m$, exactly as M1 §3 and EQ-MODEL require);
- a **common plug-in covariance $V$** (the same residual/variance structure used
  for every plausible value, not re-estimated per PV);
- **equal PV weight $1/M$** on every row (the fractional weighting of Section 1);
  and
- **identifiability** of the fixed-effect block.

These are stated as a collective list, not a one-to-one numbered map: the facts
sheet gives the seven assumptions under the single heading "R0–R5" without
assigning each a distinct index, so this vignette does **not** invent a mapping
of the form "R0 = flat prior, R1 = …". The collective label is what is locked.

## 4. Why it holds

The full proof is the companion methods paper's; what follows is a **sketch** — an
honest gesture at the mechanism, not a derivation. The intuition rests on two of
the conditions above, the **common design $X$** and the **common plug-in
covariance $V$**.

Maximising a (log-)likelihood means setting its **score** — the gradient in the
parameters — to zero. Because EQ-BRIDGE is a $1/M$-weighted *sum* of the per-PV
log-likelihoods, and differentiation is linear, the score of the stacked objective
is the $1/M$-weighted **average of the per-PV scores**:

$$
\text{(stacked score)} = \tfrac1M\textstyle\sum_m \text{(score of } \log L_m^{0}).
$$

Now use the two conditions. With a linear-Gaussian model, a common design $X$, and
a common plug-in covariance $V$, each per-PV score for the fixed effects is
*linear in its data* and *shares the same curvature* across plausible values — the
per-PV estimating equations differ only through the outcome column $y^{(m)}$, not
through $X$ or $V$. Two consequences follow. First, the root of an individual
per-PV score is that PV's estimate, $\hat\beta_m$. Second, because the equations
are linear in the outcome and share a common left-hand side, the root of the
**average** score is the **average** of the roots. Setting the stacked
(averaged) score to zero therefore yields

$$
\hat\beta_{\text{FE}}^{q} = \tfrac1M\textstyle\sum_m \hat\beta_m = \bar\beta,
$$

which is EQ-THM22. The flat prior is what lets us ignore $\log p(\psi)$ in the
fixed-effect optimum, and identifiability is what makes the root unique. Strip
away the common $X$ or the common $V$ — let the design or the plug-in covariance
vary by plausible value — and "the root of the average equals the average of the
roots" no longer holds exactly, which is precisely where Section 5 picks up. Again:
this is a sketch of the mechanism; the rigorous statement and proof live in the
companion methods paper (in preparation).

## 5. Where the identity stops

Theorem 2.2 is powerful but **narrow**, and reading it past its scope is the error
this page is most concerned to prevent. Three boundaries.

**It is a point identity, not a variance or coverage statement.** EQ-THM22 says
the stacked fit recovers the Rubin *point* $\bar\beta$. It says **nothing** about
the standard error, the interval, or the coverage of that point. The variance
answer is a different object entirely — M2's external $T_{\text{MI}}$ — and it
reaches the stacked fit only through M4's CCC calibration (EQ-CCC), not through
the bridge. The bridge gives you *where the estimate is*; M2 and M4 give you *how
uncertain it is*. Conflating the two is the central misreading to avoid.

**It is exact only under R0–R5.** The equality in Section 3 holds *as an identity*
only when all the listed conditions are met — in particular a common design and a
common plug-in covariance across plausible values, and a flat prior on the fixed
effects.

::: {.alert .alert-warning}
**EQ-THM22 is a fixed-effect *point* identity, exact only under R0–R5, and
approximate in production.** In a real fit the variance components
$(\sigma^2,\sigma^2_{\text{sch}})$ of EQ-MODEL are estimated **per plausible
value** (production uses per-PV REML), so the plug-in covariance $V$ is **not**
exactly common across the $M$ values — one of the R0–R5 conditions is only
approximately met. The "root of the average equals the average of the roots"
argument of Section 4 then holds **approximately, not exactly**: on real PISA the
stacked point and the Rubin mean agree closely but need not be byte-identical.
This is the **guardrail** framing of the facts sheet (§3, §7) — *empirical and
approximate in production* — and deliberately **not** an "exact, verified"
blueprint claim. The identity is about the **fixed-effect point only**; it is
**never** a variance or coverage result.
:::

**It is not a "one-MCMC, same-inference" claim.** That the *point* coincides does
**not** mean a single stacked fit reproduces the full $M$-fit *inference*. The
inference — the variance, the degrees of freedom, the interval, the coverage flag
— comes from M2's external target delivered through M4's CCC, and is governed by
the provenance rules of M5. The bridge buys you the point; it does not buy you
the interval.

## 6. Verify the point identity on the fixture

The cached `stack_direct` fixture lets us see EQ-THM22 realised. We read the
stacked fit's reported estimate with `get_estimates()` and the external target's
Rubin mean with `get_target()`, and check that they coincide. This is read-only —
no estimation, no MCMC — and is the same accessor pattern A1–A3 use.

```{r point-identity}
est <- get_estimates(fit)

all.equal(est$estimate, unname(tg$beta_bar))   # TRUE  -- EQ-THM22 realized on the fixture
data.frame(
  term             = est$term,
  stacked_estimate = est$estimate,
  rubin_beta_bar   = unname(tg$beta_bar)
)
```

The stacked fit's reported fixed-effect estimates are **term-by-term equal** to
the Rubin mean $\bar\beta$ from M2 — EQ-THM22 realised on this fixture. The point
identity holds.

::: {.alert .alert-warning}
**Framing precision — what this panel does and does not prove.** This fixture was
fit under the **reportable centering convention** `center = "target"` (the value
recorded in `fit$control$center`). Under that convention the stacked fit is
*centred on the external target by construction*, so the reported estimate **is**
$\bar\beta$ **by construction** — the agreement above is therefore EQ-THM22
**operationalised** on the fixture: the reportable-convention realisation of the
theorem, plus the theorem itself. It is **not**, and must not be read as, an
independent measurement of the *live-fit* identity — that would require maximising
the stacked objective EQ-BRIDGE from scratch and comparing to a separately pooled
$\bar\beta$, which is the companion paper's job, not this synthetic fixture's. A4
§3 forwards exactly this point and is careful to say the live-fit identity holds
**"approximately, not exactly"** in production (Section 5); M3 says the same.
:::

It is worth making explicit *which* quantity the bridge supplies and which it does
not. The reported **point** is the bridge's contribution (EQ-THM22). The reported
**standard error**, by contrast, is read straight off the *external* target — it
is $\sqrt{\operatorname{diag}(T_{\text{MI}})}$ from M2, not anything the bridge
produces:

```{r se-is-external}
all.equal(est$se, unname(sqrt(diag(tg$T_MI))))   # TRUE  -- the SE is M2's target, not the bridge's
```

The standard errors match the external target exactly because that is where they
come from. This is the division of labour Section 5 insisted on, made concrete:
**the bridge gives the point; M2 (the target) and M4 (the calibration) give the
variance.** The bridge contributes nothing to the SE column, and nothing in this
panel licenses a coverage claim — indeed this fixture carries
`coverage_claim_allowed == FALSE` under classic df (A3 §4, M5), so its agreement
is descriptive and algebraic, never coverage evidence.

## 7. Where to next

You now have the bridge in full — the stacked fractional objective (EQ-BRIDGE)
that averages the per-PV likelihoods, and the point identity (EQ-THM22) that
recovers the Rubin mean $\bar\beta$ from one stacked fit under R0–R5 — together
with its hard boundary: a fixed-effect *point* identity, exact only under those
conditions, approximate in production, and silent on variance and coverage.

- **M4 · CCC: Cholesky Calibration Correction** — the natural next step, and the
  *variance* counterpart to this page. Where Theorem 2.2 matched the **point**, the
  CCC affine map (EQ-CCC) matches the stacked draw cloud's first two **moments** to
  M2's external $T_{\text{MI}}$ — so the variance is delivered by *calibration*,
  not by the bridge.
- **M2 · The BRR–Fay fixed-effect target** — the other half of the answer: the
  $\bar\beta$ this identity recovers, and the $T_{\text{MI}}$ this identity does
  **not** give (EQ-TMI, EQ-FMI).
- **M5 · Methods, PSIS, and coverage** — the precise contrast of the three methods
  and why only `stack_direct` is coverage-claimable, the provenance that turns this
  point identity plus M4's calibration into a reportable interval.

To see how the point identity reads from the workflow side — the "agree on the
point, differ on the variance" pattern across methods, and the
`approximately, not exactly` caveat stated for a practitioner — return to **A4 ·
Comparing methods** (§3), which forwards its proof to this page.

Bug reports and feature requests:
<https://github.com/joonho112/pvstackr/issues>.

### Session info

```{r session-info}
sessionInfo()
```

## References
