---
title: "M5: Methods, PSIS, and coverage — choosing a path and what may be claimed"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  The capstone of the Method track, where the preceding four pages converge into
  a single contract: which of the three fitting methods may carry a coverage claim,
  and why only one of them can. It states the precise contrast between
  stack_direct, per_pv, and stack_psis, displays the Pareto-smoothed
  importance-sampling tail-shape thresholds that diagnose stack_psis, and argues
  that coverage is conferred by the design-based provenance of the external target
  alone — the M2 target reached through the M3 point identity and the M4
  calibration — not by any model-based covariance and not by a clean diagnostic. It
  is emphatic that a small Pareto tail-shape is necessary but not sufficient for a
  correct stack_psis variance, that the diagnostic is specification-dependent, and
  that the one-fit topology is an architecture statement and never a speed claim. A
  base-R classifier and a read-only flag check verify the rules on the bundled
  synthetic fixture.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{M5: Methods, PSIS, and coverage — choosing a path and what may be claimed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

The first four pages built the machinery one piece at a time. M2 derived the
**external variance answer** — the design-based BRR–Fay sandwich pooled by the
Rubin rules into $T_{\text{MI}}$ (EQ-TMI, EQ-BRRFAY), with the small-sample
Barnard–Rubin degrees of freedom (EQ-BARNARD) the path a coverage-claimable target
takes [@judkins1990fay; @rubin1987multiple; @barnard1999small]. M3 showed that one
stacked, $1/M$-weighted fit recovers the Rubin **point** $\bar\beta$ under the
regularity conditions of Theorem 2.2 (EQ-THM22) — a fixed-effect point identity,
no more. M4 showed how the CCC affine map (EQ-CCC) forces that one fit's draw
cloud to carry the target's **variance** too. This capstone closes the loop. It
asks the question the whole track has been circling: given three ways to fit the
model, *which one may carry a coverage claim, and why only that one?* The answer
is a contract about **provenance**, and the discriminator is where each method
gets its covariance.

Two warnings up front, because they bound everything below. First, coverage is a
property of **one path only** — `stack_direct` calibrated to the external,
design-based target — and even that path is descriptive under classic degrees of
freedom, exactly as the bundled fixture is. Second, the Pareto tail-shape
diagnostic this page introduces is a check on the *importance-sampling step*, not a
certificate that the resulting variance is right; reading it as the latter is the
single most seductive error in the chapter.

::: {.alert .alert-info}
**Scope of this page.** M5 states the three-method contrast, derives the PSIS
tail-shape thresholds (EQ-PSIS), and gives the coverage contract. It estimates
nothing — there is no MCMC anywhere below, only a pure base-R classifier and
read-only accessor calls against the bundled synthetic fixture. The full
three-method *comparison* — building three fits and running `pv_compare_methods()`
on injected draws — is **A4**'s job, and this page cross-references it rather than
duplicating its construction. The target is **M2**, the point identity is **M3**,
the calibration is **M4**; M5 is where their provenance is cashed out as the
`coverage_claim_allowed` flag.
:::

The shared light-path load — the same one used across the Method track — gives us
the cached fit and its external target to point at:

```{r common-load}
fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit
tg <- get_target(fit)   # external Rubin / BRR–Fay target (M2's object)

c(M = tg$M, R = tg$R, fay_k = tg$fay_k)   # 2, 4, 0.5
```

The design is the tiny synthetic one M1 introduced: $M = 2$ plausible values,
$R = 4$ replicate weights, Fay coefficient $k = 0.5$.

## 1. The three methods, precisely

pvstackr ships three fitting methods. They answer the *same* question — what is
the fixed-effect block $\beta_{\text{FE}}$? — but they get their **variance** from
different places, and that difference is the entire story of when a row may carry a
coverage claim. The locked method ↔ pipeline map (facts §4) lays them side by side:

| Package method | Blueprint pipeline | Covariance basis | Intervals | MCMC fits |
|---|---|---|---|---|
| `stack_direct` *(default)* | C-Direct | **design-based** external $T_{\text{MI}}$ | **conditional** — coverage-claimable only with an external Barnard–Rubin BRR–Fay target; descriptive under classic df | 1 |
| `per_pv` | A (orthodox) | model-based (posterior) | descriptive/reference | $M$ |
| `stack_psis` | B | model-based (posterior) | descriptive/reference | 1 (+PSIS) |

In one sentence each:

- **`stack_direct`** fits one stacked model and calibrates it — via CCC (M4,
  EQ-CCC) — to the **external, design-based** Rubin / BRR–Fay target (M2, EQ-TMI).
  Its point is the Rubin mean by the stacked bridge (M3, EQ-THM22) and its
  covariance is the design-based $T_{\text{MI}}$. Because that covariance is
  design-based, `stack_direct` rows are the only ones the package ever marks
  `coverage_claim_allowed = TRUE` [@judkins1990fay; @rubin1987multiple].
- **`per_pv`** is the orthodox multiple-imputation path: fit the model once per
  plausible value and pool the $M$ model-based results with Rubin's rules. It is
  the natural **reference** against which `stack_direct`'s point is sanity-checked,
  but its covariance is *model-based* (the posterior's own spread), so its
  intervals stay descriptive.
- **`stack_psis`** reweights a single stacked draw cloud with per-PV
  Pareto-smoothed importance sampling [@vehtari2017practical; @vehtari2024pareto]
  and pools the reweighted summaries. One fit, model-based covariance — a
  **cross-check**, never the deliverable.

**The covariance basis is the discriminator.** Read the third column, not the
first: two of the three methods carry a *model-based* covariance (the posterior's
nominal spread), and only `stack_direct` carries the *design-based* external
$T_{\text{MI}}$. Everything in this page follows from that one column — the
"Intervals" column is a *consequence* of the "Covariance basis" column, never an
independent property. The remaining columns are secondary: the "MCMC fits" count
is a topology fact (Section 5), and the "Blueprint pipeline" labels (C-Direct, A,
B) are the companion paper's internal names, given here only so a reader of that
work can line the two vocabularies up. A4 §1 prints the same map from the workflow
side; this page supplies the reason the column that matters is the covariance
basis.

## 2. PSIS in `stack_psis`

`stack_psis` is the one method that needs a diagnostic of its own, because it does
something the other two do not: it makes a **single** stacked draw cloud stand in
for all $M$ plausible values by *reweighting* it. The reweighting is
Pareto-smoothed importance sampling (PSIS), and its health is read off a single
number with published thresholds.

The mechanism, in one line. To make one cloud represent plausible value $m$,
`stack_psis` forms **per-PV importance weights** — how much each draw should count
when standing in for PV $m$ — and then **smooths the tail** of those weights by
fitting a **generalized Pareto distribution** to the largest of them
[@vehtari2017practical; @vehtari2024pareto]. Raw importance weights have notoriously
heavy, unstable tails; the generalized-Pareto fit replaces the extreme weights with
fitted order statistics, taming that instability. The single diagnostic that falls
out of the fit is the estimated **tail-shape** $\hat k$ — the shape parameter of
the fitted Pareto tail — and it is read against the Vehtari et al. thresholds.

**EQ-PSIS — the PSIS Pareto tail-shape thresholds.** Per-PV importance weights are
smoothed by a generalized-Pareto fit; the tail-shape $\hat k$ is read against the
Vehtari et al. thresholds:

$$
\hat k < 0.5 \ \text{(good)};\qquad 0.5 \le \hat k < 0.7 \ \text{(borderline)};\qquad \hat k \ge 0.7 \ \text{(unreliable)}.
$$

Read the tiers as a statement about the **importance-sampling step**, and only
that step:

- $\hat k < 0.5$ — **good.** The fitted tail is light enough that the
  importance-sampling estimate has finite variance and converges at the usual rate;
  the reweighting is trustworthy.
- $0.5 \le \hat k < 0.7$ — **borderline.** The tail is heavy enough that
  convergence slows and the estimate should be treated with caution.
- $\hat k \ge 0.7$ — **unreliable.** The tail is so heavy that the
  importance-sampling estimate is no longer dependable; the reweighting has
  effectively broken down.

The thresholds are *boundaries between tiers*, and they are **right-open**: $\hat
k = 0.5$ is the first borderline value (not the last good one), and $\hat k = 0.7$
is the first unreliable value. Section 6's classifier encodes exactly that
convention. What $\hat k$ tells you is whether the **weights are well-behaved** —
nothing more. The leap from "the weights are well-behaved" to "the resulting
variance is correct" is precisely the trap Section 4 dismantles.

## 3. Why only `stack_direct` is coverage-claimable

Now the contract. A coverage claim is the assertion that a method's nominal-95%
intervals actually capture the truth about 95% of the time. The package licenses
that assertion on **one path only**, and the reason is provenance, assembled across
the whole track.

The chain is exactly the four pages you have read:

1. **The variance has the right provenance (M2).** The interval's covariance is
   $T_{\text{MI}}$, built from the *design* — the BRR–Fay replicate-weight sandwich
   (EQ-BRRFAY) pooled by the Rubin rules (EQ-TMI) — not from the model's nominal
   posterior. M2 §6 stated the empirical result qualitatively: a correct weighted
   point paired with *model-based* standard errors can **under-cover** severely,
   and substituting the design-based $T_{\text{MI}}$ **restores nominal coverage**.
   Provenance, not arithmetic, is what licenses the claim
   [@judkins1990fay; @rubin1987multiple].
2. **The right point reaches one fit (M3).** The stacked fractional bridge means
   that single fit reports the Rubin mean $\bar\beta$ as its point (EQ-THM22), so
   the calibrated answer is centred where the orthodox $M$-fit recipe would centre
   it.
3. **The right variance reaches that fit (M4).** CCC's affine map (EQ-CCC) forces
   the stacked draw cloud's covariance to equal that same external $T_{\text{MI}}$,
   so the one fit *reports the external answer in full*.
4. **The degrees of freedom are small-sample correct (M2).** A coverage-claimable
   interval needs the **Barnard–Rubin** degrees of freedom $\nu_{\text{BR}}$
   (EQ-BARNARD) [@barnard1999small], not the classic Rubin formula — the small-$M$
   correction the tails of the reference distribution require.

`per_pv` and `stack_psis` fail at the very first link: their covariance is
**model-based**, the posterior's own spread, which never sees the survey's
clustering and weighting. However sensible their points, their intervals cannot
bear a *design*-coverage claim, so the package marks them **descriptive always** —
regardless of which degrees of freedom they happen to use. This is the H3 rule, and
the package enforces it through metadata, not trust: the licensing decision lives
in the `coverage_claim_allowed` column, which Section 6 reads off the fixture.

::: {.alert .alert-warning}
**Coverage is `stack_direct`-only — and even `stack_direct` is descriptive under
classic df.** A row is `coverage_claim_allowed = TRUE` **only** when the fit is
`stack_direct` *and* it was calibrated to an external **Barnard–Rubin** BRR–Fay
target [@barnard1999small; @rubin1987multiple; @judkins1990fay]. `per_pv` and
`stack_psis` are **always** descriptive, whatever their df. And a `stack_direct`
fit built on **classic** Rubin df — like the bundled fixture — is descriptive too,
because classic df does not deliver the small-sample calibration the claim rests
on. The lesson A3 §4 stated for practitioners holds here as a rule: **read the
flag, never infer coverage from the method name.** This fixture *is* `stack_direct`
and is still, honestly, descriptive (Section 6).
:::

It is worth being explicit about what does **not** confer coverage, because the
preceding pages each carried a version of this caveat. The stacked point identity
(M3) is a *point* statement and says nothing about variance. The CCC moment match
(M4) is *algebraic by construction* — it forces the calibrated cloud's mean and
covariance to equal the target's, so its near-perfect agreement (the $\sim10^{-14}$
center-separation numbers of M4 §6) confirms only that the linear algebra ran
cleanly, **not** that the intervals cover. Coverage comes from the *provenance* of
$T_{\text{MI}}$ and the companion paper's simulation oracle, never from a clean
diagnostic. M5 collects those caveats into a single contract: coverage is licensed
by the whole design-based chain, certified by the flag.

## 4. Caution — small $\hat k$ ≠ correct `stack_psis` variance

The most dangerous misreading in this track concerns `stack_psis`, and it is
dangerous precisely because the diagnostic that looks reassuring is reassuring
about the *wrong thing*. This is the H4 caution, and it has two parts.

**A small $\hat k$ certifies the weights, not the variance.** Section 2 was careful:
$\hat k$ diagnoses the **importance-sampling step** — a small $\hat k$ says the
weights are well-behaved and the reweighting is stable. It says **nothing** about
whether the resulting **between-imputation variance** is right. The reason is
structural. `stack_psis` reweights **one** stacked draw cloud to stand in for all
$M$ plausible values, so the per-PV "imputations" are all drawn from the *same*
cloud and are therefore **correlated**. But Rubin's between-imputation covariance
$B$ (EQ-TMI) assumes the imputations are **independent**; correlated ones make $B$
shrink, and a shrunk $B$ shrinks the pooled total $T_{\text{MI}} = \bar U +
(1+1/M)B$. The `stack_psis` interval can therefore run **narrow even when $\hat k$
is small** — the importance sampling is stable, and the variance is still wrong.
A green $\hat k$ is **necessary but not sufficient** for a trustworthy interval.

**And $\hat k$ is specification-dependent.** The diagnostic is not a fixed property
of the method; it depends on the model you actually fit. It can sit comfortably in
the *good* tier on a simplified, low-dimensional model and climb into the
*unreliable* tier on a full production specification — the regime where PSIS is
formally broken. A green $\hat k$ on a toy model is no guarantee of a green $\hat k$
on the model you report.

::: {.alert .alert-warning}
**Small Pareto-$\hat k$ ≠ correct `stack_psis` variance; and $\hat k$ is
specification-dependent.** Two distinct failures: (i) the correlated-imputation
narrowing above means a stable importance-sampling step can sit on top of a *too
small* between-imputation variance; and (ii) the tail-shape itself is **not** a
stable number — it is small in the simplified two-covariate reading demo but climbs
**above 1** in the full paper specification, where Pipeline B is unreliable. Both
behaviours are described **qualitatively** in the **companion methods paper (in
preparation)**; the package hard-codes **no** $\hat k$ value as a fact (the $\hat k$
values in Section 6's classifier are an *illustration of the thresholds*, exactly as
A4 §6 treats its injected `pareto_k = 0.2`). The standing instruction follows:
**treat `stack_psis` as a cross-check, never the deliverable**
[@vehtari2017practical; @vehtari2024pareto].
:::

This is why the rule of thumb is unambiguous (A4 §7): report `stack_direct`, keep
`stack_psis` as a one-fit cross-check while watching $\hat k$ — and *not trusting a
small $\hat k$ alone* — and use `per_pv` as the orthodox reference for the point.

## 5. The cost/benefit — `M` fits → 1

The map's "MCMC fits" column records a genuine difference between the methods, and
it is worth reading correctly because it is the one column most easily misread as a
performance claim. It is **not** one.

The orthodox `per_pv` path fits the model **$M$ separate times** — once per
plausible value — and pools afterward. The two stacked methods fit the model
**once**, on the $N\cdot M$ stacked rows with the $1/M$ row weighting M3 derived
(`stack_psis` adds the PSIS reweighting on top of that single fit). For PISA
reading, where $M = 10$ [@oecd2024pisa], that is the contrast between **ten** model
fits and **one** — "10 fits → 1".

::: {.alert .alert-warning}
**"$M$ fits → 1" is topology, not speed.** The fit count describes the **shape of
the computation** — how many model fits a method's architecture requires — not a
benchmarked runtime. "One stacked fit instead of $M$" is an **architecture /
topology** statement; the "10 fits → 1" figure is the $M = 10$ count for PISA
reading [@oecd2024pisa], **not** a claim that the stacked path is faster. The package
reports no timings and makes **no** efficiency or "faster" claim (the H5 guardrail;
A4 §7's honesty triad). The single stacked fit can be larger and slower per fit
than one per-PV fit — it carries $M$ times the rows — so the count is genuinely a
statement about *number of fits*, not wall-clock time.
:::

The benefit, then, is *not* "it runs faster." It is that the stacked architecture
collapses the multiple-imputation point recovery (M3) and the variance calibration
(M4) into a **single** object that can be calibrated to the external target and
licensed for coverage — the algebraic-equivalence headline the package may
honestly claim (facts §5), stated as an identity under stated conditions and
*never* as a speedup.

## 6. Verify in code

Two read-only panels close the page: a pure base-R classifier that encodes the
EQ-PSIS thresholds, and a flag read that confirms the coverage contract on the
fixture. Neither estimates anything.

**Panel 1 — the EQ-PSIS classifier.** This is the threshold logic of Section 2 in
one base-R function. The `right = FALSE` argument is what makes the tier boundaries
**right-open**, so $\hat k = 0.5$ lands in *borderline* and $\hat k = 0.7$ in
*unreliable*, matching EQ-PSIS exactly:

```{r psis-classifier}
## PSIS Pareto-k threshold classification (EQ-PSIS), pure base-R
classify_k <- function(k) {
  cut(k, breaks = c(-Inf, 0.5, 0.7, Inf),
      labels = c("good", "borderline", "unreliable"), right = FALSE)
}

data.frame(
  k       = c(0.04, 0.20, 0.49, 0.50, 0.69, 0.70),
  verdict = classify_k(c(0.04, 0.20, 0.49, 0.50, 0.69, 0.70))
)
```

The verdicts are good / good / good / borderline / borderline / unreliable: the
three values below $0.5$ are good, $0.50$ and $0.69$ are borderline, and $0.70$ is
unreliable. The boundary cases ($0.50$, $0.70$) fall into the *higher-risk* tier,
which is the conservative reading EQ-PSIS intends.

::: {.alert .alert-warning}
**These $\hat k$ values are an illustration of the thresholds, not package-measured
facts.** The numbers above are hand-chosen to exercise each tier boundary of
EQ-PSIS — they are *not* read from any fit, and the package hard-codes no $\hat k$
as a result (H4; facts §7). This is the same posture A4 §6 takes with its injected
`pareto_k = 0.2`: a value supplied to demonstrate the *machinery*, never asserted
as a measurement. On a real `stack_psis` fit you would read the realized
`pareto_k_max` off the comparison diagnostics (A4 §4) and classify *that* — and,
per Section 4, you would still not trust a small value on its own.
:::

**Panel 2 — read the coverage flag, not the method name.** The fixture is a
`stack_direct` fit, yet the licensing decision is recorded in the estimate table's
metadata, and it is `FALSE`. We read the flag directly (this is the H3 rule made
concrete, mirroring A3 §4):

```{r coverage-flag}
est <- get_estimates(fit)

unique(est$interval_role)            # "descriptive_classic_rubin"
unique(est$coverage_claim_allowed)   # FALSE -- even though method == "stack_direct"
est[, c("term", "interval_role", "df_method", "coverage_claim_allowed")]
```

Every row reads `interval_role == "descriptive_classic_rubin"`,
`df_method == "classic"`, and `coverage_claim_allowed == FALSE` — across all three
fixed-effect terms. This is the contract of Section 3 made visible: the fit *is*
`stack_direct` (the default, coverage-*capable* method), but because it carries
**classic** Rubin df rather than the Barnard–Rubin correction, it is descriptive,
and the flag says so. Read the column, never the method name.

For the **full three-method comparison** — building a `per_pv` reference and a
`stack_psis` cross-check from injected draws, aligning all three with
`pv_compare_methods()`, and reading `interval_role` / `coverage_claim_allowed` /
`pareto_k_max` per method — see **A4 §3–§6**, which already does exactly that on the
light path. This page deliberately does **not** rebuild it: M5 states the coverage
*rules*; A4 runs the worked comparison.

## 7. Where to next

You now hold the coverage contract in full — the three-method map whose
discriminator is the **covariance basis** (Section 1), the PSIS tail-shape
thresholds that diagnose `stack_psis` (EQ-PSIS), the provenance chain that makes
**only** `stack_direct` coverage-claimable (Section 3), the trap that a small $\hat
k$ is necessary but not sufficient and is specification-dependent (Section 4), and
the topology — not speed — reading of "$M$ fits → 1" (Section 5). That contract is
the capstone the whole Method track was built toward: M2's target, M3's point, and
M4's calibration converge here into the single flag the package reports.

**The Method track is complete.** From here the natural move is back to the Applied
track, where these rules are put to work:

- **A4 · Comparing methods** — the worked three-method comparison this page states
  the rules for: `pv_compare_methods()` on the light path, the agreement
  diagnostics, and the two cautions (design variance is not optional; small
  Pareto-$\hat k$ is not enough), with its proofs forwarded to this page and to M3.
- **A5 · Real PISA data guidance** — pointing the same workflow at a genuine PISA
  cycle: licensing and non-affiliation, the design declaration, the $M(R+1)$
  replicate-fit footprint, and the reproducibility checklist — the place the
  $M = 10$ scale behind "10 fits → 1" becomes a real job to size.

To see the licensing flag itself from the reporting side — the `interval_role`,
`coverage_claim_allowed`, `df_method`, and `df_complete` columns, and the rule to
read the flag rather than the method name — return to **A3 · Reading results** (§4).

And for the provenance behind the contract: **M2 · The BRR–Fay fixed-effect
target** (the design-based $T_{\text{MI}}$ and the Barnard–Rubin df that license
coverage), **M3 · The stacked fractional bridge** (the point identity), and **M4 ·
CCC: Cholesky Calibration Correction** (the variance match that is *not* the source
of coverage).

Bug reports and feature requests:
<https://github.com/joonho112/pvstackr/issues>.

### Session info

```{r session-info}
sessionInfo()
```

## References
