---
title: "A1: Getting started — your first calibrated plausible-value fit"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  The fastest path into pvstackr. In a few lines you install the package, see the
  single `pv_fit(method = "stack_direct")` call shape, load the bundled cached
  fixture fit, and read it three ways: the console `print()` box, the tidy
  `get_estimates()` table, and one figure. You leave knowing what a calibrated
  fixed-effect fit needs, how to read its intervals honestly, and where the full
  end-to-end workflow lives next.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{A1: Getting started — your first calibrated plausible-value fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

## 1. Why you are here

Large-scale assessments such as PISA do not report a single test score per
student. Because proficiency is a latent trait that is never observed directly,
the score is published as a set of *plausible values* (PVs): $M$ draws from each
student's posterior proficiency distribution [@mislevy1991randomization;
@vondavier2009plausible]. To honour the uncertainty those draws encode, an analysis must
fit the model once per PV and combine the $M$ results with multiple-imputation
(MI) rules [@rubin1987multiple]. The common shortcut — fit one model to `PV1` and
stop — silently discards the between-imputation variance and reports intervals
that are too narrow.

**pvstackr** gives you an honest fixed-effect fit without that shortcut and
without the bookkeeping. Its default method, `stack_direct`, produces a
fixed-effect report from **one** stacked model fit that is calibrated to an
external Rubin / BRR–Fay target. The stacked fit supplies the model-side draw
cloud and agreement diagnostics; the reportable fixed-effect estimate and
standard error come from the external target. Intervals are coverage-claimable
only on the external design-based BRR-Fay / Barnard-Rubin path; otherwise the
metadata marks them descriptive.

Two scope notes, stated up front so nothing later surprises you:

- **Fixed effects only.** The package calibrates and reports the fixed-effect block
  $\beta_{\text{FE}}$ (intercept and slopes). Variance components are estimated
  but **not** calibrated, and are not part of the reportable output.
- **"One fit" is an architecture statement, not a speed claim.** It describes the
  *topology* of the method (one stacked fit instead of $M$ separate ones); this
  vignette makes no benchmarked timing claims.

The road map for this article: install → the one call → load a cached fit →
read it three ways → read the intervals honestly → where to go next.

## 2. What you will leave with

After this vignette you will be able to:

- recognise the single canonical `pv_fit(method = "stack_direct")` call and name
  the pieces it needs (an external target and a model backend);
- load the bundled cached fixture fit and read it three ways — the console
  `print()` box, the tidy `get_estimates()` table, and one coefficient figure;
- tell at a glance, from the estimate table itself, whether a row's interval is
  **coverage-claimable** or merely **descriptive**, and say why;
- know exactly which companion vignette to open for the full workflow, the
  accessor tour, method comparison, real PISA data, and the underlying method.

This article runs in **seconds** because it never fits a model: it loads a
precomputed fit object and reads off it. There is no live MCMC anywhere below.

## 3. Install and load

Once the package is on CRAN:

```{r install-cran, eval = FALSE}
# install.packages("pvstackr")
```

Until then, install the development version from GitHub:

```{r install-dev, eval = FALSE}
# install.packages("pak")
pak::pak("joonho112/pvstackr")
```

The one line to remember — already run in the setup above — is:

```{r load-again, eval = FALSE}
library(pvstackr)
```

Reading a cached fit (everything in this vignette) needs **no** modelling
backend and no extra dependencies. A *live* fit on your own data additionally
needs a fitting engine; that is the subject of vignette A2.

## 4. The one call (conceptual)

Every pvstackr analysis funnels through a single dispatcher, `pv_fit()`. Its
canonical `stack_direct` form looks like this. The chunk is **not run** here —
it only shows the *shape* of the call, since a real fit needs an external target
and a backend that this article does not assemble:

```{r the-one-call, eval = FALSE}
fit <- pv_fit(
  data          = your_data,
  formula       = OUTCOME ~ x + female,           # OUTCOME is a placeholder
  target        = your_target,                    # from pv_brr_target()
  method        = "stack_direct",
  control       = pv_control(method = "stack_direct", backend = "brms")
)

# Or drive your own engine, which takes all three adapter functions:
fit <- pv_fit(
  data              = your_data,
  formula           = OUTCOME ~ x + female,
  target            = your_target,
  method            = "stack_direct",
  control           = pv_control(method = "stack_direct", backend = "cmdstanr"),
  fit_function      = your_fit_function,          # estimates the stacked model
  draws_function    = your_draws_function,        # extracts posterior draws
  diagnose_function = your_diagnose_function,     # reports sampler diagnostics
  cache_dir         = NULL
)
```

A few words on the pieces:

- **`OUTCOME` is a placeholder.** Internally pvstackr substitutes it with each
  plausible-value column in turn (`PV1READ`, `PV2READ`, …) so you write the model
  once, not $M$ times.
- **`target =`** supplies the external Rubin / BRR–Fay variance target built by
  `pv_brr_target()`. `stack_direct` intervals are coverage-claimable only when
  backed by an external design-based BRR-Fay target with Barnard-Rubin degrees of
  freedom; without that coverage path the fit falls back to descriptive
  intervals.
- **`fit_function =`** is the modelling backend that actually estimates the
  stacked model.

Assembling the target and a backend end to end is the job of **vignette A2 (the
end-to-end workflow)**. Here we skip straight to a fit that has already been run
for you.

## 5. Load the cached fixture fit

The package ships a small synthetic dataset and a cached `stack_direct` fit so
that every example runs offline and deterministically. Load both:

```{r load-fixture}
pisa_tiny <- read.csv(
  system.file("extdata", "pisa_tiny.csv", package = "pvstackr")
)

fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit
```

The fixture is a deliberately tiny, **synthetic** PISA-shaped table — one
make-believe country (`CNT == "SYN"`), `r nrow(pisa_tiny)` students nested in a
handful of schools. It carries one continuous covariate `x`, a binary `female`
indicator, **two** plausible reading values `PV1READ` / `PV2READ` (so $M = 2$),
**four** BRR replicate weights `W_FSTURWT1`–`W_FSTURWT4` (so $R = 4$), a final
weight `W_FSTUWT`, and student / school identifiers:

```{r fixture-peek}
str(pisa_tiny)
```

::: {.alert .alert-warning}
These are **synthetic, illustrative** numbers — included only so examples and
package checks run without licensed data. They are **not** real PISA results and
must never be cited as such. Real PISA reading uses $M = 10$ plausible values and
$R = 80$ replicate weights; here $M = 2$ and $R = 4$ keep the object small. For
working with genuine PISA files, see vignette A5.
:::

Everything below simply *reads* the `fit` object. Nothing is re-estimated.

## 6. Read it three ways

### 6.1 The console — `print()`

The quickest look is the result-first `print()` box, which leads with the answer
and the things you most need to know before trusting it:

```{r print-fit}
print(fit)
```

It tells you the method (`stack_direct`), that the fit converged (`status: ok`),
that there are 3 fixed effects, the target it was calibrated against
(`external_brr_fay_rubin`), that posterior draws were not retained, which
diagnostics are attached, and — crucially — an interval note. For this fixture
the note reads that *intervals are descriptive rather than coverage-claimable*;
Section 7 explains why, and how to change it.

### 6.2 The tidy table — `get_estimates()`

`get_estimates()` returns the reportable data frame: one row per fixed-effect
term, with the estimate, its standard error, the confidence interval, and the
metadata that says how far you may trust that interval. The full table is wide;
here is the compact view most analyses start from:

```{r get-estimates}
est <- get_estimates(fit)

est[, c("term", "estimate", "se", "df",
        "conf_low", "conf_high",
        "interval_role", "coverage_claim_allowed")]
```

The three terms are `b_Intercept`, `b_x`, and `b_female`. On this synthetic
fixture the intercept sits near 458 score points, the slope on `x` near 47
points per unit, and the `female` coefficient near 2 points (with a wide
interval — see below). The metadata columns (`interval_role`,
`coverage_claim_allowed`, and the degrees-of-freedom machinery) are the heart of
honest reporting; the full column-by-column tour, including the *fraction of
missing information*, lives in **vignette A3**.

### 6.3 One figure

A dot-and-interval plot makes the slope estimates easy to read at a glance. We
plot **only the slope coefficients** (`b_x` and `b_female`) and drop the
intercept, whose scale (~458) would otherwise flatten everything else. A
reference line at zero marks the no-effect value:

```{r coef-figure, fig.width = 7, fig.height = 3, fig.cap = "Slope coefficients from the cached synthetic stack_direct fit, with 95% descriptive intervals. The intercept is omitted because its scale (~458 score points) would dominate the axis. The dashed line marks zero (no effect). These are illustrative synthetic values, not real PISA estimates.", fig.alt = "A horizontal dot-and-interval plot of two slope coefficients from the synthetic fixture fit. The coefficient on x is about 47 score points with a narrow interval well to the right of zero. The coefficient on female is about 2 score points with a wide interval spanning zero from roughly minus 42 to plus 46. A dashed vertical reference line is drawn at zero."}
slopes <- est[est$term != "b_Intercept", ]
slopes <- slopes[order(slopes$term), ]

y    <- seq_len(nrow(slopes))
xlim <- range(c(slopes$conf_low, slopes$conf_high, 0))

op <- par(mar = c(4.5, 7, 1, 1))
plot(
  slopes$estimate, y,
  xlim = xlim, ylim = c(0.5, nrow(slopes) + 0.5),
  yaxt = "n", ylab = "",
  xlab = "Coefficient (synthetic reading-score points)",
  pch = 19, cex = 1.4, col = "#1f6f9c"
)
abline(v = 0, lty = 2, col = "grey50")
segments(slopes$conf_low, y, slopes$conf_high, y, lwd = 2, col = "#1f6f9c")
points(slopes$estimate, y, pch = 19, cex = 1.4, col = "#1f6f9c")
axis(2, at = y, labels = slopes$term, las = 1)
par(op)
```

The slope on `x` is clearly separated from zero; the `female` coefficient is
small and its interval comfortably spans zero — unsurprising in a 12-row
synthetic toy.

For completeness: this fit retained no posterior draws, so the draw accessor
returns `NULL`. That is expected for a cached `stack_direct` object and does not
affect the fixed-effect report:

```{r get-draws}
get_draws(fit)
```

## 7. How to read the intervals honestly

pvstackr never hides whether an interval can bear a coverage claim — it records
the answer in the table so you do not have to remember a rule. Read it directly:

```{r honest-columns}
est[, c("term", "interval_role", "coverage_claim_allowed")]
```

For this fixture every row shows `interval_role == "descriptive_classic_rubin"`
and `coverage_claim_allowed == FALSE`. The intervals are still useful for
*describing* the spread of the estimate — but they are **not** to be reported as
calibrated confidence intervals.

**The rule** is simple and is enforced by the metadata, not by trust:

> A row is `coverage_claim_allowed = TRUE` **only** when the fit is
> `stack_direct` *and* it was calibrated against an external **Barnard–Rubin**
> BRR–Fay target [@barnard1999small; @rubin1987multiple]. The `per_pv` and
> `stack_psis` methods are always descriptive. The bundled fixture, although
> `stack_direct`, uses **classic** Rubin degrees of freedom (`df_method ==
> "classic"`) rather than the Barnard–Rubin small-sample df, so it stays
> descriptive by design — a faithful demonstration of the object surface, not a
> coverage-claimable result.

Two practical consequences are visible in the table above:

- **Read the column, do not assume.** `coverage_claim_allowed` is the single
  field that licenses (or withholds) a confidence claim for each row. Let it
  decide.
- **Tiny $M$ means tiny df means wide intervals.** With only $M = 2$ plausible
  values, the between-imputation degrees of freedom are near 1, so the
  $t$-multiplier is enormous and the intervals — most visibly for `b_female`
  in the figure — are very wide. This is the MI machinery being honest about how
  little information two imputations carry, not a bug.

One related quantity worth a sentence: the **fraction of missing information**
(FMI) measures how much of a coefficient's uncertainty comes from imputation
rather than sampling. It drives the degrees of freedom and is reported alongside
the estimates; its full treatment — and how it interacts with
`coverage_claim_allowed` — is in **vignette A3**.

## 8. Where to next

You have loaded a fit and read it honestly. From here:

- **A2 · The end-to-end workflow** — declare the design with `pv_design()`,
  assemble the target with `pv_brr_target()`, run `pv_fit(method =
  "stack_direct")`, and report; plus the conceptual live-backend sketch and why
  reportable fits need `center = "target"`.
- **A3 · Reading results & what to report** — the full accessor family, every
  estimate-table column, the interval metadata in depth, the fraction of missing
  information, and a copy-paste reporting checklist.
- **A4 · Comparing methods** — `pv_compare_methods()` across `stack_direct`,
  `per_pv`, and `stack_psis`; agreement diagnostics; and the two cautions
  (design variance is not optional; a small Pareto-$\hat k$ does not guarantee
  correct variance).
- **A5 · Real PISA data guidance** — connecting to genuine PISA files without
  bundling licensed data: licensing and non-affiliation, the design declaration,
  memory and runtime, and a reproducibility checklist.

For the mathematics skipped here — the PV survey setting, the notation, and the
fixed-effect estimand — start the **Method track** with **M1 · Foundations and
notation**.

Bug reports and feature requests:
<https://github.com/joonho112/pvstackr/issues>.

### Session info

```{r session-info}
sessionInfo()
```

## References
