---
title: "Reading and reporting the results"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  This article reads a fitted object with `get_estimates()`, `get_target()`,
  `get_draws()` and `get_diagnostics()`. It explains the columns of the
  estimate table, including the columns that say whether pvstackr's reporting
  rule lets you read an interval as a confidence interval with nominal
  coverage, and shows where to find
  the fraction of missing information. It ends with a list of what to report.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Reading and reporting the results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

## Loading a fit

The examples read the `stack_direct` fit that ships with pvstackr:

```{r load-example-fit}
fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit
```

The example fit and its synthetic data ($M = 2$ plausible values, $R = 4$
replicate weights) are described in [Getting started with
pvstackr](a1-getting-started.html#example-fit).

## Functions that read a fit

Four functions read a fit. Use them rather than `$`: they stop with an error if
the fit was changed after it was created.

- `get_estimates(fit)` returns the estimate table: one row per fixed effect,
  with the estimate, its standard error, degrees of freedom and interval, and
  the columns that say how the interval can be read (see [the estimate
  table](#estimate-table)).
- `get_target(fit)` returns the target that the fit was calibrated to, the
  `pvstackr_brr_target` object made by `pv_brr_target()`. It holds the
  estimates combined over the plausible values, their covariance matrix
  $T_{\text{MI}}$, the degrees of freedom and the fraction of missing
  information.
- `get_draws(fit)` returns the calibrated fixed-effect draws of the stacked
  fit, or `NULL` when the fit was saved without them.
- `get_diagnostics(fit)` returns the results of pvstackr's checks and other
  records of the fit. For a `stack_direct` fit it is a list with six
  elements: `preflight`, `sampler`, `sampler_gate`, `stack_fit`,
  `stack_fit_warnings` and `ccc`.

[The target, the draws and the diagnostics](#target-and-diagnostics) shows the
last three in more detail. The example fit was saved without its draws
(`return_draws = FALSE` in `pv_control()`), so `get_draws()` returns `NULL`:

```{r get-draws}
get_draws(fit)
```

The estimate table does not need the draws, because for a `stack_direct` fit
its values come from the target.

## The estimate table {#estimate-table}

`get_estimates()` returns a data frame with one row per fixed effect. For a
`stack_direct` fit it has 17 columns:

```{r get-estimates}
est <- get_estimates(fit)

names(est)
```

- `term` names the coefficient: the column name of the model matrix with the
  prefix `b_`, here `b_Intercept`, `b_x` and `b_female`. Only the fixed
  effects (the intercept and the slopes) have rows; other parameters of the
  model, such as the residual standard deviation `sigma`, are not reported.
- `estimate`, `se` and `df` are the estimate, its standard error and its
  degrees of freedom.
- `conf_level`, `conf_low` and `conf_high` give the interval, and
  `df_method`, `df_complete`, `interval_role` and `coverage_claim_allowed`
  say how it was made and how it can be read (see [the interval
  columns](#interval-columns)).
- `parameter_scope`, `target_source` and `target_hash` say what each row is
  and where its numbers come from (see below).
- `std.error`, `conf.low` and `conf.high` repeat `se`, `conf_low` and
  `conf_high` under the names that the broom package uses for model
  summaries. pvstackr has no `tidy()` method; the repeated columns only let
  code written for these names read the table.

The first columns hold the numbers that most analyses start from:

```{r estimate-core}
est[, c("term", "estimate", "se", "df", "df_method")]
```

In the example the intercept is 457.9, the slope of `x` is 46.9 and the
coefficient of `female` is 2.1, with standard errors 1.29, 0.37 and 3.56. The
degrees of freedom are close to 1, which makes the intervals very wide. With
$M = 2$ plausible values the classic degrees of freedom cannot be smaller than
$M - 1 = 1$, and they are close to that value here because most of the
variance comes from the differences between the two plausible values (see
[the fraction of missing information](#fmi)).

The last three columns record what each row is and where it comes from:

```{r estimate-source-columns}
est[, c("term", "parameter_scope", "target_source", "target_hash")]
```

`parameter_scope` is `"fixed_effect"` in every row. `target_source` is
`"external_brr_fay_rubin"` for a `stack_direct` fit: the numbers come from the
target that `pv_brr_target()` computed with the BRR–Fay replicate weights and
Rubin's rules. `target_hash` is a SHA-256 checksum of the contents of that
target. The target stores the same value, and pvstackr uses it to detect a
target that was changed after it was made. Keep it with your results, since it
identifies the target they come from; [Real PISA data: access, memory and
reproducibility](a5-real-pisa-guidance.html#reproducibility-checklist) lists
the other records to keep.

For a `stack_direct` fit, `estimate`, `se` and `df` are copied from the target
([The full analysis workflow](a2-the-workflow.html#center-target)): `estimate`
is the target's $\bar\beta$, the average of the $M$ survey-weighted estimates,
one per plausible value, and `se` is $\sqrt{\operatorname{diag}(T_{\text{MI}})}$,
the square root of the diagonal of the target covariance matrix.

## Interval columns and pvstackr's reporting rule {#interval-columns}

Each row has an interval from `conf_low` to `conf_high`: the estimate plus or
minus a $t$ quantile with `df` degrees of freedom times `se`, at the level
`conf_level` set in `pv_control()` (0.95 by default). Four columns say how the
interval was made and how it can be read:

```{r interval-metadata}
est[, c("term", "interval_role", "df_method",
        "df_complete", "coverage_claim_allowed")]
```

- `df_method` is the rule for the degrees of freedom, chosen when the target
  is made with `pv_brr_target()`. `"classic"`, the default, is the rule of
  @rubin1987multiple, $(M - 1)/\lambda^2$, where $\lambda$ is the fraction of
  missing information of the coefficient (see [the fraction of missing
  information](#fmi)). `"barnard_rubin"` is the correction of
  @barnard1999small for small complete-data degrees of freedom ([The full
  analysis workflow](a2-the-workflow.html#target) describes both rules).
- `df_complete` is used only with `"barnard_rubin"`: the complete-data
  degrees of freedom of the replicate-weight variance, which you state when
  you make the target. They are the degrees of freedom the analysis would
  have if the outcome were observed directly instead of through plausible
  values, and the Barnard–Rubin adjustment matters when this number is
  small. pvstackr does not choose this number and does not check it against
  the replicate weights: `pv_brr_target()` accepts any positive value, `Inf`
  included. The value depends on how the survey formed its replicate weights,
  which the survey's technical documentation describes; report it with the
  intervals. With `"classic"` the column is `NA`, as here.
- `coverage_claim_allowed` is `TRUE` when pvstackr's reporting rule lets you
  read the interval as a confidence interval with nominal coverage (a
  coverage-claimable interval), and `FALSE` when pvstackr makes no statement about its
  coverage (a descriptive interval).
- `interval_role` names the case, as in the table below.

pvstackr sets `interval_role` and `coverage_claim_allowed` by its reporting
rule, from the method and `df_method`. For `stack_direct`, `df_method` and
`df_complete` are those of the target; `per_pv` and `stack_psis` take them as
their own arguments.

| Method | `df_method` | `interval_role` | `coverage_claim_allowed` |
|:--|:--|:--|:--|
| `stack_direct` | `"barnard_rubin"` | `coverage_barnard_rubin` | `TRUE` |
| `stack_direct` | `"classic"` | `descriptive_classic_rubin` | `FALSE` |
| `per_pv` | `"barnard_rubin"` | `reference_barnard_rubin` | `FALSE` |
| `per_pv` | `"classic"` | `reference_classic_rubin` | `FALSE` |
| `stack_psis` | `"barnard_rubin"` | `psis_barnard_rubin` | `FALSE` |
| `stack_psis` | `"classic"` | `psis_classic_rubin` | `FALSE` |

So `coverage_claim_allowed` is `TRUE` only for a `stack_direct` fit whose
target was made with `df_method = "barnard_rubin"` and a positive
`df_complete` (`pv_brr_target()` stops with an error without one). The labels
do not depend on the status of the fit: a fit with status `"warning"` keeps
them, and a fit with status `"blocked"` has an empty estimate table.
`per_pv` and `stack_psis` intervals are always descriptive, because their
variance within each plausible value comes from the model's posterior draws,
not from the replicate weights. The label follows the rule, not the value of
the degrees of freedom: the classic rule assumes that the complete-data
degrees of freedom are infinite [@barnard1999small, p. 948], and a target made
with the Barnard–Rubin rule keeps the label `coverage_barnard_rubin` even with
`df_complete = Inf`, which gives the classic degrees of freedom ([The
design-based target: BRR–Fay replicate weights and Rubin's
rules](m2-brr-fay-target.html#degrees-of-freedom) explains both rules). The
label records how the interval was built; it does not certify coverage ([Choosing
a method and reading its intervals](m5-methods-and-coverage.html#interval-rule)
gives the reason for the rule and its limits).

The example fit is a `stack_direct` fit whose target uses the classic degrees
of freedom, so every row is `descriptive_classic_rubin` with
`coverage_claim_allowed = FALSE`, and `print(fit)` adds the note "intervals
are descriptive rather than coverage-claimable". The method alone does not
decide the label; read it in the table. Report an interval as a confidence
interval with nominal coverage only when `coverage_claim_allowed` is `TRUE`,
and as a descriptive interval otherwise.

The calibration diagnostics are not evidence about coverage. The Cholesky
calibration correction (CCC) transforms the draws of the stacked fit so that
their mean and covariance equal the target's $\bar\beta$ and $T_{\text{MI}}$,
whatever the target is. Its diagnostics (see [the
diagnostics](#target-and-diagnostics)) show how far the stacked fit was from
the target before the calibration and whether the calibration was
numerically stable; they cannot show that the target's intervals have nominal
coverage. [Calibrating the fixed-effect draws
(CCC)](m4-ccc-calibration.html) describes the calibration, and [Choosing a
method and reading its intervals](m5-methods-and-coverage.html) explains the
intervals of each method.

## Fraction of missing information {#fmi}

For each coefficient $\ell$, `pv_brr_target()` computes

$$
\lambda_\ell = \frac{(1 + 1/M)\,B_{\ell\ell}}{T_{\text{MI},\ell\ell}},
\qquad
T_{\text{MI}} = \bar U + (1 + 1/M)\,B,
$$

where $\bar U$ is the BRR–Fay replicate covariance of the estimates averaged
over the plausible values, and $B$ is the covariance of the estimates between
plausible values. $\lambda_\ell$ is the share of the total variance of
coefficient $\ell$ that comes from the variation between plausible values;
@barnard1999small use this ratio as an approximation to the fraction of
missing information (FMI). pvstackr stores it in the target twice, as `fmi`
and as `lambda`, which are identical; the estimate table has no column for it.
Since the classic degrees of freedom are $(M - 1)/\lambda_\ell^2$, an FMI
close to 1 gives degrees of freedom close to their smallest value, $M - 1$.

The target also stores `riv`, the relative increase in variance
$(1 + 1/M)B_{\ell\ell}/\bar U_{\ell\ell}$, and the degrees of freedom:

```{r fmi}
tg <- get_target(fit)

data.frame(
  term = tg$fe_names,
  fmi  = round(as.numeric(tg$fmi), 3),
  riv  = round(as.numeric(tg$riv), 2),
  df   = round(as.numeric(tg$df),  2)
)
```

In the example the FMI is between 0.84 and 0.99 because the variance between
the two plausible values is much larger than the replicate-weight variance:
`riv` is about 5 for `b_x` and about 95 and 146 for the intercept and
`b_female`. Because `fmi` equals `riv / (1 + riv)`, a large `riv` gives an FMI
close to 1, and the degrees of freedom are then close to $M - 1 = 1$. The small
number of plausible values is not the reason: with the same $B$ and $\bar U$
and ten plausible values, the FMI would still be between 0.80 and 0.99.

PISA 2022 provides ten plausible values per domain [@oecd2024pisa]. The FMI of
an analysis of real data depends on its data and model, and this synthetic
example says nothing about it.

## The target, the draws and the diagnostics {#target-and-diagnostics}

### The target

`tg`, made in the previous chunk, holds the target of the fit. The estimate
table takes its standard errors, degrees of freedom and checksum from these
fields:

```{r target-fields}
diag(tg$T_MI)        # variance of each coefficient; se is its square root
round(tg$df, 3)      # degrees of freedom of each coefficient
tg$target_hash       # checksum of the target, also in the estimate table
```

The two comparisons below return `TRUE` because pvstackr copies `estimate`
from `beta_bar` and `se` from the square roots of the diagonal of `T_MI` when
`center = "target"`. The equality holds for every `stack_direct` fit that has
estimates; it shows how the table is made, not that two methods agree.

```{r target-identity}
all.equal(est$estimate, unname(tg$beta_bar))            # estimate = beta_bar
all.equal(est$se,       unname(sqrt(diag(tg$T_MI))))    # se = sqrt(diag(T_MI))
```

`?pv_brr_target` lists all fields of the target. Besides `beta_bar` (also
stored as `beta`), `T_MI`, `df` and `target_hash`, they include `U_bar` and
`B`, the two parts of $T_{\text{MI}} = \bar U + (1 + 1/M)B$; `fmi` and
`lambda`, which hold the same values, and `riv` (see [the fraction of missing
information](#fmi)); `M`, `R` and `fay_k`; and `fay_variance_multiplier`, the
factor $1/(R(1 - k)^2)$ of the BRR–Fay replicate covariance, where $k$ is the
Fay coefficient `fay_k`. In the example $R = 4$ and $k = 0.5$, so the factor
is 1. [The design-based target: BRR–Fay replicate weights and Rubin's
rules](m2-brr-fay-target.html) explains how the target is computed: the
BRR–Fay replicate covariance for each plausible value, Rubin's rules, and the
two rules for the degrees of freedom.

### The draws

A fit made with `return_draws = TRUE`, the default of `pv_control()`, keeps
the calibrated fixed-effect draws: a matrix with one row per draw of the
stacked fit and one column per fixed effect, named as in the estimate table.
They can be used for quantities derived from the coefficients: their mean and
covariance are the target's, so a linear combination computed from them has
the target's estimate and variance. They do not carry the target's degrees of
freedom, so their quantiles give roughly normal-reference intervals, narrower
than the $t$ intervals of the estimate table when the degrees of freedom are
small ([Calibrating the fixed-effect draws (CCC)](m4-ccc-calibration.html#map)).
For the example fit, which was saved without them, `get_draws()` returns
`NULL`, as shown above; it does the same for a blocked fit.

### The diagnostics

```{r get-diagnostics}
dg <- get_diagnostics(fit)

names(dg)
```

For a `stack_direct` fit the list has six elements:

- `preflight`: the record of the check, made before fitting, that the formula,
  the data and the target match;
- `sampler`: the sampler diagnostics of the stacked fit (the largest R-hat,
  the smallest bulk and tail effective sample sizes, the number of divergent
  transitions, and the numbers of chains and of draws per chain);
- `sampler_gate`: the result of pvstackr's check of those values;
- `stack_fit` and `stack_fit_warnings`: a record of the stacked fit and the
  notes it produced;
- `ccc`: the calibration diagnostics.

The thresholds of the checks are listed under "Status and checks" in
`?pvstackr_object_contracts`; a fit has status `"ok"` when none of them gave a
warning or a block. The sampler record of the example fit was typed in by its
fitting functions: 2 chains of 10 draws each, an R-hat of 1 and no divergent
transitions. The record reports 20 draws although the functions return 12:
pvstackr compares the reported numbers of chains and draws with the settings
in `pv_control()`, not with the draws, so the record passed pvstackr's checks.
The center-separation values in `ccc` are:

```{r ccc-peek}
dg$ccc[c("center_status", "delta_c_rel", "delta_c_max")]
```

`delta_c_max` is the largest, over the fixed effects, of the absolute
difference between the target estimate and the mean of the stacked draws
before calibration, divided by the target standard error:

$$
\max_\ell \frac{\lvert \bar\beta_\ell - \bar\beta^{\text{raw}}_\ell \rvert}
               {\sqrt{T_{\text{MI},\ell\ell}}},
$$

where $\bar\beta^{\text{raw}}_\ell$ is the mean of the stacked draws of
coefficient $\ell$ before calibration. By pvstackr's thresholds,
`center_status` is `"ok"` when `delta_c_max` is below 0.01, `"warning"` from
0.01 and `"blocked"` from 0.05. `delta_c_rel` is the root mean square (RMS) of
the same ratios; it is recorded but does not change the status. Both are
about $10^{-14}$ in the example because its draws were built around the
target. `ccc` also holds `kappa_A`, the condition number of the calibration
matrix, which has thresholds of its own. [Calibrating the fixed-effect draws
(CCC)](m4-ccc-calibration.html) explains these values.

## What to report {#reporting-checklist}

A checklist for reporting the fixed effects of a pvstackr fit:

- The method (`stack_direct`, `per_pv` or `stack_psis`); [Comparing the three
  fitting methods](a4-comparing-methods.html#choosing) gives pvstackr's
  recommendation.
- The status shown by `print(fit)` and, for `"warning"`, the reason codes it
  lists.
- `estimate` and `se` from `get_estimates()`; for `stack_direct` they are the
  target's ([The full analysis workflow](a2-the-workflow.html#center-target)
  explains why).
- `conf_low`, `conf_high` and `conf_level`, with the `interval_role` of each
  row.
- The name of each interval: "confidence interval" only where
  `coverage_claim_allowed` is `TRUE`, otherwise "descriptive interval".
- `df`, `df_method` and, with the Barnard–Rubin rule, `df_complete`.
- The fraction of missing information, `get_target(fit)$fmi` (`stack_direct`
  and `per_pv`).
- `target_hash` and the pvstackr version, `packageVersion("pvstackr")`; [Real
  PISA data: access, memory and
  reproducibility](a5-real-pisa-guidance.html#reproducibility-checklist) lists
  the other records to keep.

For the example fit a report would read: "The slope of `x` was 46.9 (standard
error 0.37; 95% descriptive interval 44.4 to 49.4, classic Rubin degrees of
freedom 1.40; `stack_direct`, status ok; fraction of missing information
0.84)." The data are synthetic.

## References
