---
title: "A3: Reading results — estimates, intervals, and what to report"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  How to read a fitted object and decide what is safe to report. This vignette
  tours the four stable accessors, walks the reportable estimate table, and
  explains the interval-metadata columns — `interval_role`,
  `coverage_claim_allowed`, `df_method`, and `df_complete` — that mark a row as
  coverage-claimable or descriptive. It covers the fraction of missing information
  and closes with a copy-paste reporting checklist.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{A3: Reading results — estimates, intervals, and what to report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

## 1. Setup — load a fit to read

Vignettes A1 and A2 produced a fitted object two ways — A1 loaded the cached
fixture and read it three ways; A2 built the pipeline that creates such a fit
(`pv_design()` → `pv_brr_target()` → `pv_fit(method = "stack_direct")`). This
article does neither. It assumes you already have a fit in hand and asks the next
question: *what is in it, and what may you safely report?*

We reuse the same bundled, synthetic `stack_direct` fit so everything runs
offline and deterministically. We only **read** it — nothing below re-estimates
anything:

```{r load-fixture}
fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit
```

::: {.alert .alert-warning}
This is the same tiny **synthetic** PISA-shaped fixture used across the Applied
track — one make-believe country, 12 students, $M = 2$ plausible values, $R = 4$
replicate weights. The coefficients below are **illustrative**, not real PISA
results, and must never be cited as such. For genuine PISA files see vignette A5.
:::

## 2. The four accessors

Everything you report comes from four stable accessors. Each takes the fit and
returns one well-defined object; together they cover the headline table, the
external variance answer, the calibrated draws, and the diagnostics.

**`get_estimates(fit)`** is the headline. It returns the reportable tidy table —
one row per fixed-effect term, with the estimate, standard error, confidence
interval, degrees of freedom, and the interval metadata that licenses (or
withholds) a coverage claim. This is the object you turn into a results table.
The full column tour is Section 3.

**`get_target(fit)`** returns the external Rubin / BRR–Fay target object (class
`pvstackr_brr_target`) that the fit was calibrated against. It carries the pooled
coefficients, the total covariance $T_{\text{MI}}$, the degrees of freedom, the
fraction of missing information, and a content hash. Under the reportable
convention (`center = "target"`), the headline estimate and SE in
`get_estimates()` *are* values read off this object — see Sections 3 and 6.

**`get_draws(fit)`** returns the calibrated fixed-effect posterior draws **when
they were retained** — an $S \times \text{(FE terms)}$ matrix — and `NULL`
otherwise. The bundled fit was produced with `return_draws = FALSE` to keep the
object small, so it returns `NULL`. That is expected and does not affect the
fixed-effect report, which is fully determined by the target:

```{r get-draws}
get_draws(fit)
```

**`get_diagnostics(fit)`** returns a named list of fit diagnostics —
`preflight`, `stack_fit`, `stack_fit_warnings`, and (for `stack_direct`) the
`ccc` calibration block with the center-separation agreement numbers. These tell
you whether the calibration behaved; you read them to *trust* the fit, not to
populate a results table. Section 6 takes a peek; the full treatment is in the
Method track, **M4**.

## 3. The estimate table tour

`get_estimates()` is the workhorse. Pull the whole table once, then look at it in
slices:

```{r get-estimates}
est <- get_estimates(fit)

names(est)
```

The table is deliberately wide because it carries both the numbers and their
*provenance*. Three things to notice before reading any value.

**The three terms.** Rows are `b_Intercept`, `b_x`, and `b_female` — the
fixed-effect block, with the `b_` prefix marking each as a fixed-effect
coefficient. There are no variance-component rows: v0.1 calibrates and reports
$\beta_{\text{FE}}$ only, a scope reminder carried over from A1/A2 and confirmed
by the `parameter_scope` column below.

**Dual snake_case / broom naming.** Several columns appear twice under two
spellings: the package-native snake_case (`se`, `conf_low`, `conf_high`) and the
`broom`-style dotted aliases (`std.error`, `conf.low`, `conf.high`). They hold
identical values — the duplication lets pvstackr tables drop into `broom`/`tidy`
pipelines unchanged while keeping a tidy snake_case surface for everything else.
Use whichever your downstream code expects.

Here is the compact "what is the estimate" slice most analyses start from:

```{r estimate-core}
est[, c("term", "estimate", "se", "df", "df_method")]
```

On this synthetic fixture the intercept sits near 458 score points, the slope on
`x` near 47 points per unit, and the `female` coefficient near 2 points with a
very wide interval (Section 4 explains the width). The `df` values hover near 1
because $M = 2$ — there is almost no between-imputation information to spend.

**Provenance is in the table, not in your memory.** Three columns record where
each row came from:

```{r estimate-provenance}
est[, c("term", "parameter_scope", "target_source", "target_hash")]
```

- `parameter_scope == "fixed_effect"` confirms every row is in the reportable
  fixed-effect block.
- `target_source == "external_brr_fay_rubin"` is the seal that the variance came
  from an **external, design-based** Rubin / BRR–Fay target — the property that
  governs coverage (Section 4).
- `target_hash` is a stable fingerprint of that target object. Record it: it lets
  you (or a reviewer) prove which target produced these numbers. The same hash
  appears on the target object itself (Section 6) and underpins the
  reproducibility checklist in A5.

**Where the numbers come from.** Because this fit uses the reportable convention
`center = "target"` (see A2 §6), the reported `estimate` and `se` are *exactly*
the external target's pooled coefficients $\bar\beta$ and
$\sqrt{\operatorname{diag}(T_{\text{MI}})}$. The single stacked, CCC-calibrated
fit is engineered to *report the external answer* — Section 6 shows the same
values living on the target object, and verifies they coincide.

## 4. Interval metadata — coverage-claimable vs descriptive

The columns that separate an honest report from an over-claim are the interval
metadata. Read them as a small block:

```{r interval-metadata}
est[, c("term", "interval_role", "df_method",
        "df_complete", "coverage_claim_allowed")]
```

What each column means:

- **`interval_role`** names the *kind* of interval. Here every row is
  `descriptive_classic_rubin`: a Rubin interval built on **classic** degrees of
  freedom, suitable for *describing* the spread of the estimate but not for a
  calibrated coverage claim.
- **`df_method`** is how the degrees of freedom were computed — `"classic"` (the
  plain Rubin formula) here, as opposed to the small-sample **Barnard–Rubin**
  adjustment that a coverage-claimable interval requires
  [@barnard1999small; @rubin1987multiple].
- **`df_complete`** is the complete-data (large-sample) degrees of freedom the
  Barnard–Rubin correction needs. It is `NA` for this fixture precisely because
  the classic path does not use it; a Barnard–Rubin target would supply it.
- **`coverage_claim_allowed`** is the single licensing flag. It is `FALSE` for
  every row here.

**The rule** — enforced by the metadata, not by trust:

> A row is `coverage_claim_allowed = TRUE` **only** when the fit is
> `stack_direct` *and* it was calibrated against an external **Barnard–Rubin**
> BRR–Fay target [@barnard1999small; @rubin1987multiple; @judkins1990fay]. The
> `per_pv` and `stack_psis` methods are **always** descriptive, regardless of
> their df. And a `stack_direct` fit that uses **classic** Rubin df — like this
> bundled fixture — is also descriptive, because classic df does not deliver the
> small-sample calibration the coverage claim rests on.

Two practical consequences:

- **Read the column, do not infer from the method name.** "It's `stack_direct`,
  so it must be coverage-claimable" is exactly the mistake the metadata exists to
  prevent. This fixture *is* `stack_direct` and is still descriptive. Let
  `coverage_claim_allowed` decide, every time.
- **Coverage comes from provenance, not from the calibration diagnostics.** It is
  tempting to read the near-perfect CCC center-separation numbers in Section 6
  (agreement to ~$10^{-14}$) as evidence of coverage. They are not. CCC is an
  affine moment match — it guarantees the calibrated draws' mean and covariance
  equal the target's, by construction — but that algebraic identity is not a
  coverage theorem. Coverage is conferred by the **provenance** of $T_{\text{MI}}$
  (external, design-based, Barnard–Rubin) and the supporting simulation evidence,
  not by how cleanly CCC ran. The precise statement is in **M4** (the CCC map)
  and **M5** (why only `stack_direct` is coverage-claimable).

The bundled fixture is therefore a faithful, deliberately *descriptive*
demonstration of the object surface — not a coverage-claimable result.

## 5. Fraction of missing information (FMI)

The **fraction of missing information** (FMI) for a coefficient is the share of
its total uncertainty that comes from imputation (the spread *between* plausible
values) rather than from sampling. For coefficient $k$,

$$
\gamma_k \;=\; \frac{\bigl(1 + \tfrac1M\bigr)\,B_{kk}}{T_{\text{MI},\,kk}},
$$

where $B$ is the between-imputation covariance and $T_{\text{MI}}$ the total
target covariance from the Rubin combining rules [@rubin1987multiple]. FMI is
what drives the degrees of freedom — high FMI means little information per
coefficient, hence small df and wide intervals.

FMI lives on the **target object**, not in the `get_estimates()` table — it is a
property of the variance target, so you read it from `get_target()`. Alongside it
the target reports the relative increase in variance (`riv`) and the degrees of
freedom that FMI drives:

```{r fmi}
tg <- get_target(fit)

data.frame(
  term = tg$fe_names,
  fmi  = round(as.numeric(tg$fmi), 3),
  riv  = round(as.numeric(tg$riv), 2),
  df   = round(as.numeric(tg$df),  2)
)
```

These FMI values are very high — roughly 0.84 to 0.99 — which looks alarming
until you remember why: with only $M = 2$ plausible values, the
$(1 + 1/M) = 1.5$ multiplier sits on a between-imputation covariance estimated
from a single pair of fits, so the between-share dominates and the degrees of
freedom collapse toward 1 (compare the `df` column in Section 3). This is the MI
machinery being *honest* about how little information two imputations carry — not
a defect.

::: {.alert .alert-warning}
Because $M = 2$ here, these FMI values illustrate the **mechanics**, not a
substantive finding. Real PISA reading uses $M = 10$ plausible values, where FMI
is far smaller and the small-sample df correction [@barnard1999small] matters in
its intended regime. Do not read this fixture's FMI as a statement about real
data.
:::

## 6. Reading the target and draws

The remaining accessors round out the picture: the target supplies the headline
numbers, draws are absent here, and the diagnostics tell you the calibration
behaved.

### 6.1 The target — `get_target()`

The target object carries the external variance answer in full. The three fields
you reach for most:

```{r target-fields}
diag(tg$T_MI)        # total target variance per coefficient
round(tg$df, 3)      # degrees of freedom per coefficient
tg$target_hash       # content fingerprint (matches the estimate table)
```

This is the *source* of the reportable SE: the standard errors in
`get_estimates()` are $\sqrt{\operatorname{diag}(T_{\text{MI}})}$, and the point
estimates are the pooled `beta_bar`. We can confirm the identity directly — under
`center = "target"`, the table and the target agree to machine precision:

```{r target-identity}
all.equal(est$estimate, unname(tg$beta_bar))            # estimates == pooled beta
all.equal(est$se,       unname(sqrt(diag(tg$T_MI))))    # SEs == sqrt(diag T_MI)
```

The target also carries `lambda`, the `fmi` of Section 5, the Fay variance
multiplier (`fay_variance_multiplier == 1` here), and the components
`U_bar` / `B` that feed $T_{\text{MI}} = \bar U + (1 + 1/M)B$. Their derivation —
Rubin pooling, the BRR–Fay sandwich, Barnard–Rubin df, and the design-variance
coverage result — is the subject of the Method track, **M2**.

### 6.2 The draws — `get_draws()`

As noted in Section 2, this fit retained no posterior draws, so the draw accessor
returns `NULL`:

```{r get-draws-again}
get_draws(fit)
```

Had the fit been run with `return_draws = TRUE`, this would be an
$S \times \text{(FE terms)}$ matrix of CCC-calibrated fixed-effect draws — useful
for derived quantities or custom intervals. Its absence does not weaken the
fixed-effect report, which is fully specified by the target.

### 6.3 The diagnostics — `get_diagnostics()`

```{r get-diagnostics}
dg <- get_diagnostics(fit)

names(dg)
```

The list holds `preflight` (the resolved formula, term names, and target
fingerprint checked before fitting), `stack_fit` and `stack_fit_warnings` (the
stacked-fit record and any messages it raised), and `ccc`, the calibration block
for `stack_direct`. A quick peek at the center-separation agreement inside `ccc`:

```{r ccc-peek}
dg$ccc[c("center_status", "delta_c_rel", "delta_c_max")]
```

`center_status == "ok"` means the calibration's center matched the target's
within tolerance; `delta_c_rel` and `delta_c_max` are the relative
center-separation summarised two ways — a root-mean-square diagnostic and a
max-over-terms reportable gate — both effectively zero (~$10^{-14}$) here.

A reminder from Section 4: these tiny numbers confirm CCC *ran cleanly*; they do
**not** confer coverage. The center-separation diagnostic, its green/yellow/red
tiers, the conditioning diagnostic $\kappa_A$, and the RMS-vs-max distinction are
defined in **M4**.

## 7. A reporting checklist

A copy-paste checklist for reporting a pvstackr fixed-effect result. Walk it once
per analysis; every field is read directly off the objects above, never
remembered.

- **Method.** State it (`stack_direct` for a reportable fit). `per_pv` and
  `stack_psis` are reference/cross-check only — see A4.
- **Centering.** Confirm `center = "target"` (the reportable convention; A2 §6).
  A `"posterior"`-centered fit is diagnostic only and must not be reported.
- **Estimates and SEs.** Take `estimate` and `se` from `get_estimates()`. Under
  `center = "target"` these are the external target's pooled coefficients and
  $\sqrt{\operatorname{diag}(T_{\text{MI}})}$.
- **Intervals — with their role.** Report `conf_low` / `conf_high` *together with*
  the `interval_role` for each row. Never strip the role off the number.
- **Degrees of freedom.** Report `df` and `df_method` (and `df_complete` when the
  Barnard–Rubin path supplies it), so the small-$M$ behaviour is visible.
- **Fraction of missing information.** Report `get_target(fit)$fmi` — high FMI
  explains small df and wide intervals.
- **The coverage flag.** Report intervals as **coverage-claimable only when
  `coverage_claim_allowed == TRUE`**. When it is `FALSE` (as for this fixture),
  label them **descriptive** explicitly.
- **Provenance.** Record `target_source`, `target_hash`, and the package version
  so the result is reproducible and auditable. The full reproducibility checklist
  is in A5.

For this synthetic fixture, that report reads: a `stack_direct`, `center =
"target"` fit; estimates ~458 / 47 / 2 with SEs from $T_{\text{MI}}$; intervals
`descriptive_classic_rubin` with `df_method == "classic"` and df ~1; FMI
~0.84–0.99; and — because `coverage_claim_allowed == FALSE` throughout — the
intervals reported as **descriptive, not coverage-claimable**.

## 8. Where to next

You can now read a fit and decide what is safe to report. From here:

- **A4 · Comparing methods** — `pv_compare_methods()` across `stack_direct`,
  `per_pv`, and `stack_psis`; how to read agreement diagnostics; and the two
  cautions (design variance is not optional; a small Pareto-$\hat k$ does not
  guarantee correct variance).
- **A5 · Real PISA data guidance** — connecting to genuine PISA files: licensing
  and non-affiliation, the design declaration, memory and runtime, and the full
  reproducibility checklist (where `target_hash` and friends earn their keep).

For the mathematics behind the columns read here — the external target, Rubin /
BRR–Fay combining, Barnard–Rubin df, and FMI in **M2**, and the precise statement
of why only `stack_direct` is coverage-claimable in **M5** — start the Method
track at **M1 · Foundations and notation**.

Bug reports and feature requests:
<https://github.com/joonho112/pvstackr/issues>.

### Session info

```{r session-info}
sessionInfo()
```

## References
