---
title: "Calibrating the fixed-effect draws (CCC)"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  In a `stack_direct` fit, the Cholesky calibration correction (CCC)
  transforms the fixed-effect draws of one Bayesian fit to the stacked
  plausible-value data so that their mean and covariance equal the estimates
  and the covariance matrix of the design-based target. This article gives
  the map, its exact moment properties and their source, the algorithm that
  pvstackr uses and when it stops with an error, and the diagnostics
  `delta_c_max`, `delta_c_rel` and `kappa_A` with pvstackr's thresholds. A
  check on simulated draws confirms the moment properties; neither the check
  nor the calibration says anything about the coverage of intervals.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Calibrating the fixed-effect draws (CCC)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

A `stack_direct` fit rests on two calculations. The target, computed by
`pv_brr_target()`, fits a survey-weighted linear regression by weighted least
squares to each of the $M$ plausible values, estimates the covariance of each
estimate from the BRR–Fay replicate weights, and combines the results with
Rubin's rules into the estimate $\bar\beta$ of the $p$ fixed effects and its
total covariance matrix $T_{\text{MI}} = \bar U + (1 + 1/M)B$. Here $\bar U$ is
the average of the $M$ replicate covariances and $B$ the covariance between
the $M$ estimates ([The design-based target: BRR–Fay replicate weights and
Rubin's rules](m2-brr-fay-target.html#rubin)). The stacked fit is one Bayesian
fit of the same regression to the stacked data: $M$ copies of the $N$ rows,
one for each plausible value, in which row $i$ has the weight $\tilde w_i/M$,
where $\tilde w_i = W_i/\overline W$ is the final weight $W_i$ divided by the
mean final weight $\overline W$.

The Cholesky calibration correction (CCC) connects the two. It transforms the
fixed-effect draws of the stacked fit so that their mean is $\bar\beta$ and
their covariance is $T_{\text{MI}}$, and before it does so it records how far
the stacked fit was from the target. The code below reads the example fit that
ships with pvstackr and its target:

```{r common-load}
fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit
tg <- get_target(fit)   # the target of the fit, made by pv_brr_target()

is.null(get_draws(fit))
```

The example fit and its synthetic data are described in [Getting started with
pvstackr](a1-getting-started.html#example-fit). The fit was saved without its
calibrated draws, so [the check of the map](#checks) uses simulated draws, and
the diagnostics that the calibration recorded are read from the fit.

## Why the stacked draws are calibrated {#why-calibrate}

The stacked fit is built to agree with the target in its point estimate. Its
log likelihood is the average of the $M$ survey-weighted log-likelihoods, one
for each plausible value, and the stacked fixed-effect point identity, Theorem
4.1 of the companion preprint [@lee2026onemcmc], gives conditions under which
the value of $\beta$ that maximizes this average equals $\bar\beta$. In
pvstackr's model the weighted least-squares estimate from the stacked data
equals $\bar\beta$ exactly, so with a flat prior on the fixed effects the mean
of the draws differs from $\bar\beta$ only by Monte Carlo error ([One stacked
fit and the Rubin point estimate](m3-stacked-bridge.html#point-identity)
explains the identity and its conditions).

The identity concerns the point estimate only and gives no variance. The
between-imputation covariance $B$ is the covariance of estimates from separate
analyses of the $M$ plausible values and exists only across such analyses, so
the draws of one stacked fit do not contain it. The preprint notes that the
covariance of the stacked fixed-effect draws approximates a one-sample
quantity that corresponds to the within-imputation part $\bar U$ only, and so
lacks the term $(1 + 1/M)B$ of $T_{\text{MI}}$ [@lee2026onemcmc, pp. 11–12,
40]. In pvstackr the target's $\bar U$ comes from the replicate weights,
whereas the spread of the stacked draws comes from the model: the likelihood
of the stacked fit carries the survey weights, but not the clustering of the
sample or the replicate weights.

pvstackr therefore takes the numbers it reports from the target ([The full
analysis workflow](a2-the-workflow.html#center-target)), and CCC transforms
the draws to agree with these numbers. The calibrated draws, returned by
`get_draws(fit)`, have the mean $\bar\beta$ and the covariance $T_{\text{MI}}$,
so a linear combination of the coefficients computed from them has the
target's estimate and variance. They
do not carry the target's degrees of freedom: quantiles of the draws give
roughly a normal-reference interval, which is narrower than the $t$ interval
of the estimate table when the degrees of freedom are small (see [the
calibration map](#map)). Before the draws are moved, the calibration compares the
stacked fit with the target (see [the diagnostics](#diagnostics)).

## The calibration map {#map}

Let $\beta_1, \dots, \beta_S$ be the $S$ draws of the $p$ fixed effects from the
stacked fit, with mean and covariance

$$
\bar\beta^{\text{raw}} = \frac{1}{S}\sum_{s=1}^{S} \beta_s, \qquad
\Sigma_{\text{raw}} = \frac{1}{S-1}\sum_{s=1}^{S}
  \big(\beta_s - \bar\beta^{\text{raw}}\big)\big(\beta_s - \bar\beta^{\text{raw}}\big)^\top ,
$$

and let $L_{\text{raw}}$ and $L_{\text{tgt}}$ be the lower-triangular Cholesky
factors of $\Sigma_{\text{raw}}$ and $T_{\text{MI}}$, so that
$L_{\text{raw}} L_{\text{raw}}^\top = \Sigma_{\text{raw}}$ and
$L_{\text{tgt}} L_{\text{tgt}}^\top = T_{\text{MI}}$. CCC maps each draw to

$$
\beta^{\text{cal}}_s = \bar\beta + A\big(\beta_s - \bar\beta^{\text{raw}}\big),
\qquad A = L_{\text{tgt}} L_{\text{raw}}^{-1} .
$$

The preprint gives this map as its Eq. (5) [@lee2026onemcmc, p. 12]; in
pvstackr, $\bar\beta$ and $T_{\text{MI}}$ are those of the target from
`pv_brr_target()`. The map centers each draw at the mean of the stacked draws,
multiplies it by the calibration matrix $A$, and adds the target estimate
$\bar\beta$. Multiplying the centered draws by $L_{\text{raw}}^{-1}$ gives
them the identity matrix as covariance, and multiplying them by
$L_{\text{tgt}}$ then gives them the covariance $T_{\text{MI}}$.

The calibrated draws have three properties, which make up exact moment
calibration, Proposition B.1 of the preprint [@lee2026onemcmc, p. 37]:

- their mean is $\bar\beta$;
- their covariance, computed like $\Sigma_{\text{raw}}$, is $T_{\text{MI}}$;
- the draws of parameters other than the fixed effects are unchanged.

The first holds because the centered draws have mean zero and $A$ maps zero to
zero. The second follows from the way a covariance matrix changes under a
linear map:

$$
A\,\Sigma_{\text{raw}}\,A^\top
= L_{\text{tgt}} L_{\text{raw}}^{-1}
  \big(L_{\text{raw}} L_{\text{raw}}^\top\big)
  L_{\text{raw}}^{-\top} L_{\text{tgt}}^\top
= L_{\text{tgt}} L_{\text{tgt}}^\top
= T_{\text{MI}} .
$$

The third holds because the map acts on the fixed-effect columns only. In
pvstackr's model the only other parameter of the stacked fit is the residual
standard deviation $\sigma$, whose draws pass through unchanged; pvstackr
neither calibrates nor reports it, and `get_draws(fit)` returns the
calibrated fixed-effect draws only.

These properties are algebraic. They hold for any draws whose covariance
matrix is positive definite, apart from the rounding error of floating-point
arithmetic, and they do not depend on how well the stacked fit was sampled:
because the map uses the mean and covariance of the draws at hand, Monte Carlo
error changes $A$ but not the mean and covariance of the calibrated draws
[@lee2026onemcmc, p. 37].

The preprint describes CCC as the scale-and-shape correction that
@williams2021uncertainty developed for pseudo-posteriors under complex
sampling, specialized to the fixed-effect block of the model and to a
covariance target from the combining rules [@lee2026onemcmc, pp. 7, 13].

CCC sets the mean and the covariance of the draws and nothing else. The
calibrated draws are an affine transformation of the stacked draws, so they
are normal only when the stacked draws are, and their other features follow
from those of the stacked draws through the same transformation. The interval
of a `stack_direct` fit does not use the draws: it is the estimate plus or
minus a $t$ quantile times the standard error, all taken from the target
([Reading and reporting the
results](a3-reading-results.html#interval-columns)). Quantiles of the
calibrated draws therefore do not reproduce this interval. In the example,
whose classic degrees of freedom are between 1.0 and 1.4, the 95% $t$
intervals are three to six times as wide as normal-reference intervals with
the same standard errors; with $M = 10$ the classic degrees of freedom are at
least 9, so the $t$ quantile is at most 2.26 against 1.96; Barnard–Rubin
degrees of freedom can be smaller, and the gap then larger.

## The center of the calibrated draws

The map above centers the calibrated draws at the target estimate $\bar\beta$.
This is the setting `center = "target"` of `pv_control()`, the default, and a
`stack_direct` fit requires it:

```{r center-control}
fit$control$center   # "target": the calibrated draws are centered at beta_bar
```

With this setting the calibrated draws have the mean and covariance of the
target, whose numbers the estimate table copies ([The full analysis
workflow](a2-the-workflow.html#center-target) explains why pvstackr requires
this setting), so the table and the draws agree. `pv_control()` also accepts
`center = "posterior"`, which would keep the mean of the stacked draws,
$\beta^{\text{cal}}_s = \bar\beta^{\text{raw}} + A(\beta_s - \bar\beta^{\text{raw}})$,
and calibrate only the covariance; no fitting function uses it ([The full
analysis workflow](a2-the-workflow.html#center-target)).

## The algorithm {#algorithm}

CCC is deterministic: pvstackr computes it with plain Cholesky factorizations,
a triangular solve and matrix products, without random numbers, iterations or
tuning, so the same draws and target always give the same calibrated draws. A
fit that pvstackr's check of the sampler diagnostics blocks is not calibrated.
For the other fits, pvstackr proceeds as follows:

1. It takes the fixed-effect columns of the draws of the stacked fit and
   computes their mean $\bar\beta^{\text{raw}}$ and covariance
   $\Sigma_{\text{raw}}$ with `colMeans()` and `cov()`.
2. It checks the target: $T_{\text{MI}}$ must be symmetric and positive
   definite, its rows and columns must name the same coefficients as the
   draws, in the same order, and the target must come from `pv_brr_target()`.
   It also stops with an error when $T_{\text{MI}}$ equals $\Sigma_{\text{raw}}$
   or is proportional to it.
3. It factors $\Sigma_{\text{raw}}$ and $T_{\text{MI}}$ with `chol()` from base
   R, without pivoting, after making each matrix exactly symmetric as
   $(\Sigma + \Sigma^\top)/2$. `chol()` returns the upper-triangular factor $U$
   with $U^\top U = \Sigma$, and pvstackr uses its transpose, $L = U^\top$.
4. It computes $L_{\text{raw}}^{-1}$ by forward substitution, with
   `forwardsolve()` applied to the identity matrix, and forms
   $A = L_{\text{tgt}} L_{\text{raw}}^{-1}$, a lower-triangular $p \times p$
   matrix.
5. It centers every draw at $\bar\beta^{\text{raw}}$, multiplies it by $A$ and
   adds $\bar\beta$, and records [the diagnostics](#diagnostics).

A positive definite matrix has exactly one lower-triangular Cholesky factor
with a positive diagonal, so $A$ is well defined. Other matrices $G$ with
$G\,\Sigma_{\text{raw}}\,G^\top = T_{\text{MI}}$ exist, for example matrices
built from symmetric square roots; used in place of $A$, they give the
calibrated draws the same mean and covariance but move each draw differently.
The preprint's algorithm, like pvstackr, uses the lower-triangular factors of
both matrices [@lee2026onemcmc, p. 37].

Both factorizations need positive definite matrices, and pvstackr applies no
automatic Tikhonov or near-PD repair: it adds no ridge (a constant on the
diagonal) and does not replace a matrix by a nearby positive definite one.
When $\Sigma_{\text{raw}}$ or $T_{\text{MI}}$ is not positive definite,
pvstackr stops with an error and returns no fit. For $\Sigma_{\text{raw}}$ the
message is "pvstackr: `Sigma_raw` must be positive definite for CCC
calibration."; `Sigma_raw` and `Sigma_target` are the names of
$\Sigma_{\text{raw}}$ and $T_{\text{MI}}$ in the calibration record of a fit
("Calibration record" in `?pvstackr_object_contracts`). The calibration in the
preprint also halts with an error when its preconditions, positive definite
matrices among them, fail; its implementation keeps a nearest
positive-semidefinite repair as an explicitly logged fallback, which pvstackr
does not have [@lee2026onemcmc, p. 38].

$\Sigma_{\text{raw}}$ is not positive definite when there are fewer than
$p + 1$ draws, or when the draws of one fixed effect are an exact linear
combination of the draws of others, as for a fixed effect that the model does
not identify. The remedy is to change the stacked fit, for example with more
draws or with fewer or less collinear fixed effects. A target covariance that
is not positive definite already stops `pv_brr_target()` with an error when
the target is made, and `pv_control(allow_target_nearpd = TRUE)`, which would
allow a repair of the target, stops with an error as well.

A covariance matrix that is positive definite but close to singular passes
the factorization, and so, because of rounding, can a matrix that is singular
in exact arithmetic. When one of the two matrices is much closer to singular
than the other, the condition number $\kappa_A$ of $A$, stored as `kappa_A`,
is large and pvstackr's thresholds apply (see [the
diagnostics](#diagnostics)); when both are close to singular in the same
direction, $\kappa_A$ can stay small.

## Diagnostics {#diagnostics}

The calibration records its diagnostics in `get_diagnostics(fit)$ccc`. Two of
them can give the fit the status `"warning"` or `"blocked"`: the center
separation and the condition number of $A$. They are combined with pvstackr's
other checks, and the most severe result sets the status ("Status and checks"
in `?pvstackr_object_contracts` lists all checks). Their thresholds are
pvstackr's rules; the preprint sets no thresholds for these quantities. The
calibration also records `a_matrix_fro_rel`, which measures how far $A$ is
from the identity matrix, and `rho1`, `rho2` and `empirical_fro_rel`, which
measure how far the covariance of the calibrated draws is from $T_{\text{MI}}$
and are zero apart from rounding; these values do not change the status.

### Center separation

The center separation compares the mean of the stacked draws before
calibration with the target estimate, coefficient by coefficient, in units of
the target standard error:

$$
\Delta_c = \max_{\ell = 1, \dots, p}
  \frac{\lvert \bar\beta_\ell - \bar\beta^{\text{raw}}_\ell \rvert}
       {\sqrt{T_{\text{MI},\ell\ell}}} .
$$

pvstackr stores $\Delta_c$ as `delta_c_max` and the ratio for each coefficient
as `delta_c_by_term`; `delta_c_rel` is the root mean square (RMS) of the same
ratios. Measured in standard errors, the ratios can be compared across
coefficients on different scales, such as an intercept in score points and the
coefficient of a 0/1 indicator.

With a flat prior on the fixed effects, which the bundled engine uses when
`prior = NULL`, the mean of the stacked draws differs from $\bar\beta$ only by
Monte Carlo error, as noted [above](#why-calibrate). It differs further when
the prior is not flat, or when a fitting function of your own does not use the
stacked weights. In Appendix A.4, the preprint treats this agreement in its
own setting as an empirical quantity to be measured and reported, not as a
consequence of the point identity [@lee2026onemcmc, p. 36].

Monte Carlo error alone is not small on this scale. The Monte Carlo standard
error of the mean of the draws of coefficient $\ell$ is their standard
deviation divided by $\sqrt{\mathrm{ESS}_\ell}$, where $\mathrm{ESS}_\ell$ is
their effective sample size. In target standard errors it is therefore the
ratio of that standard deviation to $\sqrt{T_{\text{MI},\ell\ell}}$, divided by
$\sqrt{\mathrm{ESS}_\ell}$. With a ratio of 0.5 and 4,000 effective draws it is
about 0.008, close to the warning threshold of 0.01 below. A fit that differs
from the target only by Monte Carlo error can therefore get the status
`"warning"`, and with few effective draws even `"blocked"`; more draws make
this less likely.

pvstackr sets `center_status` from `delta_c_max`. Below 0.01 it is `"ok"`.
From 0.01 the fit gets the status `"warning"`, with the reason code
`center_separation_yellow`; its estimate table is unchanged, since the table
copies the target. From 0.05 the fit is `"blocked"`, with
`center_separation_red`, and its estimate table is empty; its
`get_diagnostics(fit)$ccc` then holds single values grouped as `center`,
`conditioning`, `residual` and `prior`, for example
`get_diagnostics(fit)$ccc$center$delta_c_max`. `delta_c_rel`, the RMS of the
ratios, is recorded and printed in the warning message but does not change
the status.

### Condition number of the calibration matrix

$\kappa_A$, stored as `kappa_A`, is the condition number of $A$: the ratio of
its largest to its smallest singular value. It is at least 1 and does not
change when $A$ is multiplied by a nonzero constant. A large $\kappa_A$ means
that $A$ stretches the centered draws far more in some directions than in
others, as when $\Sigma_{\text{raw}}$ is close to singular, so that small
errors in the stacked draws can be greatly magnified. From
$\kappa_A = 10^6$ the fit gets the status `"warning"`, with the reason code
`ccc_conditioning_yellow`, and from $10^8$ it is `"blocked"`, with
`ccc_conditioning_red`.

## What the calibration does not show

CCC gives the draws the mean and covariance of the target it receives,
whatever that target is. It cannot show that the target is right, and it does
not correct a target that is wrong. The preprint calls the correction a moment
calibration that guarantees agreement with the supplied target, and in its
Appendix B.3 it treats the coverage of intervals built from the calibrated
draws as an empirical question, which its simulation addresses for targets
from model-based fits only [@lee2026onemcmc, pp. 13, 38]. The diagnostics
compare the stacked fit with the target and describe the matrix $A$; they say
nothing about coverage either.

The interval of a `stack_direct` fit is computed from the estimate, the
standard error and the degrees of freedom of the target. pvstackr's reporting
rule labels it coverage-claimable only when the target uses the Barnard–Rubin
degrees of freedom; the label records how the interval was built and does not
certify its coverage. [Reading and reporting the
results](a3-reading-results.html#interval-columns) gives the rule, [Choosing a
method and reading its intervals](m5-methods-and-coverage.html#interval-rule)
gives its reason, and [The design-based target: BRR–Fay replicate weights and
Rubin's rules](m2-brr-fay-target.html) shows how $\bar\beta$ and
$T_{\text{MI}}$ are computed.

## Checks {#checks}

The first check applies the map to simulated draws; it is an arithmetic check
of the two moment properties. The second reads the diagnostics recorded in the
example fit, which hold by construction.

### The moment properties on simulated draws

The example fit has no draws, so the chunk below simulates them: $S = 4000$
draws of $p = 3$ coefficients from a normal distribution with a mean and a
covariance chosen for the check. It sets a target mean and covariance in place
of $\bar\beta$ and $T_{\text{MI}}$ and applies the map as pvstackr computes it,
except that it inverts $L_{\text{raw}}$ with `solve()` where pvstackr uses
`forwardsolve()`, which gives the same matrix up to rounding.

```{r ccc-invariant}
set.seed(5104)                      # fixed seed, so the simulated draws are the same in every run
p <- 3L; S <- 4000L

## simulated draws of p = 3 coefficients with a chosen mean and covariance
Sigma_raw_true <- crossprod(matrix(rnorm(p * p), p, p)) + diag(p)
raw <- sweep(matrix(rnorm(S * p), S, p) %*% chol(Sigma_raw_true),
             2L, c(10, -3, 0.5), `+`)

## target mean and covariance for the check, in place of beta_bar and T_MI
c_tgt     <- c(458, 47, 2)
Sigma_tgt <- matrix(c(4.0, 0.5, 0.0,
                      0.5, 2.0, 0.3,
                      0.0, 0.3, 1.5), 3, 3, byrow = TRUE)

## the map: beta_cal = c_tgt + A (beta - colMeans(raw)), with A = L_tgt L_raw^{-1}
L_tgt <- t(chol(Sigma_tgt))         # lower Cholesky factor of the target covariance
L_raw <- t(chol(cov(raw)))          # lower Cholesky factor of the covariance of the draws
A     <- L_tgt %*% solve(L_raw)     # the calibration matrix A

cal <- sweep(raw, 2L, colMeans(raw), `-`) %*% t(A)   # center the draws, then multiply by A
cal <- sweep(cal, 2L, c_tgt, `+`)                    # add the target mean

all.equal(unname(colMeans(cal)), c_tgt)      # TRUE (largest difference about 1e-13)
all.equal(unname(cov(cal)),      Sigma_tgt)  # TRUE (largest difference about 1e-14)
```

Both comparisons return `TRUE`. `all.equal()` accepts a relative difference
below about $1.5 \times 10^{-8}$; the largest absolute differences in this run
are about $10^{-13}$ for the mean, whose entries are as large as 458, and
$10^{-14}$ for the covariance, the size of rounding errors. The agreement is
exact apart from rounding because the map is built from the mean and
covariance of the same draws that the check summarizes, as in pvstackr. The
simulated covariance is positive definite and well conditioned by
construction (the term `+ diag(p)` and 4000 draws for three coefficients), so
the check reaches neither the error for a matrix that is not positive definite
nor the thresholds of `kappa_A`.

### The diagnostics of the example fit

The example fit recorded its calibration diagnostics when it was made, and
they can be read without the draws:

```{r ccc-diagnostics}
dg <- get_diagnostics(fit)

names(dg)
dg$ccc[c("center_status", "delta_c_rel", "delta_c_max")]
```

`center_status` is `"ok"`: `delta_c_max` is about $4.4 \times 10^{-14}$ and
`delta_c_rel` about $2.5 \times 10^{-14}$, far below 0.01. These values hold by
construction. The functions that made the example fit returned draws built
around the target, whose means equal $\bar\beta$ apart from rounding ([Getting
started with pvstackr](a1-getting-started.html#example-fit)), so the values
show what the diagnostics look like, not how close the draws of a sampler come
to the target. The list also holds `delta_c_by_term` and `kappa_A` (see [the
diagnostics](#diagnostics)).

The numbers above were computed in this session:

```{r session-info}
sessionInfo()
```

## References
