---
title: "Choosing a method and reading its intervals"
author: "JoonHo Lee"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
abstract: >
  pvstackr has three fitting methods, `stack_direct`, `per_pv` and
  `stack_psis`. All three combine an estimate and a covariance matrix for each
  plausible value with Rubin's rules; they differ in how they compute these.
  This article gives the formulas of the three methods, the importance weights
  and Pareto k-hat values that `stack_psis` takes from another program
  together with pvstackr's check of them, and the reason why pvstackr's
  reporting rule labels intervals coverage-claimable only for `stack_direct`
  fits whose target uses Barnard–Rubin degrees of freedom.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Choosing a method and reading its intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", out.width = "85%")
```

```{r library}
library(pvstackr)
```

pvstackr estimates the fixed effects of a regression on plausible-value data
with one of three methods, set by the argument `method` of `pv_fit()`:
`stack_direct` (the default), `per_pv` and `stack_psis`. [Comparing the three
fitting methods](a4-comparing-methods.html) compares example fits of the three methods
and gives pvstackr's recommendation for choosing among them.

The code in this article reads the example fit that ships with pvstackr, a
`stack_direct` fit, and fits no model:

```{r common-load}
fit <- readRDS(
  system.file("extdata", "examples", "pisa_tiny_stack_direct.rds",
              package = "pvstackr")
)$fit
tg <- get_target(fit)   # the target of the fit, made by pv_brr_target()
```

The example fit and its synthetic data are described in [Getting started with
pvstackr](a1-getting-started.html#example-fit).

## The three methods {#three-methods}

Write $\beta$ for the $p$ fixed-effect coefficients and $m = 1, \dots, M$ for
the plausible values. Each method computes, for each plausible value, an
estimate $\hat\beta_m$ of $\beta$ and a covariance matrix $\hat U_m$, and
combines them with Rubin's rules [@rubin1987multiple]:

$$
\bar\beta = \frac{1}{M}\sum_{m=1}^{M} \hat\beta_m, \qquad
\bar U = \frac{1}{M}\sum_{m=1}^{M} \hat U_m, \qquad
B = \frac{1}{M-1}\sum_{m=1}^{M}
  (\hat\beta_m - \bar\beta)(\hat\beta_m - \bar\beta)^\top,
$$

$$
T_{\text{MI}} = \bar U + \Big(1 + \frac{1}{M}\Big) B .
$$

For coefficient $\ell$, the estimate table reports the estimate
$\bar\beta_\ell$, the standard error $\sqrt{T_{\text{MI},\ell\ell}}$, and an
interval: the estimate plus or minus a $t$ quantile times the standard error,
with the classic or the Barnard–Rubin degrees of freedom and the level
`conf_level` of `pv_control()` (0.95 by default). The standard errors of every
method therefore include the variation between plausible values, $B$. [The
design-based target: BRR–Fay replicate weights and Rubin's
rules](m2-brr-fay-target.html#rubin) derives these formulas and the degrees of
freedom. The methods differ in $\hat\beta_m$ and $\hat U_m$:

| Method | $\hat\beta_m$ | $\hat U_m$ | Bayesian fits | Intervals |
|:--|:--|:--|:--|:--|
| `stack_direct` (default) | weighted least-squares estimate with the final weights | BRR–Fay replicate covariance | 1 | coverage-claimable only with an external Barnard–Rubin target, otherwise descriptive |
| `per_pv` | posterior mean of the fit to plausible value $m$ | posterior covariance of that fit | $M$ | always descriptive |
| `stack_psis` | weighted mean of the draws of one stacked fit | weighted covariance of the same draws | 1 | always descriptive |

A coverage-claimable interval is one that pvstackr's reporting rule lets you
read as a confidence interval with nominal coverage; a descriptive interval
only describes the uncertainty of the estimate. The source of $\hat U_m$ is the
reason why only `stack_direct` intervals can be coverage-claimable ([the
reason for the rule](#interval-rule)).

The companion preprint [@lee2026onemcmc] calls the corresponding workflows the
calibrated stack, the per-PV workflow and the reweighted-stack variant, and its
Appendix G.1 matches them to `stack_direct`, `per_pv` and `stack_psis`. The
correspondence is one of role. The preprint's calibrated stack uses by default
a target from model-based fits of each plausible value [@lee2026onemcmc,
sec. 4.4], whereas `pv_brr_target()` computes only the replicate-weight
target, and the preprint's analysis model has a school random intercept, which
pvstackr 0.2.x does not fit.

### `stack_direct`

`stack_direct` (`pv_fit_direct()`) takes $\hat\beta_m$ and $\hat U_m$ from the
target that `pv_brr_target()` computes before the fit. $\hat\beta_m$ is the
weighted least-squares estimate for plausible value $m$ with the final survey
weights, and

$$
\hat U_m = \frac{1}{R(1-k)^2} \sum_{r=1}^{R}
  \big(\hat\beta_m^{(r)} - \hat\beta_m\big)\big(\hat\beta_m^{(r)} - \hat\beta_m\big)^\top
$$

is its BRR–Fay replicate covariance, where $\hat\beta_m^{(r)}$ is the same
estimate with replicate weight $r = 1, \dots, R$ and $k$ is the Fay
coefficient ([The design-based target: BRR–Fay replicate weights and Rubin's
rules](m2-brr-fay-target.html#replicate-covariance)). The target holds
$\bar\beta$, $\bar U$, $B$ and $T_{\text{MI}}$, and the estimate table of the
fit copies the estimates, the standard errors and the degrees of freedom from
it.

The one Bayesian fit of `stack_direct` is a fit to the stacked data, one copy
of the data for each plausible value. Its fixed-effect draws
$\beta_1, \dots, \beta_S$ do not give the reported numbers: the covariance of
the draws does not contain $B$, and it comes from the model, not from the
replicate weights ([Calibrating the fixed-effect draws
(CCC)](m4-ccc-calibration.html#why-calibrate)). The Cholesky calibration
correction (CCC) maps each draw to
$\beta^{\text{cal}}_s = \bar\beta + A(\beta_s - \bar\beta^{\text{raw}})$,
where $\bar\beta^{\text{raw}}$ is the mean of the draws and $A$ a calibration
matrix, so that the calibrated draws have the mean $\bar\beta$ and the
covariance $T_{\text{MI}}$ of the target. The stacked fixed-effect point
identity, Theorem 4.1 of the preprint [@lee2026onemcmc], gives conditions
under which the fixed-effect estimate of the stacked fit equals $\bar\beta$
([One stacked fit and the Rubin point
estimate](m3-stacked-bridge.html#point-identity)).

### `per_pv`

`per_pv` (`pv_fit_reference()`) combines $M$ Bayesian fits, one for each
plausible value. You supply their draws, either computed elsewhere
(`per_pv_draws`) or returned by fitting functions of your own, which pvstackr
calls once for each plausible value. With
$\beta_1^{(m)}, \dots, \beta_{S_m}^{(m)}$ the fixed-effect draws of the fit to
plausible value $m$,

$$
\hat\beta_m = \frac{1}{S_m}\sum_{s=1}^{S_m} \beta_s^{(m)}, \qquad
\hat U_m = \frac{1}{S_m - 1}\sum_{s=1}^{S_m}
  \big(\beta_s^{(m)} - \hat\beta_m\big)\big(\beta_s^{(m)} - \hat\beta_m\big)^\top ,
$$

the posterior mean and the posterior covariance of the draws. The model is the
one your fits use: pvstackr does not add the survey weights to its formula and
does not check the formula for random-effect terms ([Plausible values, survey
weights and notation](m1-foundations-and-notation.html#model)).

### `stack_psis`

`stack_psis` (`pv_fit_stack_psis()`) uses the fixed-effect draws
$\beta_1, \dots, \beta_S$ of one Bayesian fit to the stacked data and, for each
plausible value, importance weights $\omega_1^{(m)}, \dots, \omega_S^{(m)}$
that reweight these draws toward the posterior for plausible value $m$.
pvstackr divides the weights by their sum,
$\tilde\omega_s^{(m)} = \omega_s^{(m)} \big/ \sum_{s'=1}^{S} \omega_{s'}^{(m)}$,
and computes the weighted mean and the weighted covariance of the draws:

$$
\hat\beta_m = \sum_{s=1}^{S} \tilde\omega_s^{(m)} \beta_s, \qquad
\hat U_m = \frac{\sum_{s=1}^{S} \tilde\omega_s^{(m)}
    \big(\beta_s - \hat\beta_m\big)\big(\beta_s - \hat\beta_m\big)^\top}
  {1 - \sum_{s=1}^{S} \big(\tilde\omega_s^{(m)}\big)^2} .
$$

With equal weights $1/S$, these are the mean and the usual sample covariance
of the draws. The weights and a Pareto $\hat k$ for each plausible value come
from another program ([importance weights and the Pareto $\hat k$](#psis)).

## Importance weights and the Pareto $\hat k$ {#psis}

The importance weights of `stack_psis` reweight the draws of the stacked fit,
which serve as the proposal, toward the posterior for each plausible value.
The importance ratio of a draw is the ratio of that posterior density to the
proposal density at the draw. Suppose the stacked fit samples the stacked
fractional posterior

$$
q(\beta, \sigma) \propto p(\beta, \sigma) \prod_{m=1}^{M}
  \tilde L_m(\beta, \sigma)^{1/M} ,
$$

where $\tilde L_m$ is the survey-weighted likelihood of plausible value $m$,
$\sigma$ the residual standard deviation and $p(\beta, \sigma)$ the prior
([One stacked fit and the Rubin point
estimate](m3-stacked-bridge.html#stacked-objective)). If the posterior for
plausible value $m$ is
$p_m(\beta, \sigma) \propto p(\beta, \sigma)\, \tilde L_m(\beta, \sigma)$,
with the same prior, the log ratio at draw $(\beta_s, \sigma_s)$ is

$$
\log \frac{p_m(\beta_s, \sigma_s)}{q(\beta_s, \sigma_s)}
= \log \tilde L_m(\beta_s, \sigma_s)
  - \frac{1}{M} \sum_{m'=1}^{M} \log \tilde L_{m'}(\beta_s, \sigma_s) + c_m ,
$$

where $c_m$ does not depend on the draw and cancels when the weights are
divided by their sum.

A few draws with very large ratios can dominate raw importance weights.
Pareto smoothed importance sampling (PSIS) fits a generalized Pareto
distribution to the largest ratios, replaces them by values from the fitted
distribution, and reports the estimated shape parameter of that distribution,
the Pareto $\hat k$, as a diagnostic of how reliable the importance-sampling
estimates are [@vehtari2024pareto].

pvstackr computes no part of PSIS: it fits no Pareto distribution, smooths no
weights and estimates no $\hat k$. You supply, for each plausible value, the
weights $\omega_s^{(m)}$ and the Pareto $\hat k_m$ in one of two ways:

- `psis_weights`, a matrix with one row per stacked draw and one column per
  plausible value, with `pareto_k`, one value per plausible value;
- `log_ratios`, a matrix of log importance ratios of the same shape, with
  `psis_function`, a function of your own that pvstackr calls on `log_ratios`
  and that returns the weights and the $\hat k_m$, for example by applying the
  function `psis()` of the loo package to each column.

With `log_ratios` and `pareto_k` alone, pvstackr turns the ratios into
normalized weights without smoothing and blocks the fit, with the reason code
`psis_smoothing_not_applied` when the $\hat k_m$ pass the check below
(otherwise `psis_k_too_high` or `psis_k_not_evaluated`). `?pv_fit_stack_psis` lists all inputs.

By pvstackr's rule, a `stack_psis` fit has estimates (status `"ok"`) only when
the Pareto $\hat k_m$ of every plausible value is finite and strictly below
`psis_k_threshold` in `pv_control()`, which is 0.7 by default and cannot be set
higher, and when `psis_producer` and `psis_producer_version` name the program
that produced the weights. Otherwise the fit is blocked and has no estimates.
Its reason code is `psis_k_not_evaluated` when a $\hat k_m$ is missing or not
finite, `psis_k_too_high` when all are finite but one is at or above the
cut-off, and `psis_weight_provenance_incomplete` when the $\hat k_m$ pass but
no program is named. pvstackr records the name of the program but cannot check
it. It also records in `get_diagnostics(fit)$psis`, for each plausible value,
the effective sample size $1/\sum_s (\tilde\omega_s^{(m)})^2$ of the normalized
weights (`weight_ess_iid`) and their largest value (`max_normalized_weight`);
these values do not change the status.

The literature states the cut-off for $\hat k$ in two ways.
@vehtari2017practical [p. 1416] recommend warning the user when $\hat k$
exceeds 0.5, although they observed good performance in practice for values up
to 0.7, and above 0.7 advise considering alternatives to the
importance-sampling estimate. @vehtari2024pareto [pp. 5–6] consider requiring
$\hat k < 0.5$ stricter than necessary and recommend a warning when $\hat k$
exceeds $\min(1 - 1/\log_{10} S, 0.7)$ for $S$ draws, which is 0.67 for 1,000
draws and which they give as 0.7 for more than 2,000 draws. pvstackr uses a
single cut-off and no level at 0.5. To follow @vehtari2024pareto with fewer
than 2,000 draws, set `psis_k_threshold` to $1 - 1/\log_{10} S$.

## Why only `stack_direct` intervals can be coverage-claimable {#interval-rule}

By pvstackr's reporting rule, `coverage_claim_allowed` is `TRUE`, and
`interval_role` is `"coverage_barnard_rubin"`, only for the rows of a
`stack_direct` fit whose target was made with `df_method = "barnard_rubin"`
and a positive `df_complete`. The intervals of every `per_pv` and `stack_psis`
fit are descriptive. [Reading and reporting the
results](a3-reading-results.html#interval-columns) lists the labels of every
case and explains how to read and report them. The rule is pvstackr's own; it
is not a result of the companion preprint or of the sources of the formulas.

The rule follows the source of the variance within a plausible value. All
three methods add the variation between plausible values through $B$, and the
rule looks at $\bar U$, the average of the $\hat U_m$, whose source differs
between the methods; their $\hat\beta_m$, and with them $\bar\beta$ and $B$,
can differ as well. In a `stack_direct` fit,
$\hat U_m$ comes from the replicate weights: the estimate is recomputed with
each replicate weight, and the spread of the replicate estimates around the
full-sample estimate estimates its sampling variance [@judkins1990fay]. PISA
computes sampling variances in this way, and the method accounts for the
sampling of schools and of students within them; for an estimate from
plausible values, the variation between them is added separately
[@oecd2024pisa, chs. 10–11].

In `per_pv` and `stack_psis`, $\hat U_m$ is the covariance of posterior draws,
so it depends on the model. It reflects the clustering of the sample and the
unequal selection probabilities only as far as the model includes them; a
single-level regression, such as the one pvstackr fits, treats the students as
independent. Weighting the likelihood by the survey weights does not by itself
give a design-based variance: @williams2021uncertainty [pp. 72–73] call the
posterior of a survey-weighted likelihood a pseudo-posterior and show that
under a complex sampling design its asymptotic distribution differs in scale
and shape from that of the weighted estimate, and that its credible intervals
can fall short of nominal coverage. pvstackr therefore labels `per_pv` and
`stack_psis` intervals descriptive, whichever rule for the degrees of freedom
they use. This is pvstackr's reason, not a finding about the preprint's per-PV
workflow: in the preprint's simulation, whose model has a school random
intercept, that workflow's model-based intervals covered at 0.951 to 0.986
across strata and gradients [@lee2026onemcmc, sec. 5.2].

Among `stack_direct` fits, pvstackr also requires the Barnard–Rubin degrees
of freedom. @barnard1999small [p. 948] note that the classic degrees of
freedom of @rubin1987multiple are derived under the assumption that the
complete-data degrees of freedom $\nu_{\text{com}}$ are infinite, and that
when $\nu_{\text{com}}$ is small and little information is missing they can be
much larger than $\nu_{\text{com}}$; their adjusted degrees of freedom never
exceed $\nu_{\text{com}}$. With `df_method = "barnard_rubin"` you state
$\nu_{\text{com}}$ of the replicate-weight variance as `df_complete`; pvstackr
does not choose or check it ([The design-based target: BRR–Fay replicate
weights and Rubin's rules](m2-brr-fay-target.html#degrees-of-freedom) gives
the formulas). The labels follow `df_method` alone: `df_complete = Inf` gives
the classic degrees of freedom but keeps `coverage_claim_allowed = TRUE`, and
a fit with status `"warning"` keeps its labels.

The rule rests on the source of the variance, not on a study of coverage. The
companion preprint's simulation uses targets from model-based fits, on
simulated data without replicate weights, so it does not evaluate a BRR–Fay
target [@lee2026onemcmc, sec. 5]. Two parts of a `stack_direct` fit say
nothing about coverage either. The stacked fixed-effect point identity
concerns the point estimate only ([One stacked fit and the Rubin point
estimate](m3-stacked-bridge.html#scope)). CCC gives the draws the mean and
covariance of whatever target it receives, and its diagnostics compare the
stacked fit with that target. The preprint calls CCC a moment calibration and
treats the coverage of intervals built in this way as an empirical question
[@lee2026onemcmc, app. B.3]; [Calibrating the fixed-effect draws
(CCC)](m4-ccc-calibration.html) describes the calibration.

The label records how an interval was built; it does not certify coverage. A
coverage-claimable interval is built as in the standard replicate-weight
analysis, whose coverage is approximately nominal only under that analysis's
assumptions: the replicate covariance estimates the sampling covariance well,
the regression is congenial to the model that generated the plausible values,
and `df_complete` suits the design. pvstackr checks none of these. In the
simulations of @judkins1990fay [pp. 233, 236], intervals for a regression
coefficient based on replicate variances, Fay's method included, tended to
cover less often than their nominal level.

## Reading the Pareto $\hat k$ of a `stack_psis` fit

A Pareto $\hat k_m$ below the cut-off concerns the importance-sampling step.
@vehtari2024pareto [p. 7] find PSIS estimates of an expectation reliable, with
low bias and low variance, when the Pareto $\hat k$ for that expectation is
below 0.7 at the numbers of draws typical in practice; it can be larger than
the $\hat k$ of the importance ratios that `pareto_k` holds. When both are
small, the weighted mean and covariance of the stacked draws approximate the
posterior mean and covariance for plausible value $m$. A small $\hat k_m$ does
not change where that covariance comes from: when the weights reweight toward
the posteriors of the same model that a `per_pv` fit would use, a `stack_psis`
fit with small $\hat k_m$ approximates that `per_pv` fit, with its model-based
standard errors, and not the replicate-weight target. Its intervals stay
descriptive.

Whether the $\hat k_m$ are small depends on the data and the model. With a
common prior, the stacked fractional posterior $q$ is proportional to the
geometric mean of the $M$ posteriors for the separate plausible values,
$\prod_{m} \{p(\beta, \sigma)\, \tilde L_m(\beta, \sigma)\}^{1/M}$, and each
reweighting moves it toward one of them. The companion preprint notes that
when these posteriors are far apart relative to their spread, the importance
ratios have heavy tails and $\hat k$ rises, and it treats the reliability of
the reweighting as something to check in each analysis [@lee2026onemcmc,
sec. 4.5 and app. D.1]. The preprint reports its reweighted stack only when
$\hat k_m < 0.7$ for every plausible value, the cut-off that pvstackr uses by
default.

In the preprint's application to PISA 2022 reading, the reweighting was built
in the same way, for the same model, in the United States and in Korea. In the
United States the $\hat k$ of all ten plausible values were below 0.7, between
0.35 and 0.67; in Korea all ten were above it, between 0.82 and 1.34 and above
1 for three of them, and the Korean estimates were withheld [@lee2026onemcmc,
sec. 6.4 and table D.1]. The preprint explains the contrast by the larger and
more strongly clustered Korean sample, which makes each posterior narrow
relative to the distances between the posteriors, and calls this an
interpretation of the diagnostics rather than an established mechanism
[@lee2026onemcmc, pp. 22–23]. Its reweighting targets also differ from the
per-PV posteriors in the formula above: they place the survey weights in two
stages, a school factor on each school's integrated likelihood and a student
factor inside the integral over the school intercept (its Eq. (6)), whereas
the stacked proposal carries the final student weight at the student level.
Each of its ratios therefore spans a change of weight placement as well as a
change of plausible value, a gap that the preprint names as an additional
difficulty for the proposal [@lee2026onemcmc, secs. 4.5 and 6.4, app. D.1].
Its $\hat k$ values are thus not direct evidence about reweighting toward the
per-PV posteriors of a single-level model with the same weights. It also notes
that with the cut-off of
@vehtari2024pareto for its 1,000 draws, 0.67, the largest United States value,
0.674, would have failed as well [@lee2026onemcmc, app. D.1]. The model of the
application has a school random intercept, which pvstackr 0.2.x does not fit.

[Comparing the three fitting methods](a4-comparing-methods.html#cautions)
gives the practical cautions for reading `stack_psis` in a comparison, and its
section on [choosing a method](a4-comparing-methods.html#choosing) gives
pvstackr's recommendation, which uses `stack_psis` only as a check on a
stacked fit.

## One fit instead of $M$ {#one-fit}

`per_pv` needs $M$ Bayesian fits, one for each plausible value. `stack_direct`
and `stack_psis` need one, to the stacked data: $M$ copies of the $N$ rows of
the data, in which row $i$ of copy $m$ has the outcome $y_i^{(m)}$. In a
`stack_direct` fit that row has the weight $\tilde w_i/M$, where
$\tilde w_i = W_i/\overline W$ is the final weight $W_i$ divided by the mean
final weight $\overline W$ ([One stacked fit and the Rubin point
estimate](m3-stacked-bridge.html#stacked-objective) derives the stacked
objective from these weights). When `stack_psis` stacks the data for your
fitting function, the rows get the same weights if you give `weight_col`, and
the weight $1/M$ otherwise. With the ten plausible values per domain of PISA
2022 [@oecd2024pisa], the stacked methods need one fit where `per_pv` needs
ten.

The number of fits says how many samplers run and how many sets of
convergence diagnostics you check, not how long the analysis takes: the
stacked fit has $M$ times as many rows
as a fit to one plausible value, and `stack_direct` also needs its target,
$M(R + 1)$ weighted least-squares fits, which need no sampler. The companion
preprint likewise counts the gain as one fit, one set of convergence
diagnostics and one set of calibrated draws per model, and makes no claim
about processor time [@lee2026onemcmc, sec. 7]. `pv_compare_methods()`
records the number of model fits of each method and, when you pass times that
you measured with its argument `timings`, those times; it does not time the
fits itself. [Real PISA data: access, memory and
reproducibility](a5-real-pisa-guidance.html#memory-runtime) gives the size of
these jobs for PISA 2022.

## Checks {#checks}

The first chunk writes pvstackr's check of the Pareto $\hat k$ values in base
R and applies it to made-up values: it illustrates the rule and is not the
function that pvstackr runs. The second reads the interval labels that
pvstackr gave the example fit.

### The Pareto $\hat k$ check

```{r k-hat-check}
## pvstackr's check of the Pareto k-hat values of a stack_psis fit, written in
## base R to illustrate it; pvstackr applies it inside pv_fit_stack_psis().
## A value passes when it is finite and strictly below the cut-off.
k_passes <- function(k, threshold = 0.7) is.finite(k) & k < threshold

k <- c(0.20, 0.50, 0.69, 0.70, 0.95, NA)   # made-up k-hat values
data.frame(
  k            = k,
  passes_0.7   = k_passes(k),                               # the default cut-off
  passes_0.667 = k_passes(k, threshold = 1 - 1 / log10(1000))
)
```

Each row applies the check to one made-up value. With the default cut-off of
0.7, the values 0.20, 0.50 and 0.69 pass: pvstackr has no separate level at
0.5. The value 0.70 fails because the check is strict, and 0.95 and the
missing value fail as well. A `stack_psis` fit has estimates only when the
values of all its plausible values pass and the program that made the weights
is named. A fit with a missing or non-finite value is blocked with the reason
code `psis_k_not_evaluated`, and a fit whose values are finite but not all
below the cut-off with `psis_k_too_high`. The last column uses the cut-off
that @vehtari2024pareto recommend for 1,000 draws,
$1 - 1/\log_{10} 1000 \approx 0.667$, which you would set with
`pv_control(method = "stack_psis", psis_k_threshold = 1 - 1/log10(1000))`;
0.69 then fails too.

### The interval labels of the example fit

```{r coverage-flag}
est <- get_estimates(fit)

unique(est$interval_role)
unique(est$coverage_claim_allowed)
est[, c("term", "interval_role", "df_method", "coverage_claim_allowed")]
```

All three rows have `interval_role = "descriptive_classic_rubin"`,
`df_method = "classic"` and `coverage_claim_allowed = FALSE`. The fit is a
`stack_direct` fit, but its target was made with the classic degrees of
freedom, so by pvstackr's rule its intervals are descriptive; a target made
with `df_method = "barnard_rubin"` and a positive `df_complete` would give
`coverage_barnard_rubin` and `TRUE`. The labels do not depend on the draws:
the fit copies them from its target.

[Comparing the three fitting methods](a4-comparing-methods.html) lines up an
example fit of each method with `pv_compare_methods()` and reads their labels
side by side, and [Reading and reporting the
results](a3-reading-results.html#interval-columns) explains how to report each
kind of interval.

The numbers above were computed in this session:

```{r session-info}
sessionInfo()
```

## References
